### 2023-04-05 manuscript/manuscript.Rmd
merging the new repository into this one to avoid repitition.

---
title: "Imposed form in the Early Acheulean? Evidence from Gona, Afar, Ethiopia"
author:
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA;  dwstout@emory.edu]
- Antoine Muller^[Institute of Archaeology, Mount Scopus, The Hebrew University of Jerusalem, Jerusalem, Israel; antoine.muller@mail.huji.ac.il]
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Michael J. Rogers^[Department of Anthropology, Southern Connecticut State University, New Haven, CT, USA; rogersm1@southernct.edu]
- Sileshi Semaw^[Centro Nacional de Investigación sobre la Evolución Humana (CENIEH), Burgos, Spain; sileshi.semaw@cenieh.es]
abstract: "TBD.\\par \\textbf{Keywords:} Gona; TBD; TBD; TBD; TBD; TBD; TBD"
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
editor_options: 
  markdown: 
    wrap: sentence
---

# Introduction

The imposition of intended form on artifacts has long been viewed as a watershed in human cognitive and cultural evolution and is most commonly associated with the emergence of "Large Cutting Tools" (LCTs) in the Early Acheulean [@isaac1976; @kuhnEvolutionPaleolithicTechnologies2020; @holloway1969]. However, this interpretation of Acheulean LCTs as intentionally designed artifacts remains controversial. Alternative proposals range from the possibility that LCTs were unintended by-products of flake production [@moore2016; @noble1996] to the suggestion that their form was "at least partly under genetic control" [@corbey2016]. Even accepting that LCT form was to some extent intended, there is substantial disagreement over the specificity of design. Some analyses have indicated that shape variation in Acheulean handaxes is largely a result of resharpening [@iovita2011; @mcpherron2000] whereas others find form to be unrelated to reduction intensity and more likely to reflect normative expectations of what handaxes should look like [@garcía-medrano2019; @shiptonHandaxeReductionIts2015a; @shipton2020]. Debates about shape of Acheulean LCTs may appear narrowly technical but have broad relevance for evolutionary questions including the origins of human culture [@corbey2016; @tennie2017; @shiptonHandaxeReductionIts2015a], language [@stout2012], teaching [@gärdenfors2017], brain structure [@hecht2015], and cognition [@stout2015; @wynn2016].
To examine these questions, we studied the complete collection of Early Acheulean flaked pieces from 5 sites at Gona and compared them with Oldowan cores form 2 published sites.
By comparing shape variation to typological assignment and measures of flaking intensity, we sought to identify technological patterns that might reveal intent.

There appears to be a broad consensus that refined handaxes and cleavers from the later Acheulean resulted from procedurally elaborate, skill intensive, and socially learned production strategies [@caruana2020; @garcía-medrano2019; @moore2020; @sharon2009; @shiptonEvolutionSocialTransmission2019; @stout2014] although debate over the presence of explicit, culturally transmitted shape preferences continues [@iovita2011; @moore2020; @shipton2020; @wynn2018]. However, there is less agreement regarding the cruder LCTs that are typical of the earliest Acheulean [@beyene2013; @delatorre2018; @diez-martín2015; @lepre2011; @semaw2018] and which continue to occur with variable frequency in later time periods [@mcnabb2015] including in eastern Asia [@li2021].
Although several formal types have been recognized, many workers now see continuum, including the possibility that flake production remained an important or even primary purpose of LCT reduction (Shea).
This questions imposed form.
It is clear that LFB is new but not clear t

Types are questionable, Kuhn on LFB.
Isaac but some on cobbles.
It is even controversial whether asia is "acheulean" Prevailing opinion, but Beyene.
A conservative interpretation of available evidence is that LCT production was guided by a recurring set of functional, ergonomic, and aesthetic design preferences [@wynn2018] with other elements free to vary in response to raw materials, use life, and random population dynamics like drift, bottlenecks, and founder effects [@kuhnEvolutionPaleolithicTechnologies2020; @lycett2016].

Hypotheses: 1) Valid technological types should produce clear morphological clusters with different reduction trajectories vs. points along a continuum.
2) Debitage is indicated by relation of SDI and flaked area to core size but not shape.
3) Shaping is indicated by relation of flaked area to shape & weaker or absent relations of shape with SDI.
Shape independent of size.
4) Shaping plus resharpening means shape should be related to core size and SDI (Shipton)

# Materials and Methods

## Materials

Archaeological Sample

## Methods

### Artifact Shape Measurement

Three-dimensional scanning and geometric morphometric (3DGM) methods are becoming increasingly common in the study of LCT form [@archer2010; @caruana2020; @li2021; @lycett2006; @presnyakova2018; @shiptonHandaxeReductionIts2015a]. These methods can provide high-resolution, coordinate-based descriptions of artifact form including detailed information about whole object geometric relations that is not captured by conventional linear measures [@shott2010]. However, they also impose additional costs in terms of data collection and processing time as well as required equipment, software, and training. Insofar as these costs might present an obstacle to participation by some researchers and/or draw resources away from other activities, they must be balanced against benefits. In particular, it is not clear that these powerful methods are required in order to describe relevant variation in Acheulean LCT shape. Unlike hominin crania or even projectile points, Acheulean handaxes, cleavers, and picks are not complex shapes. Individual LCTs exhibit complex morphologies defined by idiosyncratic scar patterns, but these details are largely noise at the level of comparative analyses. Laser-scanning 3DGM studies of LCTs collect vast amounts of shape data, but typically discard upward of 50% of the observed variation in order to focus on two or three interpretable principal components. Across studies, these PCs consistently corresponding to basic features like elongation, relative thickness, pointedness, and position of maximum thickness that also emerge from lower-resolution spatial data [@archer2010; @garcía-medrano2019; @lycett2006] and studies employing linear measures rather than spatial coordinates [@crompton1993; @pargeter2019].
Thus, while the level of detail enabled by 3DGM is arguably useful for building artifact phylogenies [@okumura2019], it is of questionable behavioral/technological relevance for the study of LCTs.
For these reasons, we favored the use of simple caliper-based linear measures to quantify shape in our study.
Nevertheless, Shott and Trail [-@shott2010] do identify three potential shortcomings of linear measurements compared to 3DGM.
We considered each in the context of our particular materials and research questions.
First, conventional linear measures capture the direction (e.g. length \> breadth) but not the location of geometric relations (e.g. position of maximum breadth).
We address this by collecting linear measures defined by homologous semi-landmarks.
All artifacts were oriented along their maximum dimension, which was measured and defined as "length." The next largest dimension orthogonal to length was used to define the plane of "breadth," with the dimension orthogonal to this plane defined as "thickness." Breath and thickness measures were then collected at 25%, 50%, and 75% of length, oriented so that 25% Breadth \> 75% Breath.
To partition variation in shape from variation in size, we divided all linear measures by the geometric mean [@lycett2006].

Second, linear measures risk reducing complex forms to overly simplistic "stick figure caricatures" [@shott2010]. However, whether or not this risk actually presents a problem depends on the particular artifacts and research questions involved. We have already noted that 3DGM LCT studies typically evaluate only a small portion of the measured variation. To better evaluate the measurement density required for our study, we reanalyzed a data set of 128 experimental handaxes previously published by Pargeter et al. [-@pargeter2019]. These data comprise 19 linear measures (length plus breadth and thickness at 10% increments of length) collected from digital photos using the same orientation protocol described above. We conducted a PCA on the full set of 19 measures and again on a reduced set of 7 (length plus breadth and thickness at 30%, 50%, and 70% length). Despite this reduction, the first two components from each analysis displayed strikingly similar component loading matrices (PC1 positive on length and tip breadth, negative on thickness; PC2 positive on base breadth, negative on length and thickness) almost perfectly correlated component scores for individual pieces (PC1 r=0.919, PC2 r=0.913). As a further check, we performed the same comparison on a subset of the current archaeological sample from Gona for which photos were available for measurement (n = 50). This produced two PCs that were not only similar with each other, but also matched the PCs extracted from the experimental handaxe sample. Individual piece component scores were again highly correlated (r=0.975 and 0.927 respectively). Seven linear measures thus appear sufficient to explain technologically/behaviorally relevant shape variation in our sample. Third, linear measures may struggle to capture attributes such as cross-sectional area and shape (e.g. Caruana, 2020) more easily assessed using 3DGM. Particularly relevant here are measures of surface area used to calculate indices of reduction intensity [@shiptonHandaxeReductionIts2015a; @clarkson2013] and surface modification [@li2015] used in our study.
Clarkson [-@clarkson2013] advocates the use of 3D surface area measures as more accurate than estimation from linear measures (e.g. surface area of a rectangular prism defined by artifact dimensions).
However, he also found that the error introduced by the linear approach was a highly systematic, isometric overestimation of surface area and that results correlated with direct 3D measures with an impressive r2 = 0.944 and no effect of variation in core shape.
Insofar as it is variation in the relationship between surface area and flaking intensity that is of interest, rather than the absolute size of artifacts, such consistent overestimation is not problematic.
Here we improved on the prism-based surface area formula (2LW + 2LT + 2 WT) by using our 7 recorded dimensions to more tightly fit three prisms (Figure 1) around the artifact: SA = W1T1 + 2(.33L \* W1) + 2(.33L \* T1) + 2(.33L \* W2) + 2(.33L \* T2) + 2(.33L \* W3) + 2(.33L \* T3) + W3T3.
Surface area calculated in this way correlates with mass2/3 at r2 = 0.947 in our sample.

### Reduction Indices

Research by Clarkson and Shipton has established the Scar Density Index (SDI = number of flake scars \> 1cm per unit surface area) as a reliable indicator of mass removed from a core across technologies [@clarkson2013] and for handaxes specifically [@shiptonFlakeScarDensity2015].
We thus use SDI as an indicator of reduction intensity (mass removed) in our study.
However, reduction intensity does not constitute a full description of core modification.
Mass removal is the aim during flake production and extent of shaping are not necessarily the same thing.
For example, imposition of a desired form

# Results

A PCA (covariance matrix) on our 7 linear measures (scaled by geometric mean) for pieces identified two PCs explaining 80% of variance (56.4% and 23.7%).
Rescaled component matrix shows that PC1 reflects "flatness" (length and breadth vs. thickness).
PC

Two-step cluster analysis identified 3 clusters.

Typologically, these loosely correspond to Mode 1 cores, Large Flake/Knifes, and Picks, with handaxes split between knife vs. pick categories.

PC1 differentiates Mode 1 and Mode 2 pretty well, in that M1 cores tend not to be flat or elongated.
Mode 1 exceptions (i.e. misclassified on shape) are generally still distinguishable as smaller and more heavily reduced than Mode 2 (of Mode 1 included in Cluster 1: mean weight =159.4 vs. 635.6, p \< 0.001; Mean logSDI = .74 vs. .20, p \< 0.001).
(of Mode 1 included in Cluster 3: mean weight =224.1 vs. 398.1, p \< 0.001; Mean logSDI = .67 vs. .39, p = 0.004).
We thus treat Mode 1 as a valid techno-morphological category.
Consistent with the characterization of Mode 1 as focused on debitage rather than shaping, we observe a strong power relationship between reduction intensity (SDI) and core size (r2=0.715, p \< 0.001, b1 = -0.872):

In contrast, and also in keeping with a focus on debitage rather than shaping and resharpening, there is no such relationship with shape PCs for SDI:

Cluster 1 is divided from Cluster 3 by PC2 (pointedness).
Cluster 1 is much more likely to be executed on a flake base (91% flakes) vs. cluster 3 (35% flakes).
Cluster 1 is also significantly less reduced (Mean logSDI = .39 vs. .20, p \< 0.001).
So, cluster 1 basically comprises lightly retouched LFB acheulean, with shapes that remain largely within the range of unmodified flakes (n.s. mean difference).

The effect of reduction on LFB acheulean shape is evident only for flaked area (not SDI) and corresponds to decreases in both PCs (i.e. less elongated but more pointy).
The PC1 effect is relatively weak (r2=0.1, p =0.008, Standardized Beta = -0.215).
The PC2 effect is stronger (r2=0.244, p \< 0.001, Standardized Beta = -0.537).
This is most consistent with flaking placed to shape a point.
A weak power effect of SDI on weight (r2=0.178, p \< 0.001, b1= -.330), as well as low number of scars in general, suggests resharpening is not a major factor.

These trends mean that heavily modified flakes enter into cluster 3 (i.e. look like picks).
Indeed, 40% of identifiable bases for cluster 3 are flakes.
Cluster 3 pieces executed on flakes tend to be less pointed regardless of reduction intensity, which is likely a reflection of starting blank form.
Indeed, Mode 2 Cobble bases show no effect of reduction intensity on shape but do show SDI effect on weight (r2=0.432, p \< 0.001, b1=-0.711).
This appears to reflect the presence of cobble blanks that are already relatively pointed without substantial reduction and raises the possibility that these pieces are produced through debitage on pointed cobbles.
Could they start as LFB cores?
look at maximum flake scar size.
Large cores have few, large scars.

These patterns indicates that there is a common reduction trajectory for Mode 2 forms at Gona, regardless of typology or blank form.
Although some pieces start much closer to the terminal morphology than others (i.e. display low PC2 values without substantial reduction), none undergo substantial reduction without becoming pointed.

This uniform trajectory casts serious doubt on the likelihood that picks are a distinct morpho-functional type, although they may represent "4-dimensional design" sensu Kuhn.
edge angles up to 70 degrees are quite efficient and obtuse trimming of butt may help ergonomics.

No evidence for shaping of cobbles.

Acheulean cobbles are larger and more cylindrical (geologically “rollers”). Shape difference may reflect availability and/or selectivity for flake-able shapes.

Some Achuelean flaked cobbles might hypothetically be heavily reduced remnants of giant cores for LFB production. Size of scars overlaps with LFBs (right). However, smaller scars are present and so reduction seems to have continued past potential for LFB production. Cobbles are generally pretty heavily reduced.

Acheulean flaked flakes seem to have been (mildly) shaped to increase elongation and pointedness. This might have been an explicit design target, a passive result of preferentially flaking working edges, and or a desire to retain length for some reason.

# References

### 2023-04-05 code/Analysis.R

@@ -1,0 +1,23 @@
+# Add packages
+library(tidyverse)
+library(factoextra)
+library(FactoMineR)
+library(PCAmixdata)
+library(psych)
+# Add data
+gona_esa<-read.csv("data/EA_Reduction_Paper_Order.csv")
+
+# Principle component analysis
+gona_esa_pca <-prcomp(gona_esa[1:7],  scale = TRUE)
+pcaplot<-fviz_pca_ind(gona_esa_pca, habillage= gona_esa$Type, # color by groups
+                      addEllipses = TRUE, # Concentration ellipses
+                      ellipse.type = "convex",
+                      legend.title = "Type",
+                      label="none")
+pcaplot
+res.var <- get_pca_var(gona_esa_pca)
+variable_loadings<-data.frame(res.var$coord) 
+
+pca2 <- psych::principal(gona_esa[1:7], nfactors=2, rotate="varimax", scores=F)
+pca2
+

### 2023-04-05 code/old/txt_reading_in_mass.R

@@ -1,0 +1,236 @@
+library(tidyverse)
+
+list_of_files <- list.files(path = "/Users/raylc/Downloads/Boxgrove Taylor work/Handaxes/boxgrove_measurements/plan_2", recursive = TRUE, full.names = TRUE)
+
+# Read all the files and create a FileName column to store filenames
+df <- list_of_files %>%
+  set_names(.) %>%
+  map_df(read_table2, .id = "FileName")
+Handaxes/boxgrove_measurements
+
+
+
+
+
+
+
+
+
+
+
+
+
+plan_filenames_160<-list.files("Handaxes/boxgrove_measurements/plan_822_160", pattern="*.txt")
+profile_filenames_160<-list.files("Handaxes/boxgrove_measurements/profile_822_160", pattern="*.txt")
+plan_filenames_162<-list.files("Handaxes/boxgrove_measurements/plan_2", pattern="*.txt")
+profile_filenames_162<-list.files("Handaxes/boxgrove_measurements/profile_2", pattern="*.txt")
+
+
+
+merge_shape_data_function_160<-function(plan_names,profile_names,plan_pathway,profile_pathway) {
+  
+  # combine txt files for each recording set
+  handaxe_plan_measurements<-do.call(rbind,lapply(plan_names,open_read,plan_pathway))
+  handaxe_profile_measurements<-do.call(rbind,lapply(profile_names,open_read,profile_pathway))
+  
+  # clean and reshape plan data 
+  handaxe_plan_measurements_reshape<-handaxe_plan_measurements %>%
+    mutate(variable=recode(handaxe_plan_measurements$variable, Width = "width"),
+           measurement_point = paste(variable, measurement_point, sep="_"),
+           shape_width_mm=measurement*25.4) %>%
+    select(-c(variable,measurement)) %>%
+    dcast(individual ~ measurement_point,value.var="shape_width_mm")
+  
+  # clean and reshape profile data 
+  handaxe_profile_measurements_reshape<-handaxe_profile_measurements %>%
+    mutate(variable=dplyr::recode(variable, Width = "thickness"),
+           measurement_point = paste(variable, measurement_point, sep="_"),
+           shape_thickness_mm=measurement*25.4) %>%
+    select(-c(variable,measurement)) %>%
[…]

### 2023-04-07 manuscript/manuscript.Rmd
updated the method section based on Dietz's slides.

@@
 
+PCA on GM-transformed caliper measures (length, 3 breadth, 3 thickness). Length is maximum dimension, piece oriented so that Br1>Br3
+
+Typological and technological attributions considered unreliable. Data grouped according to context (~2.5 mya Oldowan sites vs. ~1.5 mya Acheulean sites) and blank form (cobble, flake, indeterminant).
+
+Associations between form and reduction intensity are considered as an indicator of “imposed form.” Such form could reflect mental templates and/or biased flaking patterns due to functional or technological constraints
+
 ### Reduction Indices

### 2023-07-07 manuscript/manuscript.Rmd
introduction updates by Dietz.

--- 2023-04-05 manuscript/manuscript.Rmd
@@
 - Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA;  dwstout@emory.edu]
+- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
 - Antoine Muller^[Institute of Archaeology, Mount Scopus, The Hebrew University of Jerusalem, Jerusalem, Israel; antoine.muller@mail.huji.ac.il]
-- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
 - Michael J. Rogers^[Department of Anthropology, Southern Connecticut State University, New Haven, CT, USA; rogersm1@southernct.edu]
@@
 Some analyses have indicated that shape variation in Acheulean handaxes is largely a result of resharpening [@iovita2011; @mcpherron2000] whereas others find form to be unrelated to reduction intensity and more likely to reflect normative expectations of what handaxes should look like [@garcía-medrano2019; @shiptonHandaxeReductionIts2015a; @shipton2020].
-Debates about shape of Acheulean LCTs may appear narrowly technical but have broad relevance for evolutionary questions including the origins of human culture [@corbey2016; @tennie2017; @shiptonHandaxeReductionIts2015a], language [@stout2012], teaching [@gärdenfors2017], brain structure [@hecht2015], and cognition [@stout2015; @wynn2016].
-To examine these questions, we studied the complete collection of Early Acheulean flaked pieces from 5 sites at Gona and compared them with Oldowan cores form 2 published sites.
-By comparing shape variation to typological assignment and measures of flaking intensity, we sought to identify technological patterns that might reveal intent.
+Such debates about shape of Acheulean LCTs may appear narrowly technical but have broad relevance for evolutionary questions including the origins of human culture [@corbey2016; @tennie2017; @shiptonHandaxeReductionIts2015a], language [@stout2012], teaching [@gärdenfors2017], brain structure [@hecht2015], and cognition [@stout2015; @wynn2016].
+To examine these questions, we studied the complete collection of Early Acheulean flaked pieces from 5 sites at Gona Project Area and compared them with Oldowan cores form 2 published sites at Gona.
+By comparing shape variation to measures of flaking intensity and patterning, we sought to identify technological patterns that might reveal intent.
 
-There appears to be a broad consensus that refined handaxes and cleavers from the later Acheulean resulted from procedurally elaborate, skill intensive, and socially learned production strategies [@caruana2020; @garcía-medrano2019; @moore2020; @sharon2009; @shiptonEvolutionSocialTransmission2019; @stout2014] although debate over the presence of explicit, culturally transmitted shape preferences continues [@iovita2011; @moore2020; @shipton2020; @wynn2018].
-However, there is less agreement regarding the cruder LCTs that are typical of the earliest Acheulean [@beyene2013; @delatorre2018; @diez-martín2015; @lepre2011; @semaw2018] and which continue to occur with variable frequency in later time periods [@mcnabb2015] including in eastern Asia [@li2021].
-Although several formal types have been recognized, many workers now see continuum, including the possibility that flake production remained an important or even primary purpose of LCT reduction (Shea).
-This questions imposed form.
-It is clear that LFB is new but not clear t
+## Origin of design
 
-Types are questionable, Kuhn on LFB.
-Isaac but some on cobbles.
-It is even controversial whether asia is "acheulean" Prevailing opinion, but Beyene.
-A conservative interpretation of available evidence is that LCT production was guided by a recurring set of functional, ergonomic, and aesthetic design preferences [@wynn2018] with other elements free to vary in response to raw materials, use life, and random population dynamics like drift, bottlenecks, and founder effects [@kuhnEvolutionPaleolithicTechnologies2020; @lycett2016].
+There is a broad consensus that refined handaxes and cleavers from the later Acheulean resulted from procedurally elaborate, skill intensive, and socially learned production strategies [@caruana2020; @garcía-medrano2019; @moore2020; @sharon2009; @shiptonEvolutionSocialTransmission2019; @stout2014] although debate over the presence of explicit, culturally transmitted shape preferences continues [@iovita2011; @moore2020; @shipton2020; @wynn2018].
+There is much less agreement regarding the less heavily worked and formally standardized LCTs typical of the earliest Acheulean [@beyene2013; @delatorre2018; @diez-martín2015; @lepre2011; @semaw2018].
+Such forms continue to occur with variable frequency in later time periods [@mcnabb2015], and may be especially prevalent in eastern Asia [@li2021].
+Although formal types have been recognized in the Early Acheulean and are commonly used to describe assemblages, many workers now see a continuum of morphological variation [@duke2021; @kuhn2020; @presnyakova2018]including the possibility that simple flake production remained an important [@shea2010] or even primary [@moore2016] purpose of Early Acheulean large core reduction.
+
+Typologically, LCTs are differentiated from Mode 1 pebble cores on the basis of size (\>10cm) and shape (elongation and flattening) [e.g., @isaac1977].
+This consistent production of large, flat, and elongated cores in the Achuelean has long been thought to reflect the pursuit of desired functional and ergonomic properties for hand-held cutting tools [@wynn2018].
+Unplanned flaking can sometimes produce cores that fall into the LCT shape range [@moore2016] and this is one possible explanation of the relatively small "protobifaces" that occur in low frequencies in Oldowan assemblages [@isaac1997].
+However, the Early Acheulean is clearly distinguished from the Oldowan by the production of larger artifacts necessitating the procurement and exploitation of larger raw material clasts.
+Although studies of handaxe variation often focus on shape rather than size, this shift is an important aspect of artifact design with relevance to both production and function.
+
+Production of larger tools was accomplished either through a novel process of detaching and working Large Flake Blanks (LFBs) from boulder cores or simply by using larger cobble and slab cores [@delatorre2018; @isaac1969; @semaw2018].
+Both may involve similar flaking "strategies" (e.g., bifacial or multifacial exploitation) to those present in the Oldowan [@duke2021] but require more forceful percussion to detach larger flakes.
+This increases the perceptual motor difficulty of the task [@stout2002] and in many cases may have been accomplished using different percussive techniques and supports [@semaw2009].
+These new challenges would have increased raw material procurement [@shea2010] and learning costs [@pargeter2019] as well as the risk of serious injury [@gala2023] associated with tool production.
+This strongly implies intentional pursuit of offsetting functional benefits related to size increase.
+These likely included tool ergonomics and performance [@key2017] as well as flake generation, resharpening, and reuse potential [@shea2010].
+Early Acheulean LCT production is thus widely seen as a part of shifting hominin behavioral ecological strategies including novel resources and mobility patterns [@linaresmatás2021; @rogers1994].
+
+The degree of intentional design reflected in the shape of Early Acheulean LCTs is more difficult to determine.
+For example, LFB production using a simple "least effort" bifacial/discoidal strategy will tend to generate predominantly elongated (side or end struck) flakes [@toth1982] whether or not this is an intentional design target.
+Similarly, the difficulty of flaking relatively spherical cobbles [@toth1982] might bias initial clast selection and subsequent reduction toward flat and elongated shapes even in the absence of explicit design targets.
+On the other hand, it has been argued that the shape of Early Acheulean LFBs was intentionally predetermined using core preparation techniques [@delatorre2018] and many researchers perceive efforts at intentional shaping in the organization of flake scars on Early Acheulean handaxes and picks [@beyene2013; @delatorre2018; @diez-martín2015; @duke2021; @lepre2011; @semaw2009].
+To date, however, the identification of Early Acheulean shaping has generally relied on qualitative assessment by lithic analysts.
+Such assessment may in fact be reliable, but is subject to concerns about potential selectivity, bias, and/or overinterpretation [@davidson2002; @moore2016].
+
+\
+This qualitative approach stands in contrast to investigations of Later Achuelean shaping
 
@@
 4) Shaping plus resharpening means shape should be related to core size and SDI (Shipton)
+
+It is even controversial whether asia is "acheulean" Prevailing opinion, but Beyene.
+A conservative interpretation of available evidence is that LCT production was guided by a recurring set of functional, ergonomic, and aesthetic design preferences [@wynn2018] with other elements free to vary in response to raw materials, use life, and random population dynamics like drift, bottlenecks, and founder effects [@kuhnEvolutionPaleolithicTechnologies2020; @lycett2016].
 
@@
 We have already noted that 3DGM LCT studies typically evaluate only a small portion of the measured variation.
-To better evaluate the measurement density required for our study, we reanalyzed a data set of 128 experimental handaxes previously published by Pargeter et al.
-[-@pargeter2019].
+To better evaluate the measurement density required for our study, we reanalyzed a data set of 128 experimental handaxes previously published by Pargeter et al. [-@pargeter2019].
 These data comprise 19 linear measures (length plus breadth and thickness at 10% increments of length) collected from digital photos using the same orientation protocol described above.
@@
 Seven linear measures thus appear sufficient to explain technologically/behaviorally relevant shape variation in our sample.
-Third, linear measures may struggle to capture attributes such as cross-sectional area and shape (e.g.
-Caruana, 2020) more easily assessed using 3DGM.
+Third, linear measures may struggle to capture attributes such as cross-sectional area and shape (e.g. Caruana, 2020) more easily assessed using 3DGM.
 Particularly relevant here are measures of surface area used to calculate indices of reduction intensity [@shiptonHandaxeReductionIts2015a; @clarkson2013] and surface modification [@li2015] used in our study.
@@
 Surface area calculated in this way correlates with mass2/3 at r2 = 0.947 in our sample.
+
+PCA on GM-transformed caliper measures (length, 3 breadth, 3 thickness).
+Length is maximum dimension, piece oriented so that Br1\>Br3
+
+Typological and technological attributions considered unreliable.
+Data grouped according to context (\~2.5 mya Oldowan sites vs. \~1.5 mya Acheulean sites) and blank form (cobble, flake, indeterminant).
+
+Associations between form and reduction intensity are considered as an indicator of "imposed form." Such form could reflect mental templates and/or biased flaking patterns due to functional or technological constraints
 
@@
 
-Acheulean cobbles are larger and more cylindrical (geologically “rollers”). Shape difference may reflect availability and/or selectivity for flake-able shapes.
+Acheulean cobbles are larger and more cylindrical (geologically "rollers").
+Shape difference may reflect availability and/or selectivity for flake-able shapes.
 

### 2023-08-05 manuscript/manuscript.Rmd
introduction update.

@@
 - Sileshi Semaw^[Centro Nacional de Investigación sobre la Evolución Humana (CENIEH), Burgos, Spain; sileshi.semaw@cenieh.es]
-abstract: "TBD.\\par \\textbf{Keywords:} Gona; TBD; TBD; TBD; TBD; TBD; TBD"
+abstract: 
+|
+  ```{=latex}
+  TBD.
+
+  \textbf{Keywords:} Gona; TBD; TBD; TBD; TBD; TBD; TBD
+  ```
 date: "`r Sys.Date()`"
@@
 
-## Origin of design
+## Identifying design
 
@@
 Such assessment may in fact be reliable, but is subject to concerns about potential selectivity, bias, and/or overinterpretation [@davidson2002; @moore2016].
+Notably, a 3-dimensional morphometric (3DGM) study by Presnyakova, et al.
+[-@presnyakova2018] concluded that LCT shape variation in the Okote Member (\~1.4 mya) at Koobi Fora was largely driven by reduction intensity rather than different knapping strategies.
+However, this study did not directly address the presence/absence of design targets constraining the observed range of variation.
 
-\
-This qualitative approach stands in contrast to investigations of Later Achuelean shaping
+In later Acheulean contexts, reduction intensity effects are commonly equated with resharpening and seen as an alternative to intentional form imposition [@mcpherron2000].
+Across heavily-worked and relatively standardized LCT assemblages [e.g.
+@shipton2020], a *lack* of association between morphology and reduction intensity has be used as an argument-by-elimination for the presence of imposed morphological norms or \"mental templates\" [@garcía-medrano2019; @shipton2015; @shipton2023].
+However, in the less heavily-worked and more heterogenous assemblages typical of the early Acheulean [@kuhn2020], it is equally plausible that increasing reduction intensity would reflect degree of primary reduction rather than subsequent resharpening [@archer2010].
+In this case, reduction intensity effects on morphology would have the opposite interpretation:  more reduction should result in closer approximation of a desired form if such were present.
+For example, Beyene, et al.
+[-@beyene2013] found that increasing flake scar counts were associated with increasing handaxe refinement through time at Konso, Ethiopia, which may reflect a more general trend in the African Acheulean [@shipton2018].
+
+Interpretive approaches address this quandary by \"reading\" the organization of scars on individual pieces to infer intent, but an adequate method to objectively quantify these insights has yet to be developed.
+Current measures of reduction intensity, such as the scar density index (SDI) [@clarkson2013], are designed to estimate total mass removed from a core and are reasonably effective [@lombao2023].
+However, mass removal was not the objective of Paleolithic flaking.
+Indeed, knapping efficiency is usually conceived as generating an outcome while *minimizing* required mass removal.
+This is true whether the desired outcome is a useful flake, a rejuvenated edge, or a particular core morphology.
+In simple flake production, mass removed is probably a good reflection of the completeness of exploitation (\"exhaustion\") of cores and may have implications for required skill [@toth1982; @pargeter2023] as well as raw material economy [@reeves2021; @shick1987].
+However, in core shaping and resharpening, mass removal would typically represent an energetic and raw material cost to be minimized, and might even interfere with function [@keyHandaxeShapeVariation2019].
+Without further information, relationships between artifact shape and reduction intensity are thus open to conflicting interpretations as evidence of intentional design or its absence.
+
+Li, et al.
+[-@li2015] proposed a Flaked Area Index (FAI) as an alternative to SDI as a measure of reduction intensity, arguing that its validity is supported by an observed correlation (r=0.424) with SDI.
+However, they also explain that \"flaked area does not necessarily relate to the number of flake scars\...a small number of large scars can produce a large area of scar coverage, and conversely, a large number of small scars can produce a small area of scar coverage.\" (p.
+6).
+We suggest that what FAI actually captures is the spatial extent modifications to the surface of a core.
+It is thus complementary to the measure of volume reduction provided by SDI and provides additional information to inform technological interpretations.
+For example, a correlation between FAI and artifact form without any effect of SDI would suggest a focus on \"least-effort\" shape imposition whereas the opposite pattern would be consistent with relatively intense resharpening of spatially restricted areas on the core.
+A lack of shape correlation with either measure would be expected for simple debitage with no morphological targets whereas a strong correlation with both would indicate a highly \"designed\" form achieved through extensive morphological and volumetric transformation.
+In the current study we thus considered SDI and FAI together in order to evaluate evidence of intentional shaping in the early Acheulean of Gona.
+
+## Measuring artifact form and modification
 

### 2023-08-05 manuscript/manuscript.Rmd
finished the introduction update.

@@
 
-The imposition of intended form on artifacts has long been viewed as a watershed in human cognitive and cultural evolution and is most commonly associated with the emergence of "Large Cutting Tools" (LCTs) in the Early Acheulean [@isaac1976; @kuhnEvolutionPaleolithicTechnologies2020; @holloway1969].
+The imposition of intended form on artifacts has long been viewed as a watershed in human cognitive and cultural evolution and is most commonly associated with the emergence of "Large Cutting Tools" (LCTs) in the Early Acheulean [@isaac1976; @kuhn2020; @holloway1969].
 However, this interpretation of Acheulean LCTs as intentionally designed artifacts remains controversial.
@@
 In later Acheulean contexts, reduction intensity effects are commonly equated with resharpening and seen as an alternative to intentional form imposition [@mcpherron2000].
-Across heavily-worked and relatively standardized LCT assemblages [e.g.
-@shipton2020], a *lack* of association between morphology and reduction intensity has be used as an argument-by-elimination for the presence of imposed morphological norms or \"mental templates\" [@garcía-medrano2019; @shipton2015; @shipton2023].
+Across heavily-worked and relatively standardized LCT assemblages [e.g., @shipton2020], a *lack* of association between morphology and reduction intensity has be used as an argument-by-elimination for the presence of imposed morphological norms or "mental templates" [@garcía-medrano2019; @shipton2015; @shipton2023].
 However, in the less heavily-worked and more heterogenous assemblages typical of the early Acheulean [@kuhn2020], it is equally plausible that increasing reduction intensity would reflect degree of primary reduction rather than subsequent resharpening [@archer2010].
@@
 
-Interpretive approaches address this quandary by \"reading\" the organization of scars on individual pieces to infer intent, but an adequate method to objectively quantify these insights has yet to be developed.
+Interpretive approaches address this quandary by "reading" the organization of scars on individual pieces to infer intent, but an adequate method to objectively quantify these insights has yet to be developed.
 Current measures of reduction intensity, such as the scar density index (SDI) [@clarkson2013], are designed to estimate total mass removed from a core and are reasonably effective [@lombao2023].
@@
 This is true whether the desired outcome is a useful flake, a rejuvenated edge, or a particular core morphology.
-In simple flake production, mass removed is probably a good reflection of the completeness of exploitation (\"exhaustion\") of cores and may have implications for required skill [@toth1982; @pargeter2023] as well as raw material economy [@reeves2021; @shick1987].
+In simple flake production, mass removed is probably a good reflection of the completeness of exploitation ("exhaustion") of cores and may have implications for required skill [@toth1982; @pargeter2023] as well as raw material economy [@reeves2021; @shick1987].
 However, in core shaping and resharpening, mass removal would typically represent an energetic and raw material cost to be minimized, and might even interfere with function [@keyHandaxeShapeVariation2019].
@@
 [-@li2015] proposed a Flaked Area Index (FAI) as an alternative to SDI as a measure of reduction intensity, arguing that its validity is supported by an observed correlation (r=0.424) with SDI.
-However, they also explain that \"flaked area does not necessarily relate to the number of flake scars\...a small number of large scars can produce a large area of scar coverage, and conversely, a large number of small scars can produce a small area of scar coverage.\" (p.
+However, they also explain that "flaked area does not necessarily relate to the number of flake scars\...a small number of large scars can produce a large area of scar coverage, and conversely, a large number of small scars can produce a small area of scar coverage." (p.
 6).
@@
 It is thus complementary to the measure of volume reduction provided by SDI and provides additional information to inform technological interpretations.
-For example, a correlation between FAI and artifact form without any effect of SDI would suggest a focus on \"least-effort\" shape imposition whereas the opposite pattern would be consistent with relatively intense resharpening of spatially restricted areas on the core.
-A lack of shape correlation with either measure would be expected for simple debitage with no morphological targets whereas a strong correlation with both would indicate a highly \"designed\" form achieved through extensive morphological and volumetric transformation.
+For example, a correlation between FAI and artifact form without any effect of SDI would suggest a focus on "least-effort" shape imposition whereas the opposite pattern would be consistent with relatively intense resharpening of spatially restricted areas on the core.
+A lack of shape correlation with either measure would be expected for simple debitage with no morphological targets whereas a strong correlation with both would indicate a highly "designed" form achieved through extensive morphological and volumetric transformation.
 In the current study we thus considered SDI and FAI together in order to evaluate evidence of intentional shaping in the early Acheulean of Gona.
@@
 
-Hypotheses: 1) Valid technological types should produce clear morphological clusters with different reduction trajectories vs. points along a continuum.
-2) Debitage is indicated by relation of SDI and flaked area to core size but not shape.
-3) Shaping is indicated by relation of flaked area to shape & weaker or absent relations of shape with SDI.
-Shape independent of size.
-4) Shaping plus resharpening means shape should be related to core size and SDI (Shipton)
+Three-dimensional scanning and geometric morphometric (3DGM) methods are increasingly common in the study of LCT form and reduction intensity [@shipton2015; @caruana2020; @presnyakova2018; @li2021; @archer2010; @li2015; @lycett2006].
+These methods provide high-resolution, coordinate-based descriptions of artifact form including detailed information about whole object geometric relations that is not captured by conventional linear measures [@shott2010].
+This includes measures of surface area used to compute both SDI and FAI measures [@clarkson2013; @li2015].
+At the time of writing, however, 3D scans are available for only a small number of Gona artifacts, including 33 of the Oldowan and Acheulean flaked pieces used in this study (**Figure** \@ref(fig:fig1)).
+Despite continuing improvements, 3DGM methods still impose additional costs in terms of data collection and processing time as well as required equipment, software, and training.
+Importantly, 3DGM methods cannot be applied to pre-existing photographic and metric data sets [e.g., @marshall], including available data from Gona.
+For this reason, and to better understand the relative costs and benefits of 3DGM more generally, we sought to test the degree to which conventional measurements can approximate 3DGM methods and produce reliable results by directly comparing our conventional measures with 3DGM analysis of the 33 available scans.
 
-It is even controversial whether asia is "acheulean" Prevailing opinion, but Beyene.
-A conservative interpretation of available evidence is that LCT production was guided by a recurring set of functional, ergonomic, and aesthetic design preferences [@wynn2018] with other elements free to vary in response to raw materials, use life, and random population dynamics like drift, bottlenecks, and founder effects [@kuhnEvolutionPaleolithicTechnologies2020; @lycett2016].
+```{r fig1, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="3D Artifact scans from Gona used in this study."}
+knitr::include_graphics(here::here("figure/Fig1.png"))
+```
+
+For our study, we are specifically concerned with the accurate description of morphological variation and estimation of artifact surface and flaked areas.
+With respect to morphology, we were encouraged by the fact that aspects of form variation relevant to our research question (i.e., core elongation, flattening, and pointedness) are relatively simple to describe using sparse data.
+Laser-scanning 3DGM studies of LCTs collect vast amounts of shape data but may discard upward of 50% of the observed variation in order to focus on two or three interpretable principal components.
+Across studies, these PCs consistently corresponding to basic features like elongation, relative thickness, pointedness, and position of maximum thickness that also emerge from lower-resolution spatial data [@garcía-medrano2019; @archer2010; @lycett2006] and studies employing linear measures rather than spatial coordinates [@pargeter2019; @crompton1993].
+There is less evidence that conventional methods can accurately estimate artifact surface and flaked areas.
+Clarkson [-@clarkson2013] advocated the use of 3D surface area measures as more accurate than estimation from linear measures (e.g., surface area of a rectangular prism defined by artifact dimensions).
+However, he also found that the error introduced by the linear approach was a highly systematic, isometric overestimation of surface area and that results correlated with direct 3D measures with an impressive r^2^ = 0.944 and no effect of variation in core shape.
+Insofar as it is variation in the relationship between surface area and flaking intensity that is of interest, rather than absolute artifact size, such systematic overestimation may not be problematic.
+Similar concerns apply to the estimation of flaked area.
+Traditionally, such estimates have been done \"by eye\" as a percentage of the total artifact surface [e.g., @dibble2005].
+Such estimation on have been found to be reasonably accurate when compared to 3D methods, but with the potential for substantial error on individual artifacts [@lin2010].
+The accuracy of by-eye estimation has yet to be systematically studied in Early Stone Age cores like those from Gona.
+
+## The Early Acheulean at Gona
+
+Early Acheulean sites in the Gona Project area are distributed over a wide area within the Dana Aoule North (DAN), Ounda Gona South (OGS), and Busidima North (BSN) drainages in the Busidima Formation [@quade2004] and range in age from approximately 1.7 to 1.2 mya [@semaw2018].
+The specific sites included in the current analysis are DAN-5, OGS-5, OGS-12, and BSN-17, all estimated to ca.
+1.7 -- 1.4 mya by stratigraphic position in the Gona sequence [@quade2008; @semaw2020].
+The Busidima Formation accumulated through fluvial deposition by the ancestral Awash River (Type I context) and its smaller tributary channels (Type II context) [@quade2004; @quade2008].
+Oldowan sites at Gona all occur in Type I sediments, indicating channel bank/margin (OGS-7) or proximal floodplain (EG-10, EG-12) contexts close to the large, hetero-lithic clasts available from point bars in the axial river channel [@quade2004; @stout2005].
+Acheulean sites continue to occur in Type I contexts (BSN-17, DAN-5) but are also found in Type II sediments (OGS-5, OGS-12,) reflecting increased utilization of large perennial tributaries to the ancestral Awash River [@quade2008].
+Clasts locally available in these tributaries were relatively small, implying that the large flakes and cobbles used to produce Acheulean artifacts were initially sourced from the axial river.
+A similar pattern of habitat diversification and increasing lithic transport distances has been described at other sites and may be typical of the early Acheulean [@rogers1994; @linaresmatás2021; @hay1976].
+As with other early (i.e.
+\>1.0 mya [@stoutStoneToolmakingEvolution2011; @presnyakova2018]) Acheulean assemblages, the Gona collections include numerous \"crudely made\" handaxes and picks on large flake blanks and cobbles, as well as large (\> 10cm) unmodified flakes, flaked pieces interpreted typo-technologically as Mode 1 cores (see Figure 1af), and smaller debitage [@semaw2018; @semaw2020].
 
@@
 We considered each in the context of our particular materials and research questions.
-First, conventional linear measures capture the direction (e.g. length \> breadth) but not the location of geometric relations (e.g. position of maximum breadth).
+First, conventional linear measures capture the direction (e.g., length \> breadth) but not the location of geometric relations (e.g., position of maximum breadth).
 We address this by collecting linear measures defined by homologous semi-landmarks.
@@
 Seven linear measures thus appear sufficient to explain technologically/behaviorally relevant shape variation in our sample.
-Third, linear measures may struggle to capture attributes such as cross-sectional area and shape (e.g. Caruana, 2020) more easily assessed using 3DGM.
+Third, linear measures may struggle to capture attributes such as cross-sectional area and shape (e.g., Caruana, 2020) more easily assessed using 3DGM.
 Particularly relevant here are measures of surface area used to calculate indices of reduction intensity [@shiptonHandaxeReductionIts2015a; @clarkson2013] and surface modification [@li2015] used in our study.
-Clarkson [-@clarkson2013] advocates the use of 3D surface area measures as more accurate than estimation from linear measures (e.g. surface area of a rectangular prism defined by artifact dimensions).
+Clarkson [-@clarkson2013] advocates the use of 3D surface area measures as more accurate than estimation from linear measures (e.g., surface area of a rectangular prism defined by artifact dimensions).
 However, he also found that the error introduced by the linear approach was a highly systematic, isometric overestimation of surface area and that results correlated with direct 3D measures with an impressive r2 = 0.944 and no effect of variation in core shape.
@@
 
-A PCA (covariance matrix) on our 7 linear measures (scaled by geometric mean) for pieces identified two PCs explaining 80% of variance (56.4% and 23.7%).
-Rescaled component matrix shows that PC1 reflects "flatness" (length and breadth vs. thickness).
-PC
-
-Two-step cluster analysis identified 3 clusters.
-
-Typologically, these loosely correspond to Mode 1 cores, Large Flake/Knifes, and Picks, with handaxes split between knife vs. pick categories.
-
-PC1 differentiates Mode 1 and Mode 2 pretty well, in that M1 cores tend not to be flat or elongated.
-Mode 1 exceptions (i.e.
-misclassified on shape) are generally still distinguishable as smaller and more heavily reduced than Mode 2 (of Mode 1 included in Cluster 1: mean weight =159.4 vs.
-635.6, p \< 0.001; Mean logSDI = .74 vs.
-.20, p \< 0.001).
-(of Mode 1 included in Cluster 3: mean weight =224.1 vs. 398.1, p \< 0.001; Mean logSDI = .67 vs. .39, p = 0.004).
-We thus treat Mode 1 as a valid techno-morphological category.
-Consistent with the characterization of Mode 1 as focused on debitage rather than shaping, we observe a strong power relationship between reduction intensity (SDI) and core size (r2=0.715, p \< 0.001, b1 = -0.872):
-
-In contrast, and also in keeping with a focus on debitage rather than shaping and resharpening, there is no such relationship with shape PCs for SDI:
-
-Cluster 1 is divided from Cluster 3 by PC2 (pointedness).
-Cluster 1 is much more likely to be executed on a flake base (91% flakes) vs. cluster 3 (35% flakes).
-Cluster 1 is also significantly less reduced (Mean logSDI = .39 vs. .20, p \< 0.001).
-So, cluster 1 basically comprises lightly retouched LFB acheulean, with shapes that remain largely within the range of unmodified flakes (n.s. mean difference).
-
-The effect of reduction on LFB acheulean shape is evident only for flaked area (not SDI) and corresponds to decreases in both PCs (i.e. less elongated but more pointy).
-The PC1 effect is relatively weak (r2=0.1, p =0.008, Standardized Beta = -0.215).
-The PC2 effect is stronger (r2=0.244, p \< 0.001, Standardized Beta = -0.537).
-This is most consistent with flaking placed to shape a point.
-A weak power effect of SDI on weight (r2=0.178, p \< 0.001, b1= -.330), as well as low number of scars in general, suggests resharpening is not a major factor.
-
-These trends mean that heavily modified flakes enter into cluster 3 (i.e. look like picks).
-Indeed, 40% of identifiable bases for cluster 3 are flakes.
-Cluster 3 pieces executed on flakes tend to be less pointed regardless of reduction intensity, which is likely a reflection of starting blank form.
-Indeed, Mode 2 Cobble bases show no effect of reduction intensity on shape but do show SDI effect on weight (r2=0.432, p \< 0.001, b1=-0.711).
-This appears to reflect the presence of cobble blanks that are already relatively pointed without substantial reduction and raises the possibility that these pieces are produced through debitage on pointed cobbles.
-Could they start as LFB cores?
-look at maximum flake scar size.
-Large cores have few, large scars.
-
-These patterns indicates that there is a common reduction trajectory for Mode 2 forms at Gona, regardless of typology or blank form.
-Although some pieces start much closer to the terminal morphology than others (i.e. display low PC2 values without substantial reduction), none undergo substantial reduction without becoming pointed.
-
-This uniform trajectory casts serious doubt on the likelihood that picks are a distinct morpho-functional type, although they may represent "4-dimensional design" sensu Kuhn.
-edge angles up to 70 degrees are quite efficient and obtuse trimming of butt may help ergonomics.
-
-No evidence for shaping of cobbles.
-
-Acheulean cobbles are larger and more cylindrical (geologically "rollers").
-Shape difference may reflect availability and/or selectivity for flake-able shapes.
-
-Some Achuelean flaked cobbles might hypothetically be heavily reduced remnants of giant cores for LFB production.
-Size of scars overlaps with LFBs (right).
-However, smaller scars are present and so reduction seems to have continued past potential for LFB production.
-Cobbles are generally pretty heavily reduced.
-
-Acheulean flaked flakes seem to have been (mildly) shaped to increase elongation and pointedness.
-This might have been an explicit design target, a passive result of preferentially flaking working edges, and or a desire to retain length for some reason.
-
 # References

### 2023-08-06 manuscript/manuscript.Rmd
added the method section.

@@
    - \usepackage{lineno}
+   - \usepackage[utf8]{inputenc}
    - \linenumbers
@@
 Across heavily-worked and relatively standardized LCT assemblages [e.g., @shipton2020], a *lack* of association between morphology and reduction intensity has be used as an argument-by-elimination for the presence of imposed morphological norms or "mental templates" [@garcía-medrano2019; @shipton2015; @shipton2023].
-However, in the less heavily-worked and more heterogenous assemblages typical of the early Acheulean [@kuhn2020], it is equally plausible that increasing reduction intensity would reflect degree of primary reduction rather than subsequent resharpening [@archer2010].
+However, in the less heavily-worked and more heterogeneous assemblages typical of the early Acheulean [@kuhn2020], it is equally plausible that increasing reduction intensity would reflect degree of primary reduction rather than subsequent resharpening [@archer2010].
 In this case, reduction intensity effects on morphology would have the opposite interpretation: more reduction should result in closer approximation of a desired form if such were present.
@@
 With respect to morphology, we were encouraged by the fact that aspects of form variation relevant to our research question (i.e., core elongation, flattening, and pointedness) are relatively simple to describe using sparse data.
-Laser-scanning 3DGM studies of LCTs collect vast amounts of shape data but may discard upward of 50% of the observed variation in order to focus on two or three interpretable principal components.
-Across studies, these PCs consistently corresponding to basic features like elongation, relative thickness, pointedness, and position of maximum thickness that also emerge from lower-resolution spatial data [@garcía-medrano2019; @archer2010; @lycett2006] and studies employing linear measures rather than spatial coordinates [@pargeter2019; @crompton1993].
+3DGM studies of LCTs collect vast amounts of shape data but may discard upward of 50% of the observed variation in order to focus on two or three interpretable principal components.
+Across studies, these PCs consistently correspond to basic features like elongation, relative thickness, pointedness, and position of maximum thickness that also emerge from lower-resolution spatial data [@garcía-medrano2019; @archer2010; @lycett2006] and studies employing linear measures rather than spatial coordinates [@pargeter2019; @crompton1993].
 There is less evidence that conventional methods can accurately estimate artifact surface and flaked areas.
@@
 Similar concerns apply to the estimation of flaked area.
-Traditionally, such estimates have been done \"by eye\" as a percentage of the total artifact surface [e.g., @dibble2005].
-Such estimation on have been found to be reasonably accurate when compared to 3D methods, but with the potential for substantial error on individual artifacts [@lin2010].
+Traditionally, such estimates have been done "by eye" as a percentage of the total artifact surface [e.g., @dibble2005].
+Such estimations have been found to be reasonably accurate when compared to 3D methods, but with the potential for substantial error on individual artifacts [@lin2010].
 The accuracy of by-eye estimation has yet to be systematically studied in Early Stone Age cores like those from Gona.
@@
 As with other early (i.e.
-\>1.0 mya [@stoutStoneToolmakingEvolution2011; @presnyakova2018]) Acheulean assemblages, the Gona collections include numerous \"crudely made\" handaxes and picks on large flake blanks and cobbles, as well as large (\> 10cm) unmodified flakes, flaked pieces interpreted typo-technologically as Mode 1 cores (see Figure 1af), and smaller debitage [@semaw2018; @semaw2020].
+\>1.0 mya [@stoutStoneToolmakingEvolution2011; @presnyakova2018]) Acheulean assemblages, the Gona collections include numerous "crudely made" handaxes and picks on large flake blanks and cobbles, as well as large (\> 10cm) unmodified flakes, flaked pieces interpreted typo-technologically as Mode 1 cores (see Figure 1af), and smaller debitage [@semaw2018; @semaw2020].
 
@@
 
-Archaeological Sample
+### Archaeological Sample
+
+Artifact collections analyzed here include *in situ* pieces excavated from intact stratigraphic contexts and surface pieces systematically collected from the sediments eroding from these layers.
+Surface pieces are included because the current technological analysis does not require more precise spatial association, stratigraphic, and chronological control.
+Our sample comprises the total collection of flaked pieces [@isaac1997] and large (\>10 cm) detached pieces from each site, regardless of typo-technological interpretation.
 
@@
 
-### Artifact Shape Measurement
+### Artifact Classification
+
+Artifacts were classified according to initial form (pebble/cobble, detached piece, or indeterminate), presence/absence of retouch, technological interpretation ("Mode 1" core vs.
+"Mode 2" LCT), and archaeological context (Oldowan vs.
+Early Acheulean sites).
+LCTs were additionally classified as handaxes, knives, or picks following definitions from Kleindienst [-@kleindienst1962].
+The validity of technological interpretations and typological classifications was assessed through cluster analysis based on artifact shape and reduction intensity variables.
+
+### Artifact Measurement
+
+Conventional linear measures capture the direction (e.g., length \> breadth) but not the location of geometric relations (e.g., position of maximum breadth).
+We address this by collecting linear measures defined by homologous semi-landmarks.
+All artifacts were oriented along their maximum dimension, which was measured and defined as "length." The next largest dimension orthogonal to length was used to define the plane of "breadth," with the dimension orthogonal to this plane defined as "thickness" Width (W~1~, W~2~, W~3~) and thickness (T~1~, T~2~, T~3~) measures were then collected at 25%, 50%, and 75% of length, oriented so that 25% Width \> 75% Width.
+To partition variation in shape from variation in size, we divided all linear measures by the geometric mean [@lycett2006].
+GM-transformed variables were then submitted to a Principal Components Analysis (covariance matrix) to identify the main dimensions of shape variation.
+
+Our semi-landmark measurement system allowed us to improve on the prism-based surface area formula (2LW + 2LT + 2 WT) by using our 7 recorded dimensions to more tightly fit three prisms around the artifact: SA = W~1~T~1~ + 2(.33L \* W~1~) + 2(.33L \* T~1~) + 2(.33L \* W~2~) + 2(.33L \* T~2~) + 2(.33L \* W~3~) + 2(.33L \* T~3~) + W~3~T~3~.
+Surface area calculated in this way correlates with mass^2/3^ at r^2^ = 0.947 in our sample.
+Calculated surface area was then used to derive the Scar Density Index: SDI = number of flake scars \> 1cm per unit surface area [@clarkson2013; @shiptonFlakeScarDensity2015].
+The Flaked Area Index (FAI: flaked area divided by total surface area)[@li2015] was estimated directly "by eye" as a percentage of the total artifact surface.
+
+### 3D Methods
+
+Artifact scans (N=33) were cleaned, smoothed, and re-meshed using MeshLab.
+The 3D triangular mesh of each scan was computed using Kazhdan and Hoppe's [-@kazhdan2013] method of screened Poisson surface reconstruction in MeshLab.
+3DGM analysis was conducted using the *AGMT3-D* program of Herzlinger and Grosman [-@herzlinger2018].
+Artifacts were automatically oriented according to the axis of least asymmetry, then manually oriented in the interests of standardization with the length, width, and thickness dimensions defined by the longest axis, followed by the next two longest axes perpendicular to the first.
+The wider end of the artifacts was positioned as the proximal end (base), and more protruding surfaces were oriented towards the user in the first orthogonal view.
+Then, a grid of 200 homologous semi-landmarks were overlain on each artifact's surface.
+Generalized Procrustes and Principal Component analyses were then undertaken to explore the shape variability of the sample.
+The surface area of each artifact was calculated using the *Artifact3-D* program of Grosman et al. [-@grosman2022].
+*Artifact3-D* was also used to automatically identify the flake scar boundaries and compute each scar's surface area, using the scar analysis functions of Richardson et al. [-@richardson2014].
+
+### Analyses
+
+Measurement Validation: To assess the adequacy of shape descriptions based on our linear measures, we directly compared these with shape as quantified by 3D methods on the 33 artifacts for which scans are available.
+GM-transformed linear measures from these 33 artifacts were submitted to a variance-covariance matrix PCA.
+PCs with an eigenvalue greater than the mean were retained for analysis and the results compared qualitatively (morphological interpretation of PCs) and quantitatively (correlation of artifact factor scores) to 3D results.
+Accuracy of surface area and flaked area estimates was also assessed by correlation with 3D results.
 

### 2023-08-07 manuscript/manuscript.Rmd
update the result section.

@@
 
-Three-dimensional scanning and geometric morphometric (3DGM) methods are becoming increasingly common in the study of LCT form [@archer2010; @caruana2020; @li2021; @lycett2006; @presnyakova2018; @shiptonHandaxeReductionIts2015a].
-These methods can provide high-resolution, coordinate-based descriptions of artifact form including detailed information about whole object geometric relations that is not captured by conventional linear measures [@shott2010].
-However, they also impose additional costs in terms of data collection and processing time as well as required equipment, software, and training.
-Insofar as these costs might present an obstacle to participation by some researchers and/or draw resources away from other activities, they must be balanced against benefits.
-In particular, it is not clear that these powerful methods are required in order to describe relevant variation in Acheulean LCT shape.
-Unlike hominin crania or even projectile points, Acheulean handaxes, cleavers, and picks are not complex shapes.
-Individual LCTs exhibit complex morphologies defined by idiosyncratic scar patterns, but these details are largely noise at the level of comparative analyses.
-Laser-scanning 3DGM studies of LCTs collect vast amounts of shape data, but typically discard upward of 50% of the observed variation in order to focus on two or three interpretable principal components.
-Across studies, these PCs consistently corresponding to basic features like elongation, relative thickness, pointedness, and position of maximum thickness that also emerge from lower-resolution spatial data [@archer2010; @garcía-medrano2019; @lycett2006] and studies employing linear measures rather than spatial coordinates [@crompton1993; @pargeter2019].
-Thus, while the level of detail enabled by 3DGM is arguably useful for building artifact phylogenies [@okumura2019], it is of questionable behavioral/technological relevance for the study of LCTs.
-For these reasons, we favored the use of simple caliper-based linear measures to quantify shape in our study.
-Nevertheless, Shott and Trail [-@shott2010] do identify three potential shortcomings of linear measurements compared to 3DGM.
-We considered each in the context of our particular materials and research questions.
-First, conventional linear measures capture the direction (e.g., length \> breadth) but not the location of geometric relations (e.g., position of maximum breadth).
-We address this by collecting linear measures defined by homologous semi-landmarks.
-All artifacts were oriented along their maximum dimension, which was measured and defined as "length." The next largest dimension orthogonal to length was used to define the plane of "breadth," with the dimension orthogonal to this plane defined as "thickness." Breath and thickness measures were then collected at 25%, 50%, and 75% of length, oriented so that 25% Breadth \> 75% Breath.
-To partition variation in shape from variation in size, we divided all linear measures by the geometric mean [@lycett2006].
-
-Second, linear measures risk reducing complex forms to overly simplistic "stick figure caricatures" [@shott2010].
-However, whether or not this risk actually presents a problem depends on the particular artifacts and research questions involved.
-We have already noted that 3DGM LCT studies typically evaluate only a small portion of the measured variation.
-To better evaluate the measurement density required for our study, we reanalyzed a data set of 128 experimental handaxes previously published by Pargeter et al. [-@pargeter2019].
-These data comprise 19 linear measures (length plus breadth and thickness at 10% increments of length) collected from digital photos using the same orientation protocol described above.
-We conducted a PCA on the full set of 19 measures and again on a reduced set of 7 (length plus breadth and thickness at 30%, 50%, and 70% length).
-Despite this reduction, the first two components from each analysis displayed strikingly similar component loading matrices (PC1 positive on length and tip breadth, negative on thickness; PC2 positive on base breadth, negative on length and thickness) almost perfectly correlated component scores for individual pieces (PC1 r=0.919, PC2 r=0.913).
-As a further check, we performed the same comparison on a subset of the current archaeological sample from Gona for which photos were available for measurement (n = 50).
-This produced two PCs that were not only similar with each other, but also matched the PCs extracted from the experimental handaxe sample.
-Individual piece component scores were again highly correlated (r=0.975 and 0.927 respectively).
-Seven linear measures thus appear sufficient to explain technologically/behaviorally relevant shape variation in our sample.
-Third, linear measures may struggle to capture attributes such as cross-sectional area and shape (e.g., Caruana, 2020) more easily assessed using 3DGM.
-Particularly relevant here are measures of surface area used to calculate indices of reduction intensity [@shiptonHandaxeReductionIts2015a; @clarkson2013] and surface modification [@li2015] used in our study.
-Clarkson [-@clarkson2013] advocates the use of 3D surface area measures as more accurate than estimation from linear measures (e.g., surface area of a rectangular prism defined by artifact dimensions).
-However, he also found that the error introduced by the linear approach was a highly systematic, isometric overestimation of surface area and that results correlated with direct 3D measures with an impressive r2 = 0.944 and no effect of variation in core shape.
-Insofar as it is variation in the relationship between surface area and flaking intensity that is of interest, rather than the absolute size of artifacts, such consistent overestimation is not problematic.
-Here we improved on the prism-based surface area formula (2LW + 2LT + 2 WT) by using our 7 recorded dimensions to more tightly fit three prisms (Figure 1) around the artifact: SA = W1T1 + 2(.33L \* W1) + 2(.33L \* T1) + 2(.33L \* W2) + 2(.33L \* T2) + 2(.33L \* W3) + 2(.33L \* T3) + W3T3.
-Surface area calculated in this way correlates with mass2/3 at r2 = 0.947 in our sample.
-
-PCA on GM-transformed caliper measures (length, 3 breadth, 3 thickness).
-Length is maximum dimension, piece oriented so that Br1\>Br3
-
-Typological and technological attributions considered unreliable.
-Data grouped according to context (\~2.5 mya Oldowan sites vs. \~1.5 mya Acheulean sites) and blank form (cobble, flake, indeterminant).
-
-Associations between form and reduction intensity are considered as an indicator of "imposed form." Such form could reflect mental templates and/or biased flaking patterns due to functional or technological constraints
-
-### Reduction Indices
-
-Research by Clarkson and Shipton has established the Scar Density Index (SDI = number of flake scars \> 1cm per unit surface area) as a reliable indicator of mass removed from a core across technologies [@clarkson2013] and for handaxes specifically [@shiptonFlakeScarDensity2015].
-We thus use SDI as an indicator of reduction intensity (mass removed) in our study.
-However, reduction intensity does not constitute a full description of core modification.
-Mass removal is the aim during flake production and extent of shaping are not necessarily the same thing.
-For example, imposition of a desired form
-
 # Results
 
+## Measurement Validation
+
+A PCA on GM transformed linear measures the 33 artifacts identified two PCs accounting for 75.6% of the variance.
+Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance.
+Factor loadings (Table 1) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs.
+distal width and thickness).
+This A PCA on GM transformed linear measures the 33 artifacts identified two PCs accounting for 75.6% of the variance.
+Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance.
+Factor loadings (Table 1) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs.
+distal width and thickness).
+This closely parallels the length vs.
+width and thickness tradeoff captured by 3DGM-PC1 (Figure 3a) and is reflected in a tight correlation of artifact scores produced by the two PCs (r = 0.903, p < 0.001, Figure 3c).
+Comparison with conventional shape ratios (Table 2) similarly indicates that both Linear-PC1 and 3DGM-PC1 largely capture variation in artifact elongation.
+A second factor (Linear-PC2, eigenvalue = 0.07) explained an additional 20.4% of variance.
+This factor was less strongly correlated with its 3DGM counterpart (3DGM-PC2; r = 0.344, p = 0.050) probably because Linear-PC2 describes anticorrelated variation in width and thickness (i.e., broad and flat vs.
+thick and pointed; Tables 1 & 2) whereas 3DGM-PC2 more purely isolates convergence (Table 2).
+The remainder of the shape variability explained by Linear-PC2 is captured by higher order 3DGM-PCs like 3DGM-PC3 to 5, which comprise the contribution of the left and right lateral margins to relative thickness.
+Including a third dimension in the GM analysis generates PCs that identify more specific shape attributes, but the underlying morphological variability captured by both the 2D and 3D analyses remains similar.
+Together, 3DGM-PC2 (r = 0.344, p = 0.050), 3DGM-PC3 (r = -0.416, p = 0.016), 3DGM-PC4 (r = 0.458, p = 0.007), and 3DGM-PC5 (r = -0.352, p = 0.044) correlate well with Linear-PC2, cumulatively capturing whether the items are broad and flat or thick and pointed.
+
 # References

### 2023-08-08 manuscript/manuscript.Rmd
reference updates.

@@
 Even accepting that LCT form was to some extent intended, there is substantial disagreement over the specificity of design.
-Some analyses have indicated that shape variation in Acheulean handaxes is largely a result of resharpening [@iovita2011; @mcpherron2000] whereas others find form to be unrelated to reduction intensity and more likely to reflect normative expectations of what handaxes should look like [@garcía-medrano2019; @shiptonHandaxeReductionIts2015a; @shipton2020].
-Such debates about shape of Acheulean LCTs may appear narrowly technical but have broad relevance for evolutionary questions including the origins of human culture [@corbey2016; @tennie2017; @shiptonHandaxeReductionIts2015a], language [@stout2012], teaching [@gärdenfors2017], brain structure [@hecht2015], and cognition [@stout2015; @wynn2016].
-To examine these questions, we studied the complete collection of Early Acheulean flaked pieces from 5 sites at Gona Project Area and compared them with Oldowan cores form 2 published sites at Gona.
+Some analyses have indicated that shape variation in Acheulean handaxes is largely a result of resharpening [@iovita2011; @mcpherron2000] whereas others find form to be unrelated to reduction intensity and more likely to reflect normative expectations of what handaxes should look like [@garcía-medrano2019; @shipton2015; @shipton2020].
+Such debates about shape of Acheulean LCTs may appear narrowly technical but have broad relevance for evolutionary questions including the origins of human culture [@corbey2016; @tennie2017; @shipton2015], language [@stout2012], teaching [@gärdenfors2017], brain structure [@hecht2015], and cognition [@stout2015; @wynn2016].
+To examine these questions, we studied the complete collection of Early Acheulean flaked pieces from 5 sites at Gona Project Area and compared them with Oldowan cores from 2 published sites at Gona.
 By comparing shape variation to measures of flaking intensity and patterning, we sought to identify technological patterns that might reveal intent.

### 2023-08-14 manuscript/manuscript.Rmd
added table 1.

@@
 Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance.
-Factor loadings (Table 1) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs.
+Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs.
 distal width and thickness).
@@
 
+```{r tab1, echo=FALSE, message=FALSE}
+library(tidyverse)
+df1 <- data.frame("Linear metrics (GM transformed)" = c("Length", "W1", "W2", "W3", "T1", "T2", "T3"), 
+                 "Linear-PC1" = c("0.989", "0.303", "0.403", "-0.176", "-0.135", "-0.369", "-0.607"), 
+                 "Linear-PC2" = c("-0.107", "0.350", "0.767", "0.790", "-0.679", "-0.623", "-0.282") ) 
+kableExtra::kbl(df1, booktabs = T, caption = "Component loadings for linear metric PCs .") 
+```
 # References

### 2023-09-24 manuscript/manuscript.Rmd
results update.

--- 2023-08-08 manuscript/manuscript.Rmd
@@
 
+```{r tab1, echo=FALSE, message=FALSE}
+library(tidyverse)
+df1 <- data.frame("Linear metrics (GM transformed)" = c("Length", "W1", "W2", "W3", "T1", "T2", "T3"), 
+                 "Linear-PC1" = c("0.989", "0.303", "0.403", "-0.176", "-0.135", "-0.369", "-0.607"), 
+                 "Linear-PC2" = c("-0.107", "0.350", "0.767", "0.790", "-0.679", "-0.623", "-0.282") ) 
+kableExtra::kbl(df1, booktabs = T, caption = "Component loadings for linear metric PCs .") 
+```
+
+##  Classification Validation
+
+We first conducted a stepwise DFA on all flaked pieces (n=192) with inferred technological Mode (one vs.
+two) as the grouping variable and PCs 1 and 2, corrected SDI (cSDI), and FAI as the independent variables.
+All variables were retained, yielding one canonical DF (eigenvalue=1.825, Wilks Lambda = 0.354, p< 0.001) which correctly classified 93.8% of artifacts.
+We thus accepted the validity of classification by Mode in our sample and employed this distinction in subsequent analyses.
+There was no discernable difference in discriminant scores for Mode 1 cores from Oldowan (n=37) vs.
+Acheulean (n=39) contexts (, p = 0.746).
+Mode 1 cores from Oldowan contexts (n=37) do include 10 (27%) small, retouched flakes.
+Only one retouched flake from an Acheulean context was classified by the DFA as a Mode 1 core.
+This piece, typologically classified as a “knife”, is the smallest (93mm) retouched flake in the Acheulean sample.
+When retouched flakes are excluded from the comparison, there are no significant differences in shape between Mode 1 cores from Oldowan and Achuelean contexts.
+Interestingly, however, Acheulean Mode 1 cobble-cores are much larger (mean weight 480g vs.
+186g, Cohen’s d=1.137, p<0.001) and less heavily reduced (mean cSDI 0.057 vs.
+0.103, Cohen’s d = -0.884, p < 0.001) despite having similar FAI (mean 0.52 vs.
+0.46, Cohen’s d=0.271, p=0.266).
+
+Next, we conducted a stepwise DFA on all flaked Mode 2 pieces (i.e.
+excluding unmodified large flakes, n = 115) with typology (handaxe, pick, knife) as the grouping variable and the same four independent variables.
+Both shape PCs and FAI were retained, while cSDI was not entered.
+This produced two DFs (DF1: Eigenvalue=1.536, 91.3% of variance; DF2: Eigenvalue = 0.146, 8.7% of variance; Wilks Lambda = 0.344; p<0.001) which correctly classified 71.3% of artifacts.
+Inspection of DF coefficients shows that DF1 captures an inverse correlation between flaked area and pointedness (Linear-PC2) whereas DF2 captures a positive relationship between flaked area and elongation (Linear-PC2).
+A bivariate plot (Figure 5) illustrates the fact that DF1 captures a range of variation from pointed, heavily-worked picks, through handaxes, to knives, with substantial overlap between adjacent types (Table 2).
+As an intermediate type, “handaxes” were correctly classified only 43.9% of the time.
+In agreement with others (Duke, et al., 2021, Kuhn, 2021, Presnyakova, et al., 2018) we conclude that these typological labels artificially partition a continuum of variation and abandon them in subsequent analyses.
+
+##  Effects of Reduction on Shape
+To assess the influence of flake removals on core form, we examined the association between our reduction measures (cSDI and FAI) and core shape (PC1, PC2).
+In the complete sample of flaked pieces (n=192), we observed weak but significant effects of cSDI (r=-0.294, p <0.001) and FAI (r=-0.294, p<0.001) on PC1(flatness) and of FAI only on PC2(pointedness) (r=-0.436, p<0.001).
+However, it is clear that these overall effects conflate different trends in Mode 1 vs.
+Mode 2 cores, as well as in cores executed on flake vs.
+pebble/cobble bases.
+Within categories, we observed no significant effects of cSDI whereas FAI had variable relationships with core form.
+
+In Mode 1 cores, a weak negative effect of FAI on PC2(pointedness) (n = 76, $\beta$ =-0.006, r=0.240, p= 0.037) suggests that increasing extent of modification tends to slightly increase pointedness (Figure 6a).
+No effects of FAI on PC1 (flatness) or of SDIc on either PC approach significance.
+FAI and SDIc are moderately correlated ($\beta$=0.001, r=0.524, p<0.001), an effect that reflects increases in the upper limit of scar density values as flaked area increases (i.e., flaked area constrains maximum cSDI: Figure 6b).
+
+In Mode 2 cores (n=115), reduction measures have different effects depending on initial blank form.
+In general, the contrast between flat, lightly-worked flake bases and rounder, more-heavily worked cobble bases tends to inflate relationships between FAI and core shape (Figure 7c,d).
+Mode 2 cobble cores (n=37) display a negative effect of FAI on PC1(Flatness) ($\beta$=-0.011, r=0.367, p = 0.026), no effect of FAI on pointedness, and no association between FAI and cSDI (r=0.141, p = 0.406).
+Flake cores (n=69) show negative effects of FAI on PC1(Flatness) ($\beta$=-0.007, r=0.284, p = 0.018) and, more strongly, PC2(pointedness) ($\beta$=-0.019, r=0.443, p < 0.001).
+Flake cores show a small but significant positive effect of FAI on cSDI that, in contrast to Mode 1 cores, represents actual covariance rather than simply a relaxation of constraint (i.e.
+both upper and lower limits of cSDI increase with FAI).
+Comparison of Mode 2 flake cores with unmodified large flakes from Achuelean contexts (n=35) shows that ranges of shape variation substantially overlap but that flaking generally reduces both PC1(flatness) and PC2(pointedness).
+In fact, regressions of PC1 and PC2 on FAI show y-intercepts that closely approximate unmodified flake mean values (0.852 vs.
+0.774 and 0.693 vs.
+0.594, respectively) and significantly negative slopes as FAI increases (Figure 9).
+
+A small number of Mode 2 cores (n=9) were so heavily modified (mean FAI=89%, range 65-100) that the blank form could not be determined.
+These heavily worked “indeterminate” cores tend to show similar PC1(flatness) and PC2(pointedness) values to cobble-base cores (i.e.
+thicker and more pointed than flake cores; Figure x).
+They also appear to follow general trends of shape change with increasing FAI observed across base types (Figure 10).
+
 # References

### 2023-09-25 manuscript/manuscript.Rmd
results and disucssion updates.

@@
 Such forms continue to occur with variable frequency in later time periods [@mcnabb2015], and may be especially prevalent in eastern Asia [@li2021].
-Although formal types have been recognized in the Early Acheulean and are commonly used to describe assemblages, many workers now see a continuum of morphological variation [@duke2021; @kuhn2020; @presnyakova2018]including the possibility that simple flake production remained an important [@shea2010] or even primary [@moore2016] purpose of Early Acheulean large core reduction.
+Although formal types have been recognized in the Early Acheulean and are commonly used to describe assemblages, many workers now see a continuum of morphological variation [@duke2021; @kuhn2020; @presnyakova2018] including the possibility that simple flake production remained an important [@shea2010] or even primary [@moore2016] purpose of Early Acheulean large core reduction.
 
@@
 Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance.
-Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs.
-distal width and thickness).
+Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness).
 This A PCA on GM transformed linear measures the 33 artifacts identified two PCs accounting for 75.6% of the variance.
 Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance.
-Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs.
-distal width and thickness).
+Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness).
 This closely parallels the length vs.
-width and thickness tradeoff captured by 3DGM-PC1 (Figure 3a) and is reflected in a tight correlation of artifact scores produced by the two PCs (r = 0.903, p < 0.001, Figure 3c).
+width and thickness tradeoff captured by 3DGM-PC1 (Figure 3a) and is reflected in a tight correlation of artifact scores produced by the two PCs (r = 0.903, p \< 0.001, Figure 3c).
 Comparison with conventional shape ratios (Table 2) similarly indicates that both Linear-PC1 and 3DGM-PC1 largely capture variation in artifact elongation.
@@
 This factor was less strongly correlated with its 3DGM counterpart (3DGM-PC2; r = 0.344, p = 0.050) probably because Linear-PC2 describes anticorrelated variation in width and thickness (i.e., broad and flat vs.
-thick and pointed; Tables 1 & 2) whereas 3DGM-PC2 more purely isolates convergence (Table 2).
+thick and pointed; Table 1) whereas 3DGM-PC2 more purely isolates convergence (Figure 3a).
 The remainder of the shape variability explained by Linear-PC2 is captured by higher order 3DGM-PCs like 3DGM-PC3 to 5, which comprise the contribution of the left and right lateral margins to relative thickness.
-Including a third dimension in the GM analysis generates PCs that identify more specific shape attributes, but the underlying morphological variability captured by both the 2D and 3D analyses remains similar.
-Together, 3DGM-PC2 (r = 0.344, p = 0.050), 3DGM-PC3 (r = -0.416, p = 0.016), 3DGM-PC4 (r = 0.458, p = 0.007), and 3DGM-PC5 (r = -0.352, p = 0.044) correlate well with Linear-PC2, cumulatively capturing whether the items are broad and flat or thick and pointed.
+Use of high-resolution, coordinate-based scan data thus generates PCs that identify more specific shape attributes, but the underlying morphological variability captured by the linear and 3D analyses remains similar.
+Together, 3DGM-PC2 (r = 0.344, p = 0.050), 3DGM-PC3 (r = -0.416, p = 0.016), 3DGM-PC4 (r = 0.458, p = 0.007), and 3DGM-PC5 (r = -0.352, p = 0.044) correlate well with Linear-PC2 (Figure 3c), cumulatively capturing whether the items are broad and flat or thick and pointed.
+A stepwise regression (r2=0.625, F(4,28)=11.697, p\<0.001, Probability-of-F-to-enter \<= 0.050; Probability-of-F-to-remove \>= 0.100) with Linear-PC2 as the dependent variable retained all four of these 3DGM-PCs as significant predictors.
 
@@
 
+We thus concluded that linear measures are adequate to capture relevant shape variation and proceeded with a PCA on our full sample.
+We identified two PCs accounting for 80.0% of the variance.
+PC1 (Eigenvalue = 0.216) explained 56.4% of the variance.
+Factor loadings (Table 2) for PC1 reflect artifact flatness (i.e., an anti-correlation of length and width vs. thickness) such that higher values indicate relatively thinner pieces.
+PC2 (Eigenvalue = 0.090) explained 23.6% of the variance.
+Factor loadings (Table 2) show that PC2 captures artifact convergence or "pointedness" (i.e., an anti-correlation of tip width with length and butt thickness) such that higher values indicate shorter, less pointed forms.
+
+We also tested the validity of our two reduction measures, SDI and FAI.
+In agreement with Clarkson (2013), we found that surface area estimated from caliper measures displayed a strong correlation with (r2=0.975, p \< 0.001) but linear over-estimation of ($\beta$ = 1.58) 3DGM surface area (Figure 4a).
+This results in a systematic underestimation of SDI that scales with core size (Figure 4b).
+However, a simple correction of the caliper estimate (dividing by the slope, 1.58) eliminates surface area over-estimation (mean difference = 256mm2 [\<1.7% of mean], p=0.040) and produces SDI values that agree with 3DGM (r2=0.975, p \< 0.001, $\beta$ = 0.98) (Figure 4c).
+We thus proceeded to apply this correction to surface area estimates in the full sample.
+Insofar as these relationships are driven by basic geometry, we expect these methods (including correction) to be generalizable to other ESA assemblages.
+
 ## Classification Validation
@@
 two) as the grouping variable and PCs 1 and 2, corrected SDI (cSDI), and FAI as the independent variables.
-All variables were retained, yielding one canonical DF (eigenvalue=1.825, Wilks Lambda = 0.354, p< 0.001) which correctly classified 93.8% of artifacts.
+All variables were retained, yielding one canonical DF (eigenvalue=1.825, Wilks Lambda = 0.354, p\< 0.001) which correctly classified 93.8% of artifacts.
 We thus accepted the validity of classification by Mode in our sample and employed this distinction in subsequent analyses.
-There was no discernable difference in discriminant scores for Mode 1 cores from Oldowan (n=37) vs.
-Acheulean (n=39) contexts (, p = 0.746).
+There was no discernable difference in discriminant scores for Mode 1 cores from Oldowan (n=37) vs. Acheulean (n=39) contexts (, p = 0.746).
 Mode 1 cores from Oldowan contexts (n=37) do include 10 (27%) small, retouched flakes.
 Only one retouched flake from an Acheulean context was classified by the DFA as a Mode 1 core.
-This piece, typologically classified as a “knife”, is the smallest (93mm) retouched flake in the Acheulean sample.
+This piece, typologically classified as a "knife", is the smallest (93mm) retouched flake in the Acheulean sample.
 When retouched flakes are excluded from the comparison, there are no significant differences in shape between Mode 1 cores from Oldowan and Achuelean contexts.
 Interestingly, however, Acheulean Mode 1 cobble-cores are much larger (mean weight 480g vs.
-186g, Cohen’s d=1.137, p<0.001) and less heavily reduced (mean cSDI 0.057 vs.
-0.103, Cohen’s d = -0.884, p < 0.001) despite having similar FAI (mean 0.52 vs.
-0.46, Cohen’s d=0.271, p=0.266).
+186g, Cohen's d=1.137, p\<0.001) and less heavily reduced (mean cSDI 0.057 vs.
+0.103, Cohen's d = -0.884, p \< 0.001) despite having similar FAI (mean 0.52 vs.
+0.46, Cohen's d=0.271, p=0.266).
 
@@
 Both shape PCs and FAI were retained, while cSDI was not entered.
-This produced two DFs (DF1: Eigenvalue=1.536, 91.3% of variance; DF2: Eigenvalue = 0.146, 8.7% of variance; Wilks Lambda = 0.344; p<0.001) which correctly classified 71.3% of artifacts.
+This produced two DFs (DF1: Eigenvalue=1.536, 91.3% of variance; DF2: Eigenvalue = 0.146, 8.7% of variance; Wilks Lambda = 0.344; p\<0.001) which correctly classified 71.3% of artifacts.
 Inspection of DF coefficients shows that DF1 captures an inverse correlation between flaked area and pointedness (Linear-PC2) whereas DF2 captures a positive relationship between flaked area and elongation (Linear-PC2).
 A bivariate plot (Figure 5) illustrates the fact that DF1 captures a range of variation from pointed, heavily-worked picks, through handaxes, to knives, with substantial overlap between adjacent types (Table 2).
-As an intermediate type, “handaxes” were correctly classified only 43.9% of the time.
-In agreement with others (Duke, et al., 2021, Kuhn, 2021, Presnyakova, et al., 2018) we conclude that these typological labels artificially partition a continuum of variation and abandon them in subsequent analyses.
+As an intermediate type, "handaxes" were correctly classified only 43.9% of the time.
+In agreement with others [@duke2021; @kuhn2020; @presnyakova2018] we conclude that these typological labels artificially partition a continuum of variation and abandon them in subsequent analyses.
 
 ## Effects of Reduction on Shape
+
 To assess the influence of flake removals on core form, we examined the association between our reduction measures (cSDI and FAI) and core shape (PC1, PC2).
-In the complete sample of flaked pieces (n=192), we observed weak but significant effects of cSDI (r=-0.294, p <0.001) and FAI (r=-0.294, p<0.001) on PC1(flatness) and of FAI only on PC2(pointedness) (r=-0.436, p<0.001).
-However, it is clear that these overall effects conflate different trends in Mode 1 vs.
-Mode 2 cores, as well as in cores executed on flake vs.
-pebble/cobble bases.
+In the complete sample of flaked pieces (n=192), we observed weak but significant effects of cSDI (r=-0.294, p \<0.001) and FAI (r=-0.294, p\<0.001) on PC1(flatness) and of FAI only on PC2(pointedness) (r=-0.436, p\<0.001).
+However, it is clear that these overall effects conflate different trends in Mode 1 vs. Mode 2 cores, as well as in cores executed on flake vs. pebble/cobble bases.
 Within categories, we observed no significant effects of cSDI whereas FAI had variable relationships with core form.
@@
 No effects of FAI on PC1 (flatness) or of SDIc on either PC approach significance.
-FAI and SDIc are moderately correlated ($\beta$=0.001, r=0.524, p<0.001), an effect that reflects increases in the upper limit of scar density values as flaked area increases (i.e., flaked area constrains maximum cSDI: Figure 6b).
+FAI and SDIc are moderately correlated ($\beta$=0.001, r=0.524, p\<0.001), an effect that reflects increases in the upper limit of scar density values as flaked area increases (i.e., flaked area constrains maximum cSDI: Figure 6b).
 
@@
 Mode 2 cobble cores (n=37) display a negative effect of FAI on PC1(Flatness) ($\beta$=-0.011, r=0.367, p = 0.026), no effect of FAI on pointedness, and no association between FAI and cSDI (r=0.141, p = 0.406).
-Flake cores (n=69) show negative effects of FAI on PC1(Flatness) ($\beta$=-0.007, r=0.284, p = 0.018) and, more strongly, PC2(pointedness) ($\beta$=-0.019, r=0.443, p < 0.001).
+Flake cores (n=69) show negative effects of FAI on PC1(Flatness) ($\beta$=-0.007, r=0.284, p = 0.018) and, more strongly, PC2(pointedness) ($\beta$=-0.019, r=0.443, p \< 0.001).
 Flake cores show a small but significant positive effect of FAI on cSDI that, in contrast to Mode 1 cores, represents actual covariance rather than simply a relaxation of constraint (i.e.
 both upper and lower limits of cSDI increase with FAI).
+
 Comparison of Mode 2 flake cores with unmodified large flakes from Achuelean contexts (n=35) shows that ranges of shape variation substantially overlap but that flaking generally reduces both PC1(flatness) and PC2(pointedness).
@@
 A small number of Mode 2 cores (n=9) were so heavily modified (mean FAI=89%, range 65-100) that the blank form could not be determined.
-These heavily worked “indeterminate” cores tend to show similar PC1(flatness) and PC2(pointedness) values to cobble-base cores (i.e.
-thicker and more pointed than flake cores; Figure x).
+These heavily worked "indeterminate" cores tend to show similar PC1(flatness) and PC2(pointedness) values to cobble-base cores (i.e. thicker and more pointed than flake cores; Figure x).
 They also appear to follow general trends of shape change with increasing FAI observed across base types (Figure 10).
 
+# Discussion
+
+Following the criteria we proposed in the Introduction, Large Flake base cores from Acheulean contexts show evidence of design.
+Importantly, "design" in this context is understood as "design in four dimensions" sensu Kuhn [-@kuhn2020].
+Increasing FAI is associated with progressive narrowing ( i.e., decreasing PC1) and convergence (decreasing PC2) whereas cSDI is not.
+This indicates that the size and non-overlapping placement of flake removals drives shape change, rather than simply the removal of mass in general as might be expected from conventional models of resharpening (?).
+The fact that cSDI nevertheless correlates with FAI is consistent with relatively light (number of scars per piece ranges from 2 to 17 with a mean of 6.94) and spatially distributed re-touching involving relatively little overlap between scars.
+Such a pattern could be produced by intentional imposition of a design target or "mental template" or a more procedural/functional bias in favor of establishing cutting edges at one end of the long axis.
+In the absence of effective thinning/volume management techniques, this will lead to rapid narrowing/thickening of the piece overall and especially at the "tip." This is consistent with differences between flaked vs.
+unmodified large flakes and furthers the point that blank production already can be conceived as a process of "design" -- many unmodified flakes may have been similar in shape and just as useful as many retouched tools.
+
+This contrasts with reduction of Mode 1 cores.
+Unexpectedly, FAI did have an effect to increase convergence.
+However, this effect was substantially weaker than seen in LFB Acheulean cores and may simply reflect the fact that reduction is unlikely to be perfectly symmetrical and our orienting protocol guarantees that one end is always more pointed than the other unless they are precisely equal in breath.
+In support of this "byproduct" interpretation, the relationship between FAI and cSDI is essentially what one would expect from knapping unconstrained by any systematic core-shaping strategy (greater flaked area can, but does not necessarily, accommodate a greater number of flake scars)
+
+Our most complicated results come from Mode 2 cobble cores.
+Evidence of intentional shaping is limited: FAI has a moderate effect on PC1(flatness) and no effect on pointedness.
+Furthermore, the is no association between FAI and cSDI cobble cores display a greater range of cSDI values.
+Instead, FAI is strongly correlated with scar size (mean scar length: n=33, r=.588, P\<0.001; Figure 11).
+Together, this suggests that the removal of large flakes in particular increases FAI and tends to narrow the piece.
+We propose that this pattern may actually indicate that at least some cobble cores were precisely that: cores for the generation of LFBs.
+Given the difficulty of working relatively spherical cobbles, we expect that knappers would have chosen elongated and/or tabular cobbles.
+Attempts to maximize potential flake dimensions would naturally lead to removals along the long axis rather than at either end.
+This hypothesis was suggested by informal "experiential" knapping by the lead author with Gona cobbles, during which LFB production produced pick like forms as a byproduct.
+In support of the plausibility of this hypothesis, the maximum scars length on Gona Mode 2 cobble cores substantially overlaps with the size of unmodified large flakes from Acheulean contexts.
+It is also possible that cobble cores might begin as cores for LFB production and then be finished into core tools in their own right.
+These possibilities have major implications for understanding the occurrence and apparent morphological stability [@beyene2013] of "picks" throughout the African Acheuelan but remains to be tested.
+
 # References

### 2023-11-29 code/Analysis.R
plotting.

@@ -4,6 +4,8 @@ library(factoextra)
 library(FactoMineR)
 library(PCAmixdata)
 library(psych)
+library(patchwork)
+library(figpatch)
 # Add data
 gona_esa<-read.csv("data/EA_Reduction_Paper_Order.csv")
 
@@ -62,3 +64,112 @@ ggstatsplot::ggscatterstats(
   ylab  = "3DGMM PC2",
 )
 ggplot2::ggsave("3D PC2 correlation.png", path="figure.", width = 5, height = 5, dpi = 600)
+
+
+
+
+#####################################
+gona_esa<-read.csv("data/EA_Reduction_Paper_2023_cleaned_and_corrected_for_CL.csv")
+## creating a new column of unique identifier for merging
+gona_esa$Name <- paste(gona_esa$Site, gona_esa$Catalog, sep= "-")
+gona_3dgmm<-readr::read_csv("data/Gona3Ddata_GMM.csv")
+gona_scars<-readr::read_csv("data/Gona3Ddata_scars.csv")
+## Subsetting scar data frame because starting from th 7th column it lists the area of individual scars on an artifact, the number of which vary between artifacts.
+gona_scars_6c<-gona_scars[1:6,]
+joined_gona <-dplyr::left_join(gona_esa, gona_3dgmm, by = c("Name" = "Name"))
+full_gona <-dplyr::left_join(joined_gona, gona_scars_6c, by = c("Name" = "Name"))
+## Subsetting a dataframe that contains both capliper and 3d measurements.
+overlap_gona <- full_gona %>% tidyr::drop_na(Type.y)
+write.csv(overlap_gona, file = "data/overlap_gona.csv")
+
+# Fig3b correlation between linearPC1 and 3dPC1
+
+
+Fig3b <- ggstatsplot::ggscatterstats(
+  data  = full_gona,
+  x     = "3DGMPC1(10x10)",
+  y     = "Linear_PC1",
+  xlab  = "3DGM PC1",
+  ylab  = "Linear PC1",
+  results.subtitle = FALSE,
+)
+ggplot2::ggsave("Fig3b.3D PC1 correlation.png", path="figure.", width = 5, height = 5, dpi = 600)
+
+Fig3c <- ggstatsplot::ggscatterstats(
+  data  = full_gona,
+  x     = "SDI.y",
[…]

### 2023-11-30 manuscript/manuscript.Rmd
major updaes in manuscript and code.

@@
 abstract:
-|
-  ```{=latex}
-  TBD.
 
-  \textbf{Keywords:} Gona; TBD; TBD; TBD; TBD; TBD; TBD
-  ```
 date: "`r Sys.Date()`"
@@
 
-Early Acheulean sites in the Gona Project area are distributed over a wide area within the Dana Aoule North (DAN), Ounda Gona South (OGS), and Busidima North (BSN) drainages in the Busidima Formation [@quade2004] and range in age from approximately 1.7 to 1.2 mya [@semaw2018].
+Early Acheulean sites in the Gona Project area (**Figure** \@ref(fig:fig2)) are distributed over a wide area within the Dana Aoule North (DAN), Ounda Gona South (OGS), and Busidima North (BSN) drainages in the Busidima Formation [@quade2004] and range in age from approximately 1.7 to 1.2 mya [@semaw2018].
 The specific sites included in the current analysis are DAN-5, OGS-5, OGS-12, and BSN-17, all estimated to ca.
@@
 \>1.0 mya [@stoutStoneToolmakingEvolution2011; @presnyakova2018]) Acheulean assemblages, the Gona collections include numerous "crudely made" handaxes and picks on large flake blanks and cobbles, as well as large (\> 10cm) unmodified flakes, flaked pieces interpreted typo-technologically as Mode 1 cores (see Figure 1af), and smaller debitage [@semaw2018; @semaw2020].
+
+```{r fig2, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Gona site map."}
+knitr::include_graphics(here::here("figure/site map.png"))
+```
 
@@
 
-Following the criteria we proposed in the Introduction, Large Flake base cores from Acheulean contexts show evidence of design.
-Importantly, "design" in this context is understood as "design in four dimensions" sensu Kuhn [-@kuhn2020].
-Increasing FAI is associated with progressive narrowing ( i.e., decreasing PC1) and convergence (decreasing PC2) whereas cSDI is not.
-This indicates that the size and non-overlapping placement of flake removals drives shape change, rather than simply the removal of mass in general as might be expected from conventional models of resharpening (?).
-The fact that cSDI nevertheless correlates with FAI is consistent with relatively light (number of scars per piece ranges from 2 to 17 with a mean of 6.94) and spatially distributed re-touching involving relatively little overlap between scars.
-Such a pattern could be produced by intentional imposition of a design target or "mental template" or a more procedural/functional bias in favor of establishing cutting edges at one end of the long axis.
-In the absence of effective thinning/volume management techniques, this will lead to rapid narrowing/thickening of the piece overall and especially at the "tip." This is consistent with differences between flaked vs.
-unmodified large flakes and furthers the point that blank production already can be conceived as a process of "design" -- many unmodified flakes may have been similar in shape and just as useful as many retouched tools.
+Our analyses of flaked pieces from Oldowan and Early Acheulean contexts at Gona support the presence of two distinct reduction modes (1 and 2) that can be reliably discriminated using measures of artifact shape, flaking intensity, and extent of surface modification.
+However, they fail to support further sub-division between conventional Mode 2 tool types (handaxe, pick, knife).
+Following criteria proposed in the Introduction, we find strong evidence of imposed form in the retouched large flakes from Acheulean contexts at Gona, weak evidence of shaping in Mode 2 cobble cores from Acheulean contexts, and little to no evidence of shaping in Mode 1 cores from both Acheulean and Oldowan contexts.
+Our framing of this research question and its proposed test criteria follows the preponderance of archaeological literature in conceptualizing artifact design as a binary presence/absence of intentionally imposed form [@dibbleMajorFallaciesSurrounding2017].
+However, we will now argue that our results are better interpreted as reflecting variation in the degree and nature of design expressed by ESA toolmakers.
+This shift in perspective recognizes that the purposeful production of desired morphological and functional features can be achieved by different combinations of raw material selection, blank production, and flaking strategies, including the preservation of function through patterned changes in morphology [@kuhn2020].
 
-This contrasts with reduction of Mode 1 cores.
-Unexpectedly, FAI did have an effect to increase convergence.
-However, this effect was substantially weaker than seen in LFB Acheulean cores and may simply reflect the fact that reduction is unlikely to be perfectly symmetrical and our orienting protocol guarantees that one end is always more pointed than the other unless they are precisely equal in breath.
-In support of this "byproduct" interpretation, the relationship between FAI and cSDI is essentially what one would expect from knapping unconstrained by any systematic core-shaping strategy (greater flaked area can, but does not necessarily, accommodate a greater number of flake scars)
+## Large Flakes
+
+As we argued in the Introduction, the systematic production of large flakes indicates attention to artifact size as a desired design feature.
+Many of these flakes would have possessed substantial lengths of sharp edge and could have been immediately useful as cutting tools without further retouch.
+In fact, 33% of the large flakes in our total sample are unretouched.
+Notably, this is also true of the sub-sample from sites (OGS-5 and OGS-12) located in Type II, "tributary" sediments (29 of 82 unmodified, 35%).
+This likely reflects transport and discard of unmodified flakes from production sites closer to the axial river system sources of large clasts, consistent with the idea that such flakes were themselves treated as useful tools.
+We consider the alternative possibility, that unmodified flakes were unintended byproducts of local LCT production from transported boulder cores, to be less likely considering the energetic inefficiency of this strategy as well as the absence of large boulder cores in the assemblages (but see discussion of LCTs on cobble cores below).
+
+Whether or not unmodified flakes were themselves desired "end products", a majority of large flakes in the sample were retouched to some extent.
+Our analyses indicate that this flaking was performed in a manner that produced systematic and directional shape changes.
+Specifically, increasing FAI is associated with progressive reduction in relative breadth (increasing relative thickness and elongation; PC1) and increasing convergence (PC2).
+In contrast, variation in cSDI is not associated with artifact shape.
+These effects of surface modification extent but not reduction intensity indicate that it is the size and non-overlapping placement of flake removals that drives directional shape change, rather than the removal of mass in general.
+The observed correlation of cSDI with FAI further indicates that each new scar tends to remove additional unflaked area, consistent with the observed pattern of small numbers of relatively large and spatially distributed flake removals.
+This clearly differs from classic re-sharpening models, which emphasize shape change with decreasing artifact size as a byproduct of concentrated and repeated mass removal from particular areas of the artifact, such a working edges [@dibble1987; @mcpherron2003].
+It also clearly differs from the expected effects of un-patterned debitage on flake blanks, which does tend to increase relative thickness but also decreases elongation and is not associated with convergence [@moore2016].
+
+It is possible that the observed pattern reflects the intentional pursuit of an explicit morphological target or "mental template." This is generally seen as a strong claim implying rich mental representation and manipulation of spatial features [@wynn2002], cultural mechanisms for the intersubjective sharing of ideas of "appropriate" form [@wynn1995], and social reproduction of the skills required to reliably achieve these forms [@liuDifferentialEffectsKnapping2023].
+If there is such a target, however, it would seem to be a very loose one considering the observed range of shape variation in retouched flakes (Figure 5).
+A more cognitively and culturally "lean" interpretation is that the pattern reflects a procedural and/or functional bias toward establishing (or rejuvenating original) cutting edges at one end of the long axis.
+In the absence of effective thinning and volume management techniques, such a bias would naturally lead to narrowing/thickening of the piece overall and especially at the "tip," as is seen in our sample.
+This bias might itself be a socially learned behavioral convention or "habit" (sensu Isaac, 1986), but could plausibly emerge directly from individual responses to the morphology of large flake blanks.
+Such flakes are typically elongated and thus present: 1) two longer edges along the maximum dimension that are a natural focus for retouch, and 2) an ergonomic polarity defined by a thick (platform and percussion bulb) "butt" vs.
+a feather terminated "tip" [cf.
+@key2016; @wynn2018].
+Even with this minimal interpretation, however, we would argue that LFB flaking strategies at Gona present evidence of design in the sense of being intended to produce and/or maintain desired functional features of the LCT.
+This contrasts with patterns observed in cores classified as Mode 1 in our analyses which, as expected, appear to reflect simple debitage without intended morphological targets.
+
+## Mode 1 Cores
+
+A Discriminant Function Analysis strongly supported the presence of two distinct morphological types in our sample (based on overall shape and flake scar characteristics), which we had previously typo-technologically interpreted as reflecting Mode 1 (debitage) and 2 (façonnage) flaking strategies.
+Consistent with this interpretation, cores assigned to Mode 1 showed little or no evidence of core shape management.
+One potential exception is an unexpected tendency of increasing FAI to be associated with increasing convergence.
+However, this effect is an order of magnitude weaker than that seen in LFB Acheulean cores and may be an artifact of our orienting protocol, which aligns cores such that width at 25% length is always less than width at 75% length.
+If reduction tends to increase asymmetry from initially rounded cobble forms, this could account for the weak convergence effect we observe.
+In support of this interpretation, Mode 1 cores display a relatively strong correlation between FAI and cSDI driven by an increase in the upper bound of cSDI values as FAI increases.
+In other words, we find that greater flaked area on Mode 1 cores can, but does not necessarily, accommodate a greater number of flake scars.
+This is what would be expected from knapping unconstrained by any systematic core-shaping strategy.
+
+The presence (n=14) of relatively large (\> 100mm) Mode 1 cores at Gona Acheulean sites raises the further possibility that at least some of these might be depleted cores from LFB production.
+However, none of these display scars \> 90mm (mean = 51.8 mm), most are very lightly reduced (mean = 5.7 scars), and none meet the 150mm length cut-off used by Sharon (2009) to identify "large" cores suitable for LFB production.
+The largest Mode I core in our sample is 137 mm long, has five scars ranging from 25- 85 mm long, and an FAI of 25%.
+We thus consider it unlikely that large Mode I cores from Acheulean contexts were used for LFB production.
+The behavioral significance of relatively large, lightly-worked Mode 1 cores in Acheulean contexts is a question for future research.
+
+## Mode 2 Cores
 
 Our most complicated results come from Mode 2 cobble cores.
-Evidence of intentional shaping is limited: FAI has a moderate effect on PC1(flatness) and no effect on pointedness.
-Furthermore, the is no association between FAI and cSDI cobble cores display a greater range of cSDI values.
-Instead, FAI is strongly correlated with scar size (mean scar length: n=33, r=.588, P\<0.001; Figure 11).
-Together, this suggests that the removal of large flakes in particular increases FAI and tends to narrow the piece.
-We propose that this pattern may actually indicate that at least some cobble cores were precisely that: cores for the generation of LFBs.
-Given the difficulty of working relatively spherical cobbles, we expect that knappers would have chosen elongated and/or tabular cobbles.
-Attempts to maximize potential flake dimensions would naturally lead to removals along the long axis rather than at either end.
-This hypothesis was suggested by informal "experiential" knapping by the lead author with Gona cobbles, during which LFB production produced pick like forms as a byproduct.
-In support of the plausibility of this hypothesis, the maximum scars length on Gona Mode 2 cobble cores substantially overlaps with the size of unmodified large flakes from Acheulean contexts.
-It is also possible that cobble cores might begin as cores for LFB production and then be finished into core tools in their own right.
-These possibilities have major implications for understanding the occurrence and apparent morphological stability [@beyene2013] of "picks" throughout the African Acheuelan but remains to be tested.
+Evidence of intentional shaping is limited: FAI has a moderate effect on PC1 such that increasing flaked area produces relatively shorter, thicker, and narrower pieces.
+This violates expectations that intentional LCT shaping should increase elongation and convergence.
+However, it also fails to align with a reported tendency for un-patterned debitage to decrease the thickness and increase the elongation of cobble cores (Moore & Perston, 2016).
+This may reflect the fact that Mode 2 cobble cores at Gona begin and remain relatively elongated (Table 3: mean L/W = 1.68) throughout reduction compared to starting (n=29, mean L/W = 1.29) and ending (mean L/W = 1.34) cobble cores in the experiment of Moore and Perston [-@moore2016].
+The moderate trend toward thickening and shortening of cores with increasing FAI might plausibly be an effect of particularly elongated initial core form.
+In the absence of intentional thinning techniques [@stout2014] flaking around the perimeter of an oblong core would be expected to preferentially reduce length and especially breath (due to the long axis providing more potential platforms, especially if knappers seek to maximize flake size) while preserving thickness.
+Along these lines, Toth [-@toth1985] reported that "roller" shaped cobbles tended to produce bifacial choppers in his unstructured debitage experiments.
+
+Whether or not shaping occurred, it is clear that the elongated Mode 2 cobble cores at Gona do provide evidence of initial size and shape selection within the range of rounded cobbles available from Type 1 channels in the Busidima Formation [@quade2004].
+This pattern is not simply an analytical artifact of classifying elongated cores as "Mode 2" while selectively omitting rounded forms: the complete sample of large (\>=750g) cores in our sample (n=25) is dominated by relatively elongated forms (median L/W = 1.66).
+The small number (n=5) of large, rounded (L/W \< 1.33) cores in the sample are also anomalous in being very lightly worked compared to more elongated large cores (mean scar count = 4.2 vs.
+10.45, p \< 0.001; mean cSDI = 0.14 vs.
+0.31, p= 0.002; mean FAI = 36 vs.
+65, p = 0.016).
+
+This could reflect the intentional selection of large, elongated blanks for LCT production, as has been proposed at other early Acheulean sites [@duke2021; @harmand2009; @texier2018].
+We would characterize such selection as a design choice even in the absence of subsequent shaping.
+However, large-cobble shape preferences might equally reflect biases related to LFB production.
+Geometrically, elongated cobbles afford greater potential maximum dimensions for detached flakes than do rounder cobbles of a similar mass.
+Flaking along the long axis of elongated cobbles enables the production of elongated, side-struck flakes [@delatorre2018] that are typical of the LFB Acheulean and which provide longer and more ergonomic [cf.
+@wynn2018] cutting edges than do round flakes.
+Flatter and more elongated cobbles are also generally easier to open and exploit [@texier2018; @toth1982; @whittaker1994].
+This is an increasingly important consideration as size increases due to the increased percussive force needed to detach larger flakes and the challenges of supporting and positioning larger cores [@semaw2009], particularly with cobbles that are large but not sufficiently massive to act as stationary targets for thrown hammerstones [@toth2019].
+We thus find it plausible that at least some typological picks, handaxes, and knives made on cobble cores at Gona could have been by-products of large flake production rather than intended forms.
+Toth and Schick [-@toth2019: 741] similarly note that "heavily reduced boulder cores can assume the form of smallish discoids and polyhedrons [and] may not be identified as sources of large flake blanks by many archaeologists".
+If this is true, such cores are distinguished from classic Mode 1 cores by allometric effects on raw material choice and flake placement rather than a shift to deliberate shaping.
+
+The possibility that some Mode 2 cobble cores at Gona might be the remains of cores used for large flake production was initially suggested by informal, "experiential" knapping with Gona cobbles, during which attempts at LFB production produced pick-like forms as a byproduct.
+Consistent with this hypothesis, the mean (152.9mm) and median (156mm) length of Mode 2 cobble cores are both above the 150mm cut-off proposed by Sharon [-@sharon2009] and the maximum scar length on each Mode 2 cobble core substantially overlaps with the size range of unmodified large flakes (Table 3).
+Importantly, this is also true of the largest retouched flakes at Gona, nine of which exhibit flake scars \>100mm and thus could have been used as LFB cores.
+This possibility has also been suggested at Koobi Fora [@presnyakova2018] and might be relevant to understanding very large LCTs on flakes at other sites [e.g., @beyene2013].
+
+Nevertheless, the fact remains that the preponderance of scars on large cobble cores are from the removal of smaller flakes.
+Although these smaller scars could record preparatory flaking for LFB predetermination [@delatorre2018] and/or a subsequent stage of small flake debitage [@shea2010], it is difficult to rule out a role in shaping.
+Indeed, some pieces (e.g., Figure 1 ac, ad, ae) exhibit delicate points and sharp edges that are strongly suggestive of intentional shaping.
+If these pieces did in fact begin as cores for LFB production, it is possible that they were subsequently shaped into cutting tools in their own right.
+Such lithic "upcycling" of depleted LFB cores would help to explain the transport of these pieces away from the axial river system.
+However, we should also be cautious not to overinterpret a small number of suggestive pieces pulled from a wider range of variation, as they may simply represent low frequency "spandrels" of un-structured debitage [@moore2016; @moore2020].
+These various possibilities remain to be tested, but we note that there is no a priori reason to think that Acheulean tool makers would have neglected potentially useful cores because they "belonged" to a particular reduction sequence or technological type.
+
+## Implications
+
+Early Acheulean technology is differentiated from the preceding Oldowan by changes in artifact size as well as morphology [@isaac1977].
+However, discussion of the technological, cognitive, and cultural implications of this transition has often focused on the emergence (or not) of intentional shape imposition [e.g., @duke2021; @holloway1969; @gowlett1986; @isaac1976; @roche2005; @wynn1995; @wynn2002].
+Our results suggest that this emphasis may be misplaced, and that distinctive Early Acheulean artifact forms might represent secondary accommodations to the primary goal of increasing tool size.
+From this perspective, Early Acheulean shape imposition is best understood as reflecting an interaction between functional design goals and technological constraints [@kuhn2020; @wynn2018].
+
+The consensus view is that Oldowan flaking goals focused on the production of sharp, cutting flakes through least effort debitage [@toth1985].
+Somewhat more controversially, Oldowan flaking may have included preferred debitage patterns [@stout2019] and/or intentional core maintenance and rejuvenation strategies [@delagnes2005].
+Our failure to find systematic effects of reduction on Oldowan core shape at Gona is consistent with this broad characterization.
+Starting from this Oldowan baseline, the production of larger cutting tools could in principle be achieved by increasing the size of cores and detached flakes and/or by attempting to produce and maintain cutting edges on cores themselves rather than on the smaller pieces detached from them.
+Both strategies are evident in the Early Acheulean at Gona.
+
+At Gona, large flakes appear to have been transported across the landscape and discarded either in their original form or after relatively light modification to impose desired cutting edges.
+This behavior combines the two size-maximizing strategies identified above by increasing flake size and then using these large detached pieces as supports for retouched edges.
+On the other hand, large cobble cores at Gona display shape variation more consistent with debitage than shaping.
+This leads us to suggest that these pieces might be depleted cores from large flake production (size-maximizing strategy 1) that have been "upcycled" as large core tools (size-maximizing strategy 2).
+In this way, the full range of artifact types and patterns of shape variation in the Gona Early Acheulean can be parsimoniously accounted for as the expression of two strategies for increasing cutting tool size.
+
+This interpretation has three main implications.
+First, it supports the view that the Acheulean initially emerged as a set of technological strategies for increasing cutting tool size, especially by striking large flakes [@isaac1969].
+Hypothetically, this size increase might have been favored by novel functional priorities, such as extended use-life, enhanced transportability, and utility for heavy-duty cutting [@shea2010; @toth2019].
+It may also have been motivated and/or enabled by increases in hominid body size.
+These overlapping possibilities appear likely to be mutually reinforcing rather than mutually exclusive.
+In fact, the appearance of the Acheulean starting 1.95 mya [@mussi2023] is roughly contemporaneous with an increase in the number, size, and ecogeographic range of stone tool sites in general [@plummer2004], as well as evidence of habitat diversification and greater lithic transport distances for Acheulean sites specifically [@hay1976; @quade2008; @rogers1994; @delatorre2008], increasing emphasis on large animal butchery [@linaresmatás2021], and increases in hominid brain and body size [@antón2014] including the first appearance of *Homo erectus* [@mussi2023].
+Increased cutting tool size is a plausible response and/or enabling factor to many of these shifts and is in that sense appears "overdetermined" by available evidence.
+Importantly, the current study does not resolve the extent to which the production and transport of Early Acheulean tools around the landscape [@presnyakova2018] occurred through planned logistical [@binford1980] activities of individuals and/or collaborating groups vs.
+the cumulative effects of independent transport and recycling events over time [@reeves2023].
+Implications for planning and prospection capacities [@szpunar2014] thus remain unclear.
+
+Second, it reinforces a growing consensus [e.g., @duke2021; @kuhn2020; @presnyakova2018] that early Acheulean LCT types artificially partition a continuum of variation rather than representing distinct target forms.
+More specifically, we propose that these forms emerge as the expression of a generic goal of cutting tool size maximization implemented across varying material constraints and opportunities.
+This is particularly relevant to the interpretation of "picks," which have long been viewed as a morphologically distinct but functionally mysterious tool type.
+Toth and Schick [-@toth2019] note that picks do not appear designed to provide good cutting edges but also do not show use-wear consistent with digging.
+This leads them to speculate that picks may have been specialized weapons used to dispatch large, wounded animals with a blow to the head.
+However, picks are a common, persistent, and morphologically variable artifact type in the Early Acheulean, which argues against such a narrow function.
+We suggest that typological picks may lump together a variable mix of depleted LFB cores and large cutting tools made on relatively thick (cobble or flake) blanks, possibly including upcycled LFB cores.
+Within this interpretation, picks would be part of a continuum of morphological variation produced by different raw material forms and complex reduction histories, rather than a distinct tool type designed for novel function.
+Due to such constraints, picks as a class may appear less well designed for cutting than LCTs produced on thinner blanks.
+However, the relationship between tool morphology and cutting efficiency is complex and multivariate [@keyIntegratingMechanicalErgonomic2016], and the relative utility of short and/or higher-angle edges on a more massive tool is not well understood across diverse cutting tasks.
+Picks as an artifact class have not received the same attention as handaxes and cleavers and we are unaware of any broad comparative synthesis.
+Beyene et al.
+[-@beyene2013] do report that picks at Konso are relatively conservative in shape and scar counts over a period from 1.75-1.2 Ma during which associated handaxes show clear increases in refinement.
+They thus suggest that pick function may already have been effectively optimized by 1.75-1.6 Ma whereas the cutting function of handaxes continued to be enhanced.
+This is broadly consistent with the current suggestion that typological picks identify a morphological extreme produced by raw material features (esp.
+thickness) that constrain or discourage the refinement that might be developed on other blanks.
+From 1.75-1.2 Ma, these ad hoc and/or less refined cutting tools appear to have remained a stable element in the tool kit.
+However, they are nearly absent at 0.85 Ma, coincident with the appearance of "considerably refined" handaxes.
+Indeed, picks are not commonly reported to co-occur with refined handaxes in later Acheulean contexts anywhere.
+On a traditional functional interpretation, this would imply that their originally designed function also became less important in these contexts.
+The alternative possibility suggested here is that increased investment in handaxe refinement decreased the perceived value of producing and using ad hoc core tools for similar functions.
+
+Lastly, our interpretation suggests that LCT variation at Gona probably does not reflect intentional imposition of morphological norms [cf. @holloway1969] or detailed mental templates [@deetz1967].
+Instead, we would focus attention on the cognitive and cultural implications of size maximization strategies.
+As reviewed by Stout [-@stoutStoneToolmakingEvolution2011], these stem from 1) the physical and strategic challenges of quarrying large flakes and 2) the increased complexity of knapping action hierarchies resulting from the addition of novel sub-goals (e.g.
+create a cutting edge) involved in shaping large flake and cobble cores into useful tools.
+Systematic large flake production strategies and techniques are qualitatively different from small-flake debitage [@isaac1969; @toth2019] and thus add to the volume of technical knowledge and know-how that must be acquired by individuals.
+This is directly relevant for inferring learning demands and the possible role of social support in Acheulean skill reproduction [@pargeter2019].
+Stout [-@stout2002] provides a modern example of large flake production that illustrates the potential scope of such demands.
+Interestingly, much earlier large flake production has been reported at the 3.3 Ma site of Lomekwi 3 but is argued to be poorly controlled with minimal core reduction, numerous steps and hinges, and extensive platform battering.
+Although concerns have been raised regarding the dating and context of this site [@archer2020; @dominguez-rodrigo2019], it may provide an interesting technological comparison to Early Acheulean large flake production.
+
+With respect to knapping action organization, the intentional placement of flake removals to generate desired core morphologies demonstrates a more complex goal structure than is required for simple, least-effort debitage.
+Such complexity has implications for assessing cognitive demands including relational integration, temporal abstraction, and goal abstraction [@stoutStoneToolmakingEvolution2011].
+For this reason, our study was designed to address concerns that intentional shaping might not actually be characteristic of Early Acheulean technology [e.g., @moore2016].
+We did find evidence for (at least) the imposition of cutting edges on LFBs, and offer this a foundation for "minimum necessary competence" [sensu @killin2021] cognitive interpretations.
+We remain agnostic as to whether similar knapping complexity was demonstrated in the preceding Oldowan, but note that the rare and un-standardized [@gallotti2018] occurrence of debitage on flakes prior to 2.0 mya does not provide evidence of intentional shaping.
+Stout [-@stoutStoneToolmakingEvolution2011] argued that elaborated small flake debitage methods contemporary with the Early Acheulean, such as single platform Karari scrapers [@isaac1997] and hierarchical centripetal cores [@delatorre2003], document complex goal structures including the intentional modification of the core morphology to enable subsequent flake detachments.
+Others have seen evidence of such intentions in earlier bifacial [@duke2021] and unifacial [@delagnes2005] debitage strategies.
+However, all of these interpretations are based on qualitative assessments of flake scar pattering and/or refitting and remain open to critique from quantitative and experimentally-based approaches contending that similar knapping patterns can emerge without "top down" intentions [@moore2016; @moore2020; @toth1985].
+
+However these debates are eventually resolved, we would stress that the earliest demonstration of a particular capacity is unlikely to represent its earliest presence.
+Logically, the minimum capacities required to support a novel behavior must be present before the behavioral innovation can occur [@stout2023] and may predate it substantially.
+We would thus argue that technological innovation in the Early Acheulean was more likely stimulated by changing behavioral and ecological strategies that placed a premium on cutting tool size, rather than by the sudden emergence of new cognitive capacities.
+In this context, it is important to note that the preponderance of cognitive and especially neuroarchaeological investigations of handaxe manufacture [@hecht2023; @stout2015; @stoutStoneToolmakingEvolution2011; @stout2008] have studied refined, later Acheulean forms and should not be generalized to discussions regarding the emergence of the Early Acheulean [cf.
+@duke2021].
+These neuroarchaeological studies were made possible by a robust understanding of the specific knapping behaviors being modeled [@stout2023].
+Similar studies modeling Early Acheulean technology are clearly needed and we hope that investigations like the current one can help to provide the necessary behavioral foundation.
+
+## Generalizability
+
+It remains to be seen to what extent current results can be generalized to other Early Acheulean sites.
+It is clear that LFB production as seen at Gona is typical [@isaac1969] of the African Acheulean from its first known appearance [@mussi2023].
+Similarly, the transport and discard of unmodified large flakes has also been reported at Peninj [@delatorre2008] and Koobi Fora [@presnyakova2018] and may be more widespread than has been specifically noted in publication.
+The general pattern of light, non-invasive retouch we observe also appears typical of Early Acheulean sites [reviewed by @presnyakova2018].
+However, the effects of reduction on LCT shape have not been systematically tested across sites in ways comparable to the current analyses.
+This makes it difficult to determine if rare examples of "well shaped" Early Acheulean LCTs [e.g., @diez-martín2019] represent distinct technological behaviors or extreme points along a continuum of variation.
+Presnyakova et al. [-@presnyakova2018] argue for the latter, supported by shape effects of reduction intensity identified using artifact size and edge angle as proxies.
+However, these authors do not consider SDI or FAI and their analyses excluded any flaked pieces without "clearly defined tips and bases," thus limiting comparability with current results.
+
+The occurrence of numerous large cobble cores (typological LCTs on cobble bases) at Gona is atypical for East Africa and may reflect local raw materials availability.
+Kuman [-@kuman2007] reports that river cobbles from the nearby Blaaubank gravels are the dominant blank form in Gauteng "Cradle of Humankind" Early Acheulean collections, possibly due to the rarity of larger boulder-size clasts.
+Relatively high frequencies of LCTs on cobbles have also been reported from Gadeb in Ethiopia [@delatorreEarlyStoneAge2011].
+However, the large cobble cores at these sites have not been systematically studied as a distinct technological component or potential source of LFBs.
+Substantial use of large slabs/blocks for LCT production has been reported at Kokiselei 4 [@harmand2009] and Olduvai Gorge [@delatorre2005] but again it is not clear whether these could also have been sources for LFBs during some part of their reduction sequence.
+Interestingly, Presnyakova et al.
+[-@presnyakova2018: Fig.
+8] illustrate a boulder core with refitting large flake from FxJj65 that appears to be comparable in size (\~150mm long), shape (L/W $\approx$ 1.5), and reduction intensity to some of the more lightly worked cobble cores at Gona, providing at least anecdotal support for our interpretations.
+In sum, however, systematic comparison across published Early Acheulean sites is not currently possible due to differences in research questions, methods, and data reporting practices.
+Clearly much work is needed to enable such comparison, but we do find some encouragement in the current finding that conventional linear measures, simple scar counts, and visual flaked area estimations are sufficient to describe relevant variation in shape and reduction
+
+# Conclusion
+
+Characteristic Early Acheulean artifact forms including large retouched flakes and core tools began to appear ca.
+2.0 Ma [@mussi2023] as part of a pervasive ecological and behavioral shift encompassing changing diet, ranging patterns, and habitat usage; increasing site frequency, density, and ecogeographic distribution; and the first appearance of *Homo erectus*.
+Although the cognitive and cultural significance of this new lithic industry is often framed in terms of the emergence of discrete tool types and/or intentionally imposed morphological norms, we find little evidence at Gona to support this.
+Instead, we observe systematic patterns of raw material selection and core surface modification organized around the production and maintenance of useful cutting edges on large(r) supports.
+Such larger cutting tools may have been prioritized in newly emerging hominid behavioral adaptations due to their extended use-life, enhanced transportability, and/or suitability for heavy-duty cutting.
+
+This interpretation is consistent with a characterization of Early Acheulean tool forms as emerging from a set of "design imperatives" for large hand-held cutting tools [@gowlett2006; @wynn2018], specifically including elongation and ergonomic positioning of cutting edges [[\@key2016; \@keyIntegratingMechanicalErgonomic2016](Key%20et%20al.,%202016;%20Key,%202016)].
+Our results thus provide support for the presence of imposed form in the broad theoretical sense developed by Kuhn [-@kuhn2020]: not as the realization of a "holographic model" in the mind of the maker, but rather as the outcome of technological choices made in view of material constraints and affordances [@moore2020].
+To the extent that some, but not all [@muller2022], later Acheulean assemblages provide stronger evidence of normative form imposition [@garcía-medrano2019; @liuDifferentialEffectsKnapping2023; @shipton2020] this underlines the heterogeneity of lithic technologies subsumed within the "Acheulean Industry" and argues for more context-specific approaches to evaluating behavioral, cognitive, and cultural implications.
 

### 2023-11-30 code/Analysis.R

@@ -6,45 +6,57 @@ library(PCAmixdata)
 library(psych)
 library(patchwork)
 library(figpatch)
-# Add data
-gona_esa<-read.csv("data/EA_Reduction_Paper_Order.csv")
+# Data preparation
+## loading data
+gona_linear<-readr::read_csv("data/Gona_ESA_Linear.csv")
+gona_3d<-readr::read_csv("data/Gona_ESA_3D.csv")
+## creating a new column of unique identifier for merging
+gona_linear$Name <- paste(gona_linear$Site, gona_linear$Catalog, sep= "-")
+## merging two datasets based on Name
+joined_gona <-dplyr::left_join(gona_linear, gona_3d, by = c("Name" = "Name"))
+## Reordering column order for principle component analysis
+joined_gona <- joined_gona %>% dplyr::relocate(GM_L,GM_Br1,GM_Br2,GM_Br3,GM_T1,GM_T2,GM_T3)
+## Subsetting a dataframe that contains both linear and 3d data for measuremnt validation
+overlap_gona <- joined_gona %>% tidyr::drop_na(Flaked.y)
 
-# Principle component analysis
-gona_esa_pca <-prcomp(gona_esa[1:7],  scale = TRUE)
-pcaplot<-fviz_pca_ind(gona_esa_pca, habillage= gona_esa$Type, # color by groups
-                      addEllipses = TRUE, # Concentration ellipses
-                      ellipse.type = "convex",
-                      legend.title = "Type",
-                      label="none")
-pcaplot
+### #######################Trial
+gona_linear1<-readr::read_csv("data/EA_Reduction_Paper_41.csv")
+gona_linear1$Name <- paste(gona_linear1$Site, gona_linear1$Catalog, sep= "-")
+joined_gona1 <-dplyr::left_join(gona_linear1, gona_3d, by = c("Name" = "Name"))
+joined_gona1 <- joined_gona1 %>% dplyr::relocate(GM_L,GM_Br1,GM_Br2,GM_Br3,GM_T1,GM_T2,GM_T3)
+## Subsetting a dataframe that contains both linear and 3d data for measuremnt validation
+overlap_gona1 <- joined_gona1 %>% tidyr::drop_na(Flaked)
+
+gona_esa_pca12 <-prcomp(overlap_gona1[1:7],  scale = FALSE)
+res.var <- get_pca_var(gona_esa_pca12)
[…]

### 2023-12-01 manuscript/manuscript.Rmd
text update and figure reproduction.

@@
 A lack of shape correlation with either measure would be expected for simple debitage with no morphological targets whereas a strong correlation with both would indicate a highly "designed" form achieved through extensive morphological and volumetric transformation.
+(**Figure** \@ref(fig:FAISDI))
 In the current study we thus considered SDI and FAI together in order to evaluate evidence of intentional shaping in the early Acheulean of Gona.
+
+```{r FAISDI, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A conceptual model of inferring intention from artifact morphology."}
+knitr::include_graphics(here::here("figure/FAI_SDI.png"))
+```
 
@@
 We address this by collecting linear measures defined by homologous semi-landmarks.
-All artifacts were oriented along their maximum dimension, which was measured and defined as "length." The next largest dimension orthogonal to length was used to define the plane of "breadth," with the dimension orthogonal to this plane defined as "thickness" Width (W~1~, W~2~, W~3~) and thickness (T~1~, T~2~, T~3~) measures were then collected at 25%, 50%, and 75% of length, oriented so that 25% Width \> 75% Width.
+All artifacts were oriented along their maximum dimension, which was measured and defined as "length".
+The next largest dimension orthogonal to length was used to define the plane of "breadth", with the dimension orthogonal to this plane defined as "thickness" Width (W~1~, W~2~, W~3~) and thickness (T~1~, T~2~, T~3~) measures were then collected at 25%, 50%, and 75% of length, oriented so that 25% Width \> 75% Width.
 To partition variation in shape from variation in size, we divided all linear measures by the geometric mean [@lycett2006].
@@
 
-### 3D Methods
-
-Artifact scans (N=33) were cleaned, smoothed, and re-meshed using MeshLab.
-The 3D triangular mesh of each scan was computed using Kazhdan and Hoppe's [-@kazhdan2013] method of screened Poisson surface reconstruction in MeshLab.
-3DGM analysis was conducted using the *AGMT3-D* program of Herzlinger and Grosman [-@herzlinger2018].
-Artifacts were automatically oriented according to the axis of least asymmetry, then manually oriented in the interests of standardization with the length, width, and thickness dimensions defined by the longest axis, followed by the next two longest axes perpendicular to the first.
-The wider end of the artifacts was positioned as the proximal end (base), and more protruding surfaces were oriented towards the user in the first orthogonal view.
-Then, a grid of 200 homologous semi-landmarks were overlain on each artifact's surface.
-Generalized Procrustes and Principal Component analyses were then undertaken to explore the shape variability of the sample.
-The surface area of each artifact was calculated using the *Artifact3-D* program of Grosman et al. [-@grosman2022].
-*Artifact3-D* was also used to automatically identify the flake scar boundaries and compute each scar's surface area, using the scar analysis functions of Richardson et al. [-@richardson2014].
-
-### Analyses
-
-Measurement Validation: To assess the adequacy of shape descriptions based on our linear measures, we directly compared these with shape as quantified by 3D methods on the 33 artifacts for which scans are available.
+To assess the adequacy of shape descriptions based on our linear measures, we directly compared these with shape as quantified by 3D methods on the 33 artifacts for which scans are available.
 GM-transformed linear measures from these 33 artifacts were submitted to a variance-covariance matrix PCA.
@@
 Accuracy of surface area and flaked area estimates was also assessed by correlation with 3D results.
+
+### 3D Methods
+
+3DGM analysis was conducted using the *AGMT3-D* program of Herzlinger and Grosman [-@herzlinger2018].
+Artifacts were automatically oriented according to the axis of least asymmetry, then manually oriented following the same orthogonal conventions described in section 2.3.
+Then, a grid of 200 homologous semi-landmarks were overlain on each artifact's surface.
+Generalized Procrustes and Principal Component analyses were then undertaken to explore the shape variability of the sample.
+The surface area of each artifact was calculated using the *Artifact3-D* program of Grosman et al.
+[-@grosman2022].
+*Artifact3-D* was also used to automatically identify the flake scar boundaries and compute each scar's surface area, using the scar analysis functions of Richardson et al.
+[-@richardson2014].
 
@@
 
-A PCA on GM transformed linear measures the 33 artifacts identified two PCs accounting for 75.6% of the variance.
+A PCA on GM transformed linear measures of the 33 artifacts identified two PCs accounting for 75.6% of the variance.
 Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance.
@@
 This closely parallels the length vs.
-width and thickness tradeoff captured by 3DGM-PC1 (Figure 3a) and is reflected in a tight correlation of artifact scores produced by the two PCs (r = 0.903, p \< 0.001, Figure 3c).
-Comparison with conventional shape ratios (Table 2) similarly indicates that both Linear-PC1 and 3DGM-PC1 largely capture variation in artifact elongation.
+width and thickness tradeoff captured by 3DGM-PC1 (**Figure** \@ref(fig:fig3d)) and is reflected in a tight correlation of artifact scores produced by the two PCs (r = 0.903, p \< 0.001, **Figure** \@ref(fig:figcomparison)A).
 A second factor (Linear-PC2, eigenvalue = 0.07) explained an additional 20.4% of variance.
 This factor was less strongly correlated with its 3DGM counterpart (3DGM-PC2; r = 0.344, p = 0.050) probably because Linear-PC2 describes anticorrelated variation in width and thickness (i.e., broad and flat vs.
-thick and pointed; Table 1) whereas 3DGM-PC2 more purely isolates convergence (Figure 3a).
-The remainder of the shape variability explained by Linear-PC2 is captured by higher order 3DGM-PCs like 3DGM-PC3 to 5, which comprise the contribution of the left and right lateral margins to relative thickness.
+thick and pointed; **Table** \@ref(tab:tab1)) whereas 3DGM-PC2 more purely isolates convergence (**Figure** \@ref(fig:fig3d)).
+The remainder of the shape variability explained by Linear-PC2 is captured by higher order 3DGM-PCs 3 through 5, which comprise the contribution of the left and right lateral margins to relative thickness.
 Use of high-resolution, coordinate-based scan data thus generates PCs that identify more specific shape attributes, but the underlying morphological variability captured by the linear and 3D analyses remains similar.
-Together, 3DGM-PC2 (r = 0.344, p = 0.050), 3DGM-PC3 (r = -0.416, p = 0.016), 3DGM-PC4 (r = 0.458, p = 0.007), and 3DGM-PC5 (r = -0.352, p = 0.044) correlate well with Linear-PC2 (Figure 3c), cumulatively capturing whether the items are broad and flat or thick and pointed.
-A stepwise regression (r2=0.625, F(4,28)=11.697, p\<0.001, Probability-of-F-to-enter \<= 0.050; Probability-of-F-to-remove \>= 0.100) with Linear-PC2 as the dependent variable retained all four of these 3DGM-PCs as significant predictors.
+Together, 3DGM-PC2 (r = 0.344, p = 0.050), 3DGM-PC3 (r = -0.416, p = 0.016), 3DGM-PC4 (r = 0.458, p = 0.007), and 3DGM-PC5 (r = -0.352, p = 0.044) correlate well with Linear-PC2, cumulatively capturing whether the items are broad and flat or thick and pointed.
+A stepwise regression ($r^2$=0.625, F(4,28)=11.697, p<0.001, Probability-of-F-to-enter <= 0.050; Probability-of-F-to-remove >= 0.100) with Linear-PC2 as the dependent variable retained all four of these 3DGM-PCs as significant predictors.
 
@@
                  "Linear-PC2" = c("-0.107", "0.350", "0.767", "0.790", "-0.679", "-0.623", "-0.282") ) 
-kableExtra::kbl(df1, booktabs = T, caption = "Component loadings for linear metric PCs .") 
+kableExtra::kbl(df1, booktabs = T, caption = "Component loadings for linear metric PCs on scanned sample.") 
+```
+
+```{r fig3d, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="3D models displaying the first two PCs."}
+knitr::include_graphics(here::here("figure/Fig3a.png"))
+```
+
+```{r figcomparison, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Comparison between linear measurement and 3d measurement."}
+knitr::include_graphics(here::here("figure/Fig.3D comparison.png"))
 ```
@@
 PC1 (Eigenvalue = 0.216) explained 56.4% of the variance.
-Factor loadings (Table 2) for PC1 reflect artifact flatness (i.e., an anti-correlation of length and width vs. thickness) such that higher values indicate relatively thinner pieces.
+Factor loadings (**Table** \@ref(tab:tab2)) for PC1 reflect artifact flatness (i.e., an anti-correlation of length and width vs. thickness) such that higher values indicate relatively thinner pieces.
 PC2 (Eigenvalue = 0.090) explained 23.6% of the variance.
-Factor loadings (Table 2) show that PC2 captures artifact convergence or "pointedness" (i.e., an anti-correlation of tip width with length and butt thickness) such that higher values indicate shorter, less pointed forms.
+Factor loadings (**Table** \@ref(tab:tab2)) show that PC2 captures artifact convergence or "pointedness" (i.e., an anti-correlation of tip width with length and butt thickness) such that higher values indicate shorter, less pointed forms.
+
+```{r tab2, echo=FALSE, message=FALSE}
+library(tidyverse)
+df1 <- data.frame("Linear metrics (GM transformed)" = c("Length", "W1", "W2", "W3", "T1", "T2", "T3"), 
+                 "PC1" = c("0.905", "0.635", "0.680", "0.486", "-0.510", "-0.684", "-0.719"), 
+                 "PC2" = c("-0.414", "0.299", "0.571", "0.761", "-0.526", "-0.525", "0.023") ) 
+kableExtra::kbl(df1, booktabs = T, caption = "Component loadings for linear metric PCs in the full sample.") 
+```
 
 We also tested the validity of our two reduction measures, SDI and FAI.
-In agreement with Clarkson (2013), we found that surface area estimated from caliper measures displayed a strong correlation with (r2=0.975, p \< 0.001) but linear over-estimation of ($\beta$ = 1.58) 3DGM surface area (Figure 4a).
-This results in a systematic underestimation of SDI that scales with core size (Figure 4b).
-However, a simple correction of the caliper estimate (dividing by the slope, 1.58) eliminates surface area over-estimation (mean difference = 256mm2 [\<1.7% of mean], p=0.040) and produces SDI values that agree with 3DGM (r2=0.975, p \< 0.001, $\beta$ = 0.98) (Figure 4c).
+In agreement with Clarkson (2013), we found that surface area estimated from caliper measures displayed a strong correlation with (r2=0.975, p \< 0.001) but linear over-estimation of ($\beta$ = 1.58) 3D surface area.
+This results in a systematic underestimation of SDI that scales with core size.
+However, a simple correction of the caliper estimate (dividing by the slope, 1.58) eliminates surface area over-estimation (mean difference = $256mm^2$ [\<1.7% of mean], p=0.040) and produces SDI values that agree with 3D values (r2=0.975, p \< 0.001, $\beta$ = 0.98) (**Figure** \@ref(fig:figcomparison)B).
 We thus proceeded to apply this correction to surface area estimates in the full sample.
 Insofar as these relationships are driven by basic geometry, we expect these methods (including correction) to be generalizable to other ESA assemblages.
+
+Visual estimation of flaked area approximated 3DGM measurement very closely (r2=0.932, p < 0.001, $\beta$ = 1.051; **Figure** \@ref(fig:figcomparison)C) and without any systematic bias (paired t-test: mean difference = -0.015%, 95% CI = 2.02% to 1.99%, p = 0.987).
+Individual errors ranged between -10.46% and 14.24%.
+We thus considered visual estimation to be reliable in our sample.
 
@@
 This piece, typologically classified as a "knife", is the smallest (93mm) retouched flake in the Acheulean sample.
-When retouched flakes are excluded from the comparison, there are no significant differences in shape between Mode 1 cores from Oldowan and Achuelean contexts.
+When retouched flakes are excluded from the comparison, there are no significant differences in shape between Mode 1 cores from Oldowan and Acheulean contexts.
 Interestingly, however, Acheulean Mode 1 cobble-cores are much larger (mean weight 480g vs.
@@
 In Mode 2 cores (n=115), reduction measures have different effects depending on initial blank form.
-In general, the contrast between flat, lightly-worked flake bases and rounder, more-heavily worked cobble bases tends to inflate relationships between FAI and core shape (Figure 7c,d).
+In general, the contrast between flat, lightly-worked flake bases and rounder, more-heavily worked cobble bases tends to inflate relationships between FAI and core shape (**Figure** \@ref(fig:FAIBASE)).
 Mode 2 cobble cores (n=37) display a negative effect of FAI on PC1(Flatness) ($\beta$=-0.011, r=0.367, p = 0.026), no effect of FAI on pointedness, and no association between FAI and cSDI (r=0.141, p = 0.406).
 Flake cores (n=69) show negative effects of FAI on PC1(Flatness) ($\beta$=-0.007, r=0.284, p = 0.018) and, more strongly, PC2(pointedness) ($\beta$=-0.019, r=0.443, p \< 0.001).
+Comparison of Mode 2 flake cores with unmodified large flakes from Acheulean contexts (n=35) shows that ranges of shape variation substantially overlap but that flaking generally reduces both PC1 and PC2.
+In fact, regressions of PC1 and PC2 on FAI show y-intercepts that closely approximate unmodified flake mean values (0.852 vs.
+0.774 and 0.693 vs.
+0.594, respectively) and significantly negative slopes as FAI increases (**Figure** \@ref(fig:FAIFLAKED)).
 Flake cores show a small but significant positive effect of FAI on cSDI that, in contrast to Mode 1 cores, represents actual covariance rather than simply a relaxation of constraint (i.e.
@@
 
-Comparison of Mode 2 flake cores with unmodified large flakes from Achuelean contexts (n=35) shows that ranges of shape variation substantially overlap but that flaking generally reduces both PC1(flatness) and PC2(pointedness).
-In fact, regressions of PC1 and PC2 on FAI show y-intercepts that closely approximate unmodified flake mean values (0.852 vs.
-0.774 and 0.693 vs.
-0.594, respectively) and significantly negative slopes as FAI increases (Figure 9).
+```{r FAIBASE, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Mode 2 trends across blank types."}
+knitr::include_graphics(here::here("figure/Fig.FAI by base.png"))
+```
+
+```{r FAIFLAKED, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Effect of FAI compared to unmodified large flakes from Acheulean contexts."}
+knitr::include_graphics(here::here("figure/Fig.FAI by flaked.png"))
+```
 
 A small number of Mode 2 cores (n=9) were so heavily modified (mean FAI=89%, range 65-100) that the blank form could not be determined.
-These heavily worked "indeterminate" cores tend to show similar PC1(flatness) and PC2(pointedness) values to cobble-base cores (i.e. thicker and more pointed than flake cores; Figure x).
-They also appear to follow general trends of shape change with increasing FAI observed across base types (Figure 10).
+These heavily worked "indeterminate" cores tend to show similar PC1 (flatness) and PC2 (pointedness) values to cobble-base cores (i.e.
+thicker and more pointed than flake cores) and follow general trends of shape change with increasing FAI observed across base types (**Figure** \@ref(fig:FAIBASE)).
 

### 2023-12-01 code/Analysis.R

@@ -1,11 +1,8 @@
 # Add packages
 library(tidyverse)
 library(factoextra)
-library(FactoMineR)
-library(PCAmixdata)
-library(psych)
 library(patchwork)
-library(figpatch)
+
 # Data preparation
 ## loading data
 gona_linear<-readr::read_csv("data/Gona_ESA_Linear.csv")
@@ -19,24 +16,91 @@ joined_gona <- joined_gona %>% dplyr::relocate(GM_L,GM_Br1,GM_Br2,GM_Br3,GM_T1,G
 ## Subsetting a dataframe that contains both linear and 3d data for measuremnt validation
 overlap_gona <- joined_gona %>% tidyr::drop_na(Flaked.y)
 
-### #######################Trial
-gona_linear1<-readr::read_csv("data/EA_Reduction_Paper_41.csv")
-gona_linear1$Name <- paste(gona_linear1$Site, gona_linear1$Catalog, sep= "-")
-joined_gona1 <-dplyr::left_join(gona_linear1, gona_3d, by = c("Name" = "Name"))
-joined_gona1 <- joined_gona1 %>% dplyr::relocate(GM_L,GM_Br1,GM_Br2,GM_Br3,GM_T1,GM_T2,GM_T3)
-## Subsetting a dataframe that contains both linear and 3d data for measuremnt validation
-overlap_gona1 <- joined_gona1 %>% tidyr::drop_na(Flaked)
-
-gona_esa_pca12 <-prcomp(overlap_gona1[1:7],  scale = FALSE)
-res.var <- get_pca_var(gona_esa_pca12)
-variable_loadings123<-data.frame(res.var$coord) 
-fviz_eig(gona_esa_pca12, addlabels=TRUE, hjust = -0.1)
-
-##################################################
 
 # Measurement Validation
 ## Principle component analysis
-gona_esa_pca <-prcomp(overlap_gona[1:7],  scale = TRUE)
+gona_pca_overlap <-prcomp(overlap_gona[1:7], scale = FALSE)
+overlap.res.var <- get_pca_var(gona_pca_overlap)
+overlap_variable_loadings<-data.frame(overlap.res.var$coord) 
+## Screeplot showing percentage of explained variances
[…]

### 2023-12-01 code/Visualization.R

@@ -1,0 +1,112 @@
+# Add packages
+library(tidyverse)
+library(patchwork)
+
+# Data preparation
+## loading data
+gona_linear<-readr::read_csv("data/Gona_ESA_Linear1.csv")
+gona_3d<-read.csv("data/Gona_ESA_3D.csv")
+## creating a new column of unique identifier for merging
+gona_linear$Name <- paste(gona_linear$Site, gona_linear$Catalog, sep= "-")
+## merging two datasets based on Name
+joined_gona <-dplyr::left_join(gona_linear, gona_3d, by = c("Name" = "Name"))
+## Reordering column order for principle component analysis
+joined_gona <- joined_gona %>% dplyr::relocate(GM_L,GM_Br1,GM_Br2,GM_Br3,GM_T1,GM_T2,GM_T3)
+## Subsetting a dataframe that contains both linear and 3d data for measuremnt validation
+overlap_gona <- joined_gona %>% tidyr::drop_na(Flaked.y)
+
+## Figure for 3d comparison
+
+fig4a <- ggplot(overlap_gona, aes(X.3DGMPC110x10, Linear_PC1)) + 
+  geom_point() + 
+  geom_smooth(method=lm)+
+  labs(x ="3DGM PC1", y = "Linear PC1")
+
+fig4b <- ggplot(overlap_gona, aes(SDI.y, SDILinearCorrected)) + 
+  geom_point() + 
+  geom_smooth(method=lm)+
+  labs(x ="SDI (3DGM)", y = "SDILinearCorrected")
+
+fig4c <- ggplot(overlap_gona, aes(Flaked.y, FlakedVisual)) + 
+  geom_point() + 
+  geom_smooth(method=lm)+
+  labs(x ="Flaked percentage (3DGM)", y = "Flaked percentage (visual)")
+
+patchwork <- (fig4a + fig4b + fig4c)
+patchwork + plot_annotation(tag_levels = 'A')
+ggplot2::ggsave("Fig.3D comparison.png", path="figure.", width = 9, height = 3, dpi = 300)
+
+
+## Figure for FAI by base
+
+fig5a<-joined_gona %>%
+  filter(Contexts == "Acheulean", Mode == "2") %>%
+  ggplot(aes(FAI, PC1, color = as.factor(Base))) + 
+  geom_point(alpha=0.5) + 
[…]

### 2023-12-02 code/Visualization.R
figure revisions.

@@ -1,6 +1,7 @@
 # Add packages
 library(tidyverse)
 library(patchwork)
+library(cowplot)
 
 # Data preparation
 ## loading data
@@ -40,24 +41,27 @@ ggplot2::ggsave("Fig.3D comparison.png", path="figure.", width = 9, height = 3, 
 ## Figure for FAI by base
 
 fig5a<-joined_gona %>%
-  filter(Contexts == "Acheulean", Mode == "2") %>%
-  ggplot(aes(FAI, PC1, color = as.factor(Base))) + 
-  geom_point(alpha=0.5) + 
+  filter(Contexts == "Acheulean", Mode == "2", Flaked.x == "Flaked") %>%
+  ggplot(aes(FAI, PC1)) + 
+  geom_point(alpha=0.6,size=2,aes(color= as.factor(Base),shape=as.factor(Base))) + 
   geom_smooth(method=lm, aes(group = 1))+
   labs(x ="FAI", y = "PC1 (flatness)")+
   theme(legend.position="none")
 
+
 fig5b<-joined_gona %>%
-  filter(Contexts == "Acheulean", Mode == "2") %>%
-  ggplot(aes(FAI, PC2, color = as.factor(Base))) + 
-  geom_point(alpha=0.5) + 
+  filter(Contexts == "Acheulean", Mode == "2", Flaked.x == "Flaked") %>%
+  ggplot(aes(FAI, PC2)) + 
+  geom_point(alpha=0.6,size=2,aes(color = as.factor(Base),shape=as.factor(Base))) + 
   geom_smooth(method=lm, aes(group = 1))+
-  labs(x ="FAI", y = "PC2 (pointedness)")+
-  labs(color='Base')
+  labs(x ="FAI", y = "PC2 (convergence)")+
+  labs(color='Base', shape = 'Base')+
+  theme(legend.key.size = unit(0.2, "cm"))
+
 
 patchwork <- (fig5a + fig5b)
 patchwork + plot_annotation(tag_levels = 'A')
-ggplot2::ggsave("Fig.FAI by base.png", path="figure.", width = 7, height = 3, dpi = 300)
+ggplot2::ggsave("Fig.FAI by base new.png", path="figure.", width = 9, height = 3, dpi = 300)
 
 
 ## Figure for FAI by flaked
@@ -69,10 +73,11 @@ PC1.mean <- joined_gona %>%
 fig6a<-joined_gona %>%
[…]

### 2024-08-10 manuscript/manuscript.Rmd
new figure testing.

@@
 bibliography: bibliography.bib
-csl: apa.csl
+csl: frontiers.csl
 editor_options: 
@@
 ```{r FAIBASE, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Mode 2 trends across blank types."}
-knitr::include_graphics(here::here("figure/Fig.FAI by base.png"))
+knitr::include_graphics(here::here("figure/Fig.FAI by base new NEW.png"))
 ```
@@
 ```{r FAIFLAKED, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Effect of FAI compared to unmodified large flakes from Acheulean contexts."}
-knitr::include_graphics(here::here("figure/Fig.FAI by flaked.png"))
+knitr::include_graphics(here::here("figure/Fig.FAI by flaked new123.png"))
 ```

### 2024-08-10 code/Visualization.R

@@ -19,16 +19,18 @@ fig4a <- ggplot(gona_3d, aes(X.3DGMPC110x10, FAC1_2)) +
 fig4b <- ggplot(gona_3d, aes(SDI_3D, SDIBoxTimesPointFive)) + 
   geom_point() + 
   geom_smooth(method=lm)+
+  geom_abline(slope=1, intercept = 0, colour = "red", linewidth=1.2, alpha=0.5)+
   labs(x ="3D SDI", y = "cSDI")
 
 fig4c <- ggplot(gona_3d, aes(Flaked, FlakedVisual)) + 
   geom_point() + 
   geom_smooth(method=lm)+
+  geom_abline(slope=1, intercept = 0, colour = "red", linewidth=1.2, alpha=0.5)+       
   labs(x ="Flaked percentage (3D)", y = "Flaked percentage (visual)")
 
 patchwork <- (fig4a + fig4b + fig4c)
 patchwork + plot_annotation(tag_levels = 'A')
-ggplot2::ggsave("Fig.3D comparison new.png", path="figure.", width = 9, height = 3, dpi = 300)
+ggplot2::ggsave("Fig.3D comparison new1.png", path="figure.", width = 9, height = 3, dpi = 300)
 
 
 ## Figure for FAI by base

### 2024-08-12 code/Visualization.R
experimenting with new figures as requested by R3.

@@ -143,3 +143,35 @@ car::ncvTest(lmMod2)
 leftfig<-joined_gona %>%
   filter(Contexts == "Acheulean", Base == "flake")
 
+
+
+## figures for demonstrating the conceptual model as requested by R3
+gona_linear<- gona_linear %>% unite("FULLNAME", Locality:Catalog, remove = FALSE)
+### mapping SDI and FAI
+fig0a <- ggplot(gona_linear, aes(cSDI, FAI)) + 
+  geom_point() + 
+  geom_smooth(method=lm)+
+  ggrepel::geom_text_repel(aes(label=FULLNAME))+
+  labs(x ="cSDI", y = "FAI")
+fig0a < fig0a+
+
+### correlation between SDI/PC1 correlation and FAI/PC1
+  
+fig0b <- ggplot(gona_linear, aes(SDI_3D, SDIBoxTimesPointFive)) + 
+  geom_point() + 
+  geom_smooth(method=lm)+
+  geom_abline(slope=1, intercept = 0, colour = "red", linewidth=1.2, alpha=0.5)+
+  labs(x ="3D SDI", y = "cSDI")
+
+fig0c <- ggplot(gona_linear, aes(Flaked, FlakedVisual)) + 
+  geom_point() + 
+  geom_smooth(method=lm)+
+  geom_abline(slope=1, intercept = 0, colour = "red", linewidth=1.2, alpha=0.5)+       
+  labs(x ="Flaked percentage (3D)", y = "Flaked percentage (visual)")
+
+patchwork <- (fig0a + fig0b + fig0c)
+patchwork + plot_annotation(tag_levels = 'A')
+ggplot2::ggsave("Fig.CONCEPTUAL.png", path="figure.", width = 9, height = 9, dpi = 300)
+
+
+### correlation between SDI/PC2 correlation and FAI/PC2

### 2024-08-12 code/Visualization.R
experiment figures done.

@@ -148,30 +148,52 @@ leftfig<-joined_gona %>%
 ## figures for demonstrating the conceptual model as requested by R3
 gona_linear<- gona_linear %>% unite("FULLNAME", Locality:Catalog, remove = FALSE)
 ### mapping SDI and FAI
-fig0a <- ggplot(gona_linear, aes(cSDI, FAI)) + 
+fig0a <- ggplot(gona_linear, aes(cSDI, FAI, color = Typology)) +
   geom_point() + 
-  geom_smooth(method=lm)+
+  # geom_smooth(method=lm)+
   ggrepel::geom_text_repel(aes(label=FULLNAME))+
   labs(x ="cSDI", y = "FAI")
-fig0a < fig0a+
 
 ### correlation between SDI/PC1 correlation and FAI/PC1
-  
-fig0b <- ggplot(gona_linear, aes(SDI_3D, SDIBoxTimesPointFive)) + 
-  geom_point() + 
-  geom_smooth(method=lm)+
-  geom_abline(slope=1, intercept = 0, colour = "red", linewidth=1.2, alpha=0.5)+
-  labs(x ="3D SDI", y = "cSDI")
+correlation1<- gona_linear %>% group_by(Typology) %>%  summarise(SDIPC1R = cor(cSDI, FAC1_1))
+correlation2<- gona_linear %>% group_by(Typology) %>%  summarise(FAIPC1R = cor(FAI, FAC1_1))
+correlation12<- merge(correlation1,correlation2,by="Typology")
 
-fig0c <- ggplot(gona_linear, aes(Flaked, FlakedVisual)) + 
+### correlation between SDI/PC2 correlation and FAI/PC2
+
+correlation3<- gona_linear %>% group_by(Typology) %>%  summarise(SDIPC2R = cor(cSDI, FAC2_1))
+correlation4<- gona_linear %>% group_by(Typology) %>%  summarise(FAIPC2R = cor(FAI, FAC2_1))
+correlation34<- merge(correlation3,correlation4,by="Typology")
+  
+fig0b <- ggplot(correlation12, aes(SDIPC1R, FAIPC1R)) + 
   geom_point() + 
-  geom_smooth(method=lm)+
-  geom_abline(slope=1, intercept = 0, colour = "red", linewidth=1.2, alpha=0.5)+       
-  labs(x ="Flaked percentage (3D)", y = "Flaked percentage (visual)")
[…]

### 2024-08-17 code/Visualization.R
dip test for multimodality.

@@ -6,6 +6,7 @@ library(mvtnorm)
 library(MASS) 
 library(caret) 
 library(lsr)
+library(diptest)
 
 # Data preparation
 ## loading data
@@ -499,3 +500,10 @@ fig0c <- ggplot(pc2effectsize, aes(abs(estimate.x), abs(estimate.y))) +
 patchwork <- (fig0b + fig0c)
 patchwork + plot_annotation(tag_levels = 'A')
 ggplot2::ggsave("Fig.sdifai by typology(slope absolute value).png", path="figure.", width = 10, height = 5, dpi = 300)
+
+
+dip_test_result <- dip.test(lda.data$LD1)
+print(dip_test_result)
+
+dip_test_result <- dip.test(lda.data$LD2)
+print(dip_test_result)

### 2024-08-18 manuscript/manuscript.Rmd
started to update the manuscript.

@@
 bibliography: bibliography.bib
-csl: frontiers.csl
+csl: jhe.csl
 editor_options: 
@@
 
-The imposition of intended form on artifacts has long been viewed as a watershed in human cognitive and cultural evolution and is most commonly associated with the emergence of "Large Cutting Tools" (LCTs) in the Early Acheulean [@isaac1976; @kuhn2020; @holloway1969].
+The imposition of intended form on artifacts has long been viewed as a watershed in human cognitive and cultural evolution and is most commonly associated with the emergence of “Large Cutting Tools” (LCTs; cf.
+Long Core Tools [Shea, -@shea2013]) in the Early Acheulean [@isaac1976; @kuhn2020; @holloway1969].
 However, this interpretation of Acheulean LCTs as intentionally designed artifacts remains controversial.
-Alternative proposals range from the possibility that LCTs were unintended by-products of flake production [@moore2016; @noble1996] to the suggestion that their form was "at least partly under genetic control" [@corbey2016].
+Alternative proposals range from the possibility that LCTs were unintended by-products of flake production [@moore2016; @noble1996] to the suggestion that their form was “at least partly under genetic control” [@corbey2016: 6].
 Even accepting that LCT form was to some extent intended, there is substantial disagreement over the specificity of design.
-Some analyses have indicated that shape variation in Acheulean handaxes is largely a result of resharpening [@iovita2011; @mcpherron2000] whereas others find form to be unrelated to reduction intensity and more likely to reflect normative expectations of what handaxes should look like [@garcía-medrano2019; @shipton2015; @shipton2020].
-Such debates about shape of Acheulean LCTs may appear narrowly technical but have broad relevance for evolutionary questions including the origins of human culture [@corbey2016; @tennie2017; @shipton2015], language [@stout2012], teaching [@gärdenfors2017], brain structure [@hecht2015], and cognition [@stout2015; @wynn2016].
-To examine these questions, we studied the complete collection of Early Acheulean flaked pieces from 5 sites at Gona Project Area and compared them with Oldowan cores from 2 published sites at Gona.
-By comparing shape variation to measures of flaking intensity and patterning, we sought to identify technological patterns that might reveal intent.
+Some analyses have indicated that shape variation in Acheulean handaxes is largely a result of function-driven resharpening [@iovita2011; @mcpherron2000] whereas others find it more likely to reflect normative expectations of what handaxes should look like [@garcía-medrano2019; @shipton2015; @shipton2020].
+Such debates about the shape of Acheulean LCTs may appear narrowly technical but have broad relevance for evolutionary questions, including the origins of human culture [@corbey2016; @liu2023; @tennie2017; @shipton2015], language [@stout2012], teaching [@gärdenfors2017], brain structure [@hecht2015], and cognition [@stout2015; @wynn2016].
+To examine these questions, we studied the complete collection of Early Acheulean (ca.
+1.7 –1.2 Ma) flaked pieces from four sites (BSN17, DAN5, OGS12, and OGS5) in the Gona Project Area and compared them with all of the flaked pieces from two published Oldowan (\>2.5 Ma) sites at Gona (OGS7 and EG10; Semaw, 2000; Semaw et al., 2003; Stout et al., 2010).
+By comparing variation in overall artifact shape (or “Bauplan” [Lycett and Gowlett, 2008)) to measures of flaking intensity and coverage, we sought to identify technological patterns that might reveal intent.
+The possibilities we considered include: a) intentional shape standardization reflecting normative expectations of what handaxes should look like, b) regularities of overall form arising as by-products of the intentional pursuit of particular morpho-functional attributes, and c) the absence of any intentional form imposition as in the case of simple debitage.
 
@@
 [-@li2015] proposed a Flaked Area Index (FAI) as an alternative to SDI as a measure of reduction intensity, arguing that its validity is supported by an observed correlation (r=0.424) with SDI.
-However, they also explain that "flaked area does not necessarily relate to the number of flake scars\...a small number of large scars can produce a large area of scar coverage, and conversely, a large number of small scars can produce a small area of scar coverage." (p.
+However, they also explain that "flaked area does not necessarily relate to the number of flake scars...a small number of large scars can produce a large area of scar coverage, and conversely, a large number of small scars can produce a small area of scar coverage." (p.
 6).
@@
 A lack of shape correlation with either measure would be expected for simple debitage with no morphological targets whereas a strong correlation with both would indicate a highly "designed" form achieved through extensive morphological and volumetric transformation.
-(**Figure** \@ref(fig:FAISDI))
-In the current study we thus considered SDI and FAI together in order to evaluate evidence of intentional shaping in the early Acheulean of Gona.
+(**Figure** \@ref(fig:FAISDI)) In the current study we thus considered SDI and FAI together in order to evaluate evidence of intentional shaping in the early Acheulean of Gona.
 
@@
 
-```{r fig2, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Gona site map."}
-knitr::include_graphics(here::here("figure/site map.png"))
-```
+<!-- ```{r fig2, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Gona site map."} -->
+
+<!-- knitr::include_graphics(here::here("figure/site map.png")) -->
+
+<!-- ``` -->
 
@@
 Generalized Procrustes and Principal Component analyses were then undertaken to explore the shape variability of the sample.
-The surface area of each artifact was calculated using the *Artifact3-D* program of Grosman et al.
-[-@grosman2022].
-*Artifact3-D* was also used to automatically identify the flake scar boundaries and compute each scar's surface area, using the scar analysis functions of Richardson et al.
-[-@richardson2014].
+The surface area of each artifact was calculated using the *Artifact3-D* program of Grosman et al. [-@grosman2022].
+*Artifact3-D* was also used to automatically identify the flake scar boundaries and compute each scar's surface area, using the scar analysis functions of Richardson et al. [-@richardson2014].
 
@@
 Together, 3DGM-PC2 (r = 0.344, p = 0.050), 3DGM-PC3 (r = -0.416, p = 0.016), 3DGM-PC4 (r = 0.458, p = 0.007), and 3DGM-PC5 (r = -0.352, p = 0.044) correlate well with Linear-PC2, cumulatively capturing whether the items are broad and flat or thick and pointed.
-A stepwise regression ($r^2$=0.625, F(4,28)=11.697, p<0.001, Probability-of-F-to-enter <= 0.050; Probability-of-F-to-remove >= 0.100) with Linear-PC2 as the dependent variable retained all four of these 3DGM-PCs as significant predictors.
+A stepwise regression ($r^2$=0.625, F(4,28)=11.697, p\<0.001, Probability-of-F-to-enter \<= 0.050; Probability-of-F-to-remove \>= 0.100) with Linear-PC2 as the dependent variable retained all four of these 3DGM-PCs as significant predictors.
 
@@
 ```{r figcomparison, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Comparison between linear measurement and 3d measurement."}
-knitr::include_graphics(here::here("figure/Fig.3D comparison.png"))
+knitr::include_graphics(here::here("figure/Fig.3D comparison new1.png"))
 ```
@@
 
-Visual estimation of flaked area approximated 3DGM measurement very closely (r2=0.932, p < 0.001, $\beta$ = 1.051; **Figure** \@ref(fig:figcomparison)C) and without any systematic bias (paired t-test: mean difference = -0.015%, 95% CI = 2.02% to 1.99%, p = 0.987).
+Visual estimation of flaked area approximated 3DGM measurement very closely (r2=0.932, p \< 0.001, $\beta$ = 1.051; **Figure** \@ref(fig:figcomparison)C) and without any systematic bias (paired t-test: mean difference = -0.015%, 95% CI = 2.02% to 1.99%, p = 0.987).
 Individual errors ranged between -10.46% and 14.24%.

### 2024-08-18 manuscript/manuscript.Rmd
manuscript update.

@@
 The imposition of intended form on artifacts has long been viewed as a watershed in human cognitive and cultural evolution and is most commonly associated with the emergence of “Large Cutting Tools” (LCTs; cf.
-Long Core Tools [Shea, -@shea2013]) in the Early Acheulean [@isaac1976; @kuhn2020; @holloway1969].
+Long Core Tools [Shea, [-@shea2013]]) in the Early Acheulean [@isaac1976; @kuhn2020; @holloway1969].
 However, this interpretation of Acheulean LCTs as intentionally designed artifacts remains controversial.
@@
 ## Identifying design
+
+There is a broad consensus that refined handaxes and cleavers known from some later Acheulean (\<1.0 Ma) contexts resulted from procedurally elaborate, skill intensive and socially learned production strategies [\@caruana2020; \@garcía-medrano2019; \@moore2020; \@sharon2009; \@shiptonEvolutionSocialTransmission2019; \@stout2014], although debate over the presence of explicit, culturally transmitted shape preferences not determined by function continues [\@iovita2011; \@moore2020; \@shipton2020; \@wynn2018].
+However, archaeologists have also noted the presence of less heavily worked and morphologically standardized LCTs (e.g., handaxes, cleavers, picks) in earlier Acheulean contexts (Leakey et al., 1951; Kleindienst, 1962; Wynn, 1979; Gowlett, 1986; Clark, 1994; Ludwig and Harris, 1998; Stout, 2011; Presnyakova et al., 2018).
+Such forms are typical of Acheulean assemblages in Africa prior to 1.0 Ma (e.g., Kleindienst, 1962; Ludwig and Harris, 1998; Lepre et al., 2011; Beyene et al., 2013; Diez-Martín et al., 2015; de la Torre and Mora, 2018; Presnyakova et al., 2018; Semaw et al., 2018), but continue to occur with variable frequency in later time periods [\@mcnabb2015] and may be especially prevalent in eastern Asian [\@li2021].
+Although formal tool types are commonly used to describe such early Acheulean assemblages, many workers now see a continuum of morphological variation [\@duke2021; \@kuhn2020; \@presnyakova2018] including the possibility that simple flake production remained an important [@shea2010] or even primary [@moore2016] purpose of reduction.
 

### 2024-08-18 manuscript/manuscript.Rmd
manuscript update.

--- 2024-08-18 manuscript/manuscript.Rmd
@@
 The imposition of intended form on artifacts has long been viewed as a watershed in human cognitive and cultural evolution and is most commonly associated with the emergence of “Large Cutting Tools” (LCTs; cf.
-Long Core Tools [Shea, -@shea2013]) in the Early Acheulean [@isaac1976; @kuhn2020; @holloway1969].
+Long Core Tools [@shea2013]) in the Early Acheulean [@isaac1976; @kuhn2020; @holloway1969].
 However, this interpretation of Acheulean LCTs as intentionally designed artifacts remains controversial.
@@
 To examine these questions, we studied the complete collection of Early Acheulean (ca.
-1.7 –1.2 Ma) flaked pieces from four sites (BSN17, DAN5, OGS12, and OGS5) in the Gona Project Area and compared them with all of the flaked pieces from two published Oldowan (\>2.5 Ma) sites at Gona (OGS7 and EG10; Semaw, 2000; Semaw et al., 2003; Stout et al., 2010).
-By comparing variation in overall artifact shape (or “Bauplan” [Lycett and Gowlett, 2008)) to measures of flaking intensity and coverage, we sought to identify technological patterns that might reveal intent.
+1.7–1.2 Ma) flaked pieces from four sites (BSN17, DAN5, OGS12, and OGS5) in the Gona Project Area and compared them with all of the flaked pieces from two published Oldowan (\>2.5 Ma) sites at Gona (OGS7 and EG10; @semaw2000; @semaw2003; @stoutTechnologicalVariationEarliest2010).
+By comparing variation in overall artifact shape (or “Bauplan” [@lycettQuestionsSurroundingAcheulean2008]) to measures of flaking intensity and coverage, we sought to identify technological patterns that might reveal intent.
 The possibilities we considered include: a) intentional shape standardization reflecting normative expectations of what handaxes should look like, b) regularities of overall form arising as by-products of the intentional pursuit of particular morpho-functional attributes, and c) the absence of any intentional form imposition as in the case of simple debitage.
@@
 
-There is a broad consensus that refined handaxes and cleavers from the later Acheulean resulted from procedurally elaborate, skill intensive, and socially learned production strategies [@caruana2020; @garcía-medrano2019; @moore2020; @sharon2009; @shiptonEvolutionSocialTransmission2019; @stout2014] although debate over the presence of explicit, culturally transmitted shape preferences continues [@iovita2011; @moore2020; @shipton2020; @wynn2018].
-There is much less agreement regarding the less heavily worked and formally standardized LCTs typical of the earliest Acheulean [@beyene2013; @delatorre2018; @diez-martín2015; @lepre2011; @semaw2018].
-Such forms continue to occur with variable frequency in later time periods [@mcnabb2015], and may be especially prevalent in eastern Asia [@li2021].
-Although formal types have been recognized in the Early Acheulean and are commonly used to describe assemblages, many workers now see a continuum of morphological variation [@duke2021; @kuhn2020; @presnyakova2018] including the possibility that simple flake production remained an important [@shea2010] or even primary [@moore2016] purpose of Early Acheulean large core reduction.
+There is a broad consensus that refined handaxes and cleavers known from some later Acheulean (\<1.0 Ma) contexts resulted from procedurally elaborate, skill intensive and socially learned production strategies [@caruana2020; @garcía-medrano2019; @moore2020; @sharon2009; @shiptonEvolutionSocialTransmission2019; @stout2014], although debate over the presence of explicit, culturally transmitted shape preferences not determined by function continues [@iovita2011; @moore2020; @shipton2020; @wynn2018].
+However, archaeologists have also noted the presence of less heavily worked and morphologically standardized LCTs (e.g., handaxes, cleavers, picks) in earlier Acheulean contexts [@leakey1951; @kleindienst1962; @wynn1979; @gowlett1986; @clark1994; @ludwig1998; @stoutStoneToolmakingEvolution2011; @presnyakova2018].
+Such forms are typical of Acheulean assemblages in Africa prior to 1.0 Ma [e.g., @kleindienst1962; @ludwig1998; @lepre2011; @beyene2013; @diez-martín2015; @delatorre2018; @presnyakova2018; @semaw2018], but continue to occur with variable frequency in later time periods [@mcnabb2015] and may be especially prevalent in eastern Asian [@li2021].
+Although formal tool types are commonly used to describe such early Acheulean assemblages, many workers now see a continuum of morphological variation [@duke2021; @kuhn2020; @presnyakova2018] including the possibility that simple flake production remained an important [@shea2010] or even primary [@moore2016] purpose of reduction.
 

### 2024-08-19 manuscript/manuscript.Rmd
finished another round of proofreading.

@@
 abstract:
+|
+  ```{=latex}
+  The appearance of ‘Large Cutting Tools’ in the early Acheulean is widely regarded as the first evidence for the imposition of intended forms on artifacts, with major implications for hominin cognitive and cultural capacities.
+However, the nature and extent of explicit design documented by these forms remains open to debate.
+To address this issue, we analyzed the complete collection of early Acheulean (ca.
+1.7–1.2 Ma) flaked pieces from four sites (BSN17, DAN5, OGS12, and OGS5) in the Gona Project Area and compared these with all of the flaked pieces from two published Oldowan (>2.5 Ma) sites at Gona.
+By comparing shape variation to measures of flaking intensity and coverage, we sought to identify technological patterns indicative of intent.
+Results provide little support for the presence of discrete tool types or imposed morphological norms in our sample.
+Instead, we observe systematic patterns of raw material selection and core surface modification aimed at the production and maintenance of useful cutting edges on relatively large supports (cobbles and large flake blanks).
+This is consistent with prior characterizations of Acheulean tool form as arising from functional and ergonomic design imperatives for large hand-held cutting tools.
+Although the generalizability of these results to other sites remains to be seen, we propose that distinctive early Acheulean artifact forms may represent secondary accommodations to the primary goal of increasing tool size to meet the novel demands (e.g., extended use-life, enhanced transportability, utility for heavy-duty cutting) of a more general shift in hominin behavioral ecology at this time.
+Our results thus provide support for the presence of Acheulean design, not as the concrete realization of a ‘holographic model’ in the mind of the maker, but rather in the broader sense of deliberate technological choices made in view of behavioral goals and material constraints.
 
+  \textbf{Keywords:} Lithic technology; Early Stone Age; Handaxe; Large Cutting Tool; Pick; Shaping
+  ```
 date: "`r Sys.Date()`"
@@
 csl: jhe.csl
-editor_options: 
-  markdown: 
-    wrap: sentence
 ---
@@
 
-The imposition of intended form on artifacts has long been viewed as a watershed in human cognitive and cultural evolution and is most commonly associated with the emergence of “Large Cutting Tools” (LCTs; cf.
-Long Core Tools [@shea2013]) in the Early Acheulean [@isaac1976; @kuhn2020; @holloway1969].
+The imposition of intended form on artifacts has long been viewed as a watershed in human cognitive and cultural evolution and is most commonly associated with the emergence of “Large Cutting Tools” [LCTs; cf.
+Long Core Tools; @shea2013] in the Early Acheulean [@isaac1976; @kuhn2020; @holloway1969].
 However, this interpretation of Acheulean LCTs as intentionally designed artifacts remains controversial.
@@
 To examine these questions, we studied the complete collection of Early Acheulean (ca.
-1.7–1.2 Ma) flaked pieces from four sites (BSN17, DAN5, OGS12, and OGS5) in the Gona Project Area and compared them with all of the flaked pieces from two published Oldowan (\>2.5 Ma) sites at Gona (OGS7 and EG10; @semaw2000; @semaw2003; @stoutTechnologicalVariationEarliest2010).
+1.7–1.2 Ma) flaked pieces from four sites (BSN17, DAN5, OGS12, and OGS5) in the Gona Project Area and compared them with all of the flaked pieces from two published Oldowan (\> 2.5 Ma) sites at Gona [OGS7 and EG10; @semaw2000; @semaw2003; @stoutTechnologicalVariationEarliest2010].
 By comparing variation in overall artifact shape (or “Bauplan” [@lycettQuestionsSurroundingAcheulean2008]) to measures of flaking intensity and coverage, we sought to identify technological patterns that might reveal intent.
@@
 
-Typologically, LCTs are differentiated from Mode 1 pebble cores on the basis of size (\>10cm) and shape (elongation and flattening) [e.g., @isaac1977].
-This consistent production of large, flat, and elongated cores in the Achuelean has long been thought to reflect the pursuit of desired functional and ergonomic properties for hand-held cutting tools [@wynn2018].
-Unplanned flaking can sometimes produce cores that fall into the LCT shape range [@moore2016] and this is one possible explanation of the relatively small "protobifaces" that occur in low frequencies in Oldowan assemblages [@isaac1997].
-However, the Early Acheulean is clearly distinguished from the Oldowan by the production of larger artifacts necessitating the procurement and exploitation of larger raw material clasts.
-Although studies of handaxe variation often focus on shape rather than size, this shift is an important aspect of artifact design with relevance to both production and function.
+Classically, LCTs are differentiated from Mode 1 pebble cores on the basis of size (\> 10 cm) and basic aspects of overall artifact shape including elongation and flattening [@isaac1977] and the creation of a point [@kleindienst1962] .
+The consistent production of large, flat, elongated, and pointed forms in the Acheulean is generally thought to reflect the pursuit of desired functional and ergonomic properties for hand-held cutting tools [@wynn2018].
+Within this broad frame, numerous studies have examined the variation in overall LCT shape captured by various combinations of length, breadth, and thickness measures in an attempt to disentangle cultural, functional, raw material, allometric, and resharpening influences [e.g., @roe1969; @crompton1993; @mcpherron2000].
+It has been demonstrated that unplanned flaking can sometimes produce cores that fall into the LCT shape range [@moore2016] and this is one possible explanation of the relatively small “protobifaces” that occur in low frequencies in Oldowan assemblages [@isaac1997].
+However, the early Acheulean is also distinguished from the Oldowan by the production of larger artifacts necessitating the procurement and exploitation of larger raw material clasts.
+Although studies of handaxe variation often focus on shape rather than size, this size increase is an important aspect of artifact design with relevance to both production and function.
 
-Production of larger tools was accomplished either through a novel process of detaching and working Large Flake Blanks (LFBs) from boulder cores or simply by using larger cobble and slab cores [@delatorre2018; @isaac1969; @semaw2018].
-Both may involve similar flaking "strategies" (e.g., bifacial or multifacial exploitation) to those present in the Oldowan [@duke2021] but require more forceful percussion to detach larger flakes.
-This increases the perceptual motor difficulty of the task [@stout2002] and in many cases may have been accomplished using different percussive techniques and supports [@semaw2009].
-These new challenges would have increased raw material procurement [@shea2010] and learning costs [@pargeter2019] as well as the risk of serious injury [@gala2023] associated with tool production.
-This strongly implies intentional pursuit of offsetting functional benefits related to size increase.
-These likely included tool ergonomics and performance [@key2017] as well as flake generation, resharpening, and reuse potential [@shea2010].
-Early Acheulean LCT production is thus widely seen as a part of shifting hominin behavioral ecological strategies including novel resources and mobility patterns [@linaresmatás2021; @rogers1994].
+Production of larger tools was accomplished either through a novel process of detaching and working Large Flake Blanks (LFBs) from boulder cores or by selecting larger cobble and slab cores [@delatorre2018; @isaac1969; @semaw2018].
+Both approaches may involve similar flaking “strategies” (e.g., bifacial or multifacial exploitation) to those present in the Oldowan [@duke2021] but require more forceful percussion to detach larger flakes.
+This increases the perceptual-motor difficulty of the task [@stout2002] and in many cases may have been accomplished using different percussive techniques and supports [@semaw2009].
+These new challenges would have increased raw material procurement [@shea2010] and learning [@pargeter2019] costs, as well as the risk of serious injury [@gala2023].
+These costs strongly imply intentional pursuit of offsetting functional benefits related to size increase.
+Such benefits likely included tool ergonomics and performance [@key2017; @keyReassessingProductionHandaxes2017], as well as flake generation, resharpening, and reuse potential [@shea2010].
+Early Acheulean LCT production is thus widely seen as a part of shifting hominin behavioral ecological strategies including novel resources and mobility patterns[@linaresmatás2021; @rogers1994; @hay1976].
+Importantly, such functional and ecological interpretation speaks primarily to motivation (goals), and need not imply a lack of hominin technological understanding and intention.
 
-The degree of intentional design reflected in the shape of Early Acheulean LCTs is more difficult to determine.
-For example, LFB production using a simple "least effort" bifacial/discoidal strategy will tend to generate predominantly elongated (side or end struck) flakes [@toth1982] whether or not this is an intentional design target.
-Similarly, the difficulty of flaking relatively spherical cobbles [@toth1982] might bias initial clast selection and subsequent reduction toward flat and elongated shapes even in the absence of explicit design targets.
-On the other hand, it has been argued that the shape of Early Acheulean LFBs was intentionally predetermined using core preparation techniques [@delatorre2018] and many researchers perceive efforts at intentional shaping in the organization of flake scars on Early Acheulean handaxes and picks [@beyene2013; @delatorre2018; @diez-martín2015; @duke2021; @lepre2011; @semaw2009].
-To date, however, the identification of Early Acheulean shaping has generally relied on qualitative assessment by lithic analysts.
-Such assessment may in fact be reliable, but is subject to concerns about potential selectivity, bias, and/or overinterpretation [@davidson2002; @moore2016].
-Notably, a 3-dimensional morphometric (3DGM) study by Presnyakova, et al.
-[-@presnyakova2018] concluded that LCT shape variation in the Okote Member (\~1.4 mya) at Koobi Fora was largely driven by reduction intensity rather than different knapping strategies.
+The degree of intentional design reflected in the overall shape of Early Acheulean LCTs is more difficult to determine.
+For example, LFB production using a simple “least effort” bifacial/discoidal strategy will tend to generate predominantly elongated (side or end struck) flakes [@toth1982; @leader2018] whether or not this is an intentional design target.
+Similarly, the difficulty of flaking relatively spherical cobbles [@toth1982] might bias initial clast selection and subsequent reduction toward flat and elongated shapes, even in the absence of normative shape targets.
+Nevertheless, it has been argued that the shape of Early Acheulean LFBs was predetermined using intentional core preparation techniques [@delatorre2018; @leader2018] and many researchers perceive efforts at intentional shaping in the organization of flake scars on Early Acheulean handaxes and picks [@beyene2013; @delatorre2018; @diez-martín2015; @lepre2011; @semaw2018; @presnyakova2018].
+
+To date, evidence of early Acheulean shaping has generally come from qualitative assessments by lithic analysts.
+We do not dispute these assessments, but note that they are subject to reasonable concerns about possible selectivity, bias and/or overinterpretation [e.g., @davidson2002; @moore2016].
+The attribution of overall artifact shape imposition would be more compelling if supported by multiple lines of qualitative and quantitative evidence [i.e., a consilience of inductions; @whewell1837].
+This is particularly true as qualitative and quantitative approaches often have complementary strengths and weaknesses.
+Currently, however, quantitative methods bearing on this question are underdeveloped and evidence is mixed.
+A three-dimensional geometric morphometric (3DGM) study by Presnyakova et al.
+[-@presnyakova2018] concluded that LCT shape variation in the Okote Member (ca.
+1.4 Ma) at Koobi Fora was largely driven by reduction intensity rather than different knapping strategies.
 However, this study did not directly address the presence/absence of design targets constraining the observed range of variation.
+Shipton et al.
+[-@shipton2019] found that there was a correlation between LCT symmetry and reduction intensity (scar density index) in the lightly worked cleavers from Olorgesailie CL1-1 and conclude that this, along with several examples of flake scar patterning, provides evidence of the intentional imposition of symmetry.
+However, they remain agnostic as to whether this was motivated by functional, aesthetic, or social goals, leaving open the possibility that overall symmetry was a byproduct of more localized morpho-functional objectives.
+The purpose of the current study is to add to these various lines of evidence by developing and applying novel quantitative methods for assessing the imposition of overall artifact form in the early Acheulean at Gona.
 
-In later Acheulean contexts, reduction intensity effects are commonly equated with resharpening and seen as an alternative to intentional form imposition [@mcpherron2000].
-Across heavily-worked and relatively standardized LCT assemblages [e.g., @shipton2020], a *lack* of association between morphology and reduction intensity has be used as an argument-by-elimination for the presence of imposed morphological norms or "mental templates" [@garcía-medrano2019; @shipton2015; @shipton2023].
-However, in the less heavily-worked and more heterogeneous assemblages typical of the early Acheulean [@kuhn2020], it is equally plausible that increasing reduction intensity would reflect degree of primary reduction rather than subsequent resharpening [@archer2010].
-In this case, reduction intensity effects on morphology would have the opposite interpretation: more reduction should result in closer approximation of a desired form if such were present.
-For example, Beyene, et al.
+In later Acheulean contexts, reduction intensity effects are commonly equated with resharpening and seen as supporting an alternative “byproduct” explanation of artifact shape regularity, rather than intentional form imposition [@mcpherron2000].
+Across heavily worked and relatively standardized LCT assemblages [e.g., @shipton2020], a lack of association between morphology and reduction intensity has thus been used as an argument-by-elimination for the presence of imposed morphological norms or “mental templates” [@garcía-medrano2019; @shipton2015; @shipton2023].
+However, in the less heavily worked and more heterogenous assemblages typical of the early Acheulean [@kuhn2020], reduction intensity is more plausibly interpreted as an indicator of the extent of primary reduction rather than subsequent resharpening [@archer2010; @shipton2019].
+In this case associations between reduction intensity and shape would indicate intended effects of primary reduction, rather than byproducts of subsequent rounds of resharpening.
+For example, Beyene et al.
 [-@beyene2013] found that increasing flake scar counts were associated with increasing handaxe refinement through time at Konso, Ethiopia, which may reflect a more general trend in the African Acheulean [@shipton2018].
+Disentangling such alternative explanations is a major challenge for quantitative approaches to the question of shape imposition.
 
-Interpretive approaches address this quandary by "reading" the organization of scars on individual pieces to infer intent, but an adequate method to objectively quantify these insights has yet to be developed.
-Current measures of reduction intensity, such as the scar density index (SDI) [@clarkson2013], are designed to estimate total mass removed from a core and are reasonably effective [@lombao2023].
+Interpretive approaches address this quandary by “reading” the organization of scars on individual pieces to infer intent [@pelegrin2005], but an adequate method to objectively quantify these insights has yet to be developed.
+Lin et al.
+[-@lin2024] recently developed a method for quantifying flake scar orientations patterns, however this method still requires reading the direction of individual scars and it is not yet clear how results might be related to the question of shape imposition.
+The WEAP method of Garcia-Medrano et al.
+[-@garcia-medranoWEAPMethodNew2020] similarly provides a useful framework for integrating qualitative/interpretive technological attributes with quantitative shape metrics but does not address the reliability of individual attributes.
+Current measures of reduction intensity, such as the scar density index [SDI; @clarkson2013], are designed to estimate total mass removed from a core and have been shown to be reasonably effective [@lombao2023].
 However, mass removal was not the objective of Paleolithic flaking.
-Indeed, knapping efficiency is usually conceived as generating an outcome while *minimizing* required mass removal.
+Indeed, knapping efficiency is usually conceived as generating an outcome while minimizing required mass removal and energy expenditure [@rezek2018].
 This is true whether the desired outcome is a useful flake, a rejuvenated edge, or a particular core morphology.
-In simple flake production, mass removed is probably a good reflection of the completeness of exploitation ("exhaustion") of cores and may have implications for required skill [@toth1982; @pargeter2023] as well as raw material economy [@reeves2021; @shick1987].
-However, in core shaping and resharpening, mass removal would typically represent an energetic and raw material cost to be minimized, and might even interfere with function [@keyHandaxeShapeVariation2019].
+In simple flake production, mass removed is probably a good reflection of the completeness of exploitation (“exhaustion”) of cores and may have implications for required skill [@toth1982; @pargeter2023], as well as raw material economy [@reeves2021; @shick1987].
+However, in core shaping and resharpening, mass removal would typically represent an energetic and raw material cost to be minimized and might even interfere with function [@keyHandaxeShapeVariation2019].
 Without further information, relationships between artifact shape and reduction intensity are thus open to conflicting interpretations as evidence of intentional design or its absence.
 
-Li, et al.
-[-@li2015] proposed a Flaked Area Index (FAI) as an alternative to SDI as a measure of reduction intensity, arguing that its validity is supported by an observed correlation (r=0.424) with SDI.
-However, they also explain that "flaked area does not necessarily relate to the number of flake scars...a small number of large scars can produce a large area of scar coverage, and conversely, a large number of small scars can produce a small area of scar coverage." (p.
-6).
-We suggest that what FAI actually captures is the spatial extent modifications to the surface of a core.
-It is thus complementary to the measure of volume reduction provided by SDI and provides additional information to inform technological interpretations.
-For example, a correlation between FAI and artifact form without any effect of SDI would suggest a focus on "least-effort" shape imposition whereas the opposite pattern would be consistent with relatively intense resharpening of spatially restricted areas on the core.
-A lack of shape correlation with either measure would be expected for simple debitage with no morphological targets whereas a strong correlation with both would indicate a highly "designed" form achieved through extensive morphological and volumetric transformation.
-(**Figure** \@ref(fig:FAISDI)) In the current study we thus considered SDI and FAI together in order to evaluate evidence of intentional shaping in the early Acheulean of Gona.
+Li et al.
+[-@li2015] proposed a Flaked Area Index (FAI) as an alternative to SDI as a measure of reduction intensity, arguing that its validity is supported by a moderate correlation ($r = 0.424$) with SDI.
+However, they also explain that “flaked area does not necessarily relate to the number of flake scars...a small number of large scars can produce a large area of scar coverage, and conversely, a large number of small scars can produce a small area of scar coverage.” [@li2015: 6].
+We suggest that FAI actually captures the spatial extent of modifications to the surface of a core, rather than the volume of material removed.
+It is thus complementary to the measure of volume reduction provided by SDI and provides additional information to inform technological interpretations (**Fig.** \@ref(fig:fig1)).
+For example, a correlation between FAI and artifact form without any effect of SDI would suggest a focus on “least-effort” shape imposition through large and/or strategically placed flake removals.
+On the other hand, an association between SDI and core form with no effect of FAI would be consistent with reduction (shaping and/or resharpening) focusing on spatially restricted areas (e.g., non-invasive retouch) on the core.
+In both cases we would also expect a breakdown in the expected association between FAI and SDI.
+For simple debitage with no systematic morphological constraints we would expect a lack of any effect of FAI or SDI on core shape.
+Conversely, strong correlation of core shape with both FAI and SDI would indicate a highly “designed” form achieved through extensive surface modification and volumetric reduction.
+In the current study we employed these theoretical expectations to evaluate evidence of the presence and nature of intentional shaping in the early Acheulean of Gona.
 
-```{r FAISDI, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A conceptual model of inferring intention from artifact morphology."}
-knitr::include_graphics(here::here("figure/FAI_SDI.png"))
+```{r fig1, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Technological interpretations of the relationship between artifact shape, flaked area, and scar density.
+Abbreviations: SDI: corrected scar density index; FAI: Flaked area index."}
+knitr::include_graphics(here::here("figure/Fig.1.png"))
 ```
@@
 
-Three-dimensional scanning and geometric morphometric (3DGM) methods are increasingly common in the study of LCT form and reduction intensity [@shipton2015; @caruana2020; @presnyakova2018; @li2021; @archer2010; @li2015; @lycett2006].
-These methods provide high-resolution, coordinate-based descriptions of artifact form including detailed information about whole object geometric relations that is not captured by conventional linear measures [@shott2010].
-This includes measures of surface area used to compute both SDI and FAI measures [@clarkson2013; @li2015].
-At the time of writing, however, 3D scans are available for only a small number of Gona artifacts, including 33 of the Oldowan and Acheulean flaked pieces used in this study (**Figure** \@ref(fig:fig1)).
-Despite continuing improvements, 3DGM methods still impose additional costs in terms of data collection and processing time as well as required equipment, software, and training.
-Importantly, 3DGM methods cannot be applied to pre-existing photographic and metric data sets [e.g., @marshall], including available data from Gona.
-For this reason, and to better understand the relative costs and benefits of 3DGM more generally, we sought to test the degree to which conventional measurements can approximate 3DGM methods and produce reliable results by directly comparing our conventional measures with 3DGM analysis of the 33 available scans.
+For our study, we are specifically concerned with the accurate description of morphological variation and estimation of artifact surface and flaked areas.
+Historically, debates over imposed form in the Acheulean have focused on basic aspects of overall artifact shape, such as elongation, flattening, and pointedness that are relatively easy to capture using sparse data.
+For example, McNabb [-@mcnabb2022] recently concluded that the influential “Roe method” using 7 linear measurements is sufficient to “fairly represent the basic *bauplan* of LCTs.” Typological groupings derived in this fashion have similarly been validated [@shipton2020] by analyses based on the much more detailed morphological information provided by three-dimensional geometric morphometric (3DGM) methods.
+Across 3DGM studies, the first two or three principle components (PCs) typically account for about 50% of variation and consistently correspond to basic features like elongation, relative thickness, pointedness, and position of maximum thickness that also emerge from lower resolution spatial data [@lycett2006; @archer2010; @garcía-medrano2019] and studies employing linear measures rather than spatial coordinates [@crompton1993; @pargeter2019].
+We are thus confident that the seven linear measures used in our study effectively capture relevant aspects of shape variation for our research question.
+To further confirm this, we compared shape PCs derived from our linear measures with the results of a 3DGM analysis of all scanned artifacts in our sample ($n = 31$; **Fig.** \@ref(fig:fig2)).
 
-```{r fig1, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="3D Artifact scans from Gona used in this study."}
-knitr::include_graphics(here::here("figure/Fig1.png"))
+```{r fig2, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="3D Artifact scans from Gona used in this study.Artifact maximum dimension is vertical, and three views shown 90 degree rotation.
+The pieces, catalog number and typological classifications are: EG10: 1125 M1 FC (a), 1171 M1 FC (b), 1215 M1 FC(c), 1223 M1 FC (d).
+OGS7: 334 M1 FC (e), 335 M1 FC (f), 437 M1 FF (g), 485 M1 FC (h), 486 M1 FIB (i), 2016 “M1 FF “protopick” (j), 523 M1 FC (k), 2006 M1 FF (l), 2066 M1 FC (m), 2081 M1 FIB (n), 2092 M1 FC (o).
+OGS12: 16 M2 FC “handaxe” (p), 63 M2 FF “handaxe” (q), 226 M2 FF “handaxe” (r), 13 M2 FF “pick” (s), 74 M2 FC “pick” (t), 102 M2 FF “pick” (u), 2012 M2 FF “pick” (v), 115 M2 FF “knife” (w), 110 M2 FF “knife” (x), 113 M2 FF “knife” (y).
+BSN17: 35 M2 FC “handaxe” (z), 21 M2 FC “pick”  (aa), 24 M2 FC “pick” (ab), 34 M2 FC “pick” (ac), 29 M2 FC “broken pick” (ad), 231 M1 FC “core” (ae) (M1: Mode 1, M2: Mode 2, FC: Flaked Cobble, FF: Flaked Flake, FIB: Flaked Indeterminate Base)"}
+knitr::include_graphics(here::here("figure/Fig.2.png"))
 ```
 
-For our study, we are specifically concerned with the accurate description of morphological variation and estimation of artifact surface and flaked areas.
-With respect to morphology, we were encouraged by the fact that aspects of form variation relevant to our research question (i.e., core elongation, flattening, and pointedness) are relatively simple to describe using sparse data.
-3DGM studies of LCTs collect vast amounts of shape data but may discard upward of 50% of the observed variation in order to focus on two or three interpretable principal components.
-Across studies, these PCs consistently correspond to basic features like elongation, relative thickness, pointedness, and position of maximum thickness that also emerge from lower-resolution spatial data [@garcía-medrano2019; @archer2010; @lycett2006] and studies employing linear measures rather than spatial coordinates [@pargeter2019; @crompton1993].
 There is less evidence that conventional methods can accurately estimate artifact surface and flaked areas.
 Clarkson [-@clarkson2013] advocated the use of 3D surface area measures as more accurate than estimation from linear measures (e.g., surface area of a rectangular prism defined by artifact dimensions).
-However, he also found that the error introduced by the linear approach was a highly systematic, isometric overestimation of surface area and that results correlated with direct 3D measures with an impressive r^2^ = 0.944 and no effect of variation in core shape.
+However, he also found that the error introduced by the linear approach was a highly systematic, isometric overestimation of surface area and that results correlated with direct 3D measures with an impressive $r^2 = 0.944$ and no effect of variation in core shape.
 Insofar as it is variation in the relationship between surface area and flaking intensity that is of interest, rather than absolute artifact size, such systematic overestimation may not be problematic.
 Similar concerns apply to the estimation of flaked area.
-Traditionally, such estimates have been done "by eye" as a percentage of the total artifact surface [e.g., @dibble2005].
+Traditionally, such estimates have been done “by eye” as a percentage of the total artifact surface [e.g., @dibble2005].
 Such estimations have been found to be reasonably accurate when compared to 3D methods, but with the potential for substantial error on individual artifacts [@lin2010].
-The accuracy of by-eye estimation has yet to be systematically studied in Early Stone Age cores like those from Gona.
+The accuracy of visual estimation has yet to be systematically studied in Early Stone Age cores like those from Gona.
+We addressed this question by comparing SDI and FAI estimates derived from linear measures and “by eye” estimation with the results of 3DGM analysis.
 
@@
 
-Early Acheulean sites in the Gona Project area (**Figure** \@ref(fig:fig2)) are distributed over a wide area within the Dana Aoule North (DAN), Ounda Gona South (OGS), and Busidima North (BSN) drainages in the Busidima Formation [@quade2004] and range in age from approximately 1.7 to 1.2 mya [@semaw2018].
-The specific sites included in the current analysis are DAN-5, OGS-5, OGS-12, and BSN-17, all estimated to ca.
-1.7 -- 1.4 mya by stratigraphic position in the Gona sequence [@quade2008; @semaw2020].
-The Busidima Formation accumulated through fluvial deposition by the ancestral Awash River (Type I context) and its smaller tributary channels (Type II context) [@quade2004; @quade2008].
+Early Acheulean sites in the Gona Project area are distributed over a wide area within the Dana Aoule North (DAN), Ounda Gona South (OGS), and Busidima North (BSN) drainages in the Busidima Formation [@quade2004] and range in age from approximately 1.7–1.2 Ma [@semaw2018].
+The specific sites included in the current analysis are DAN-5, OGS-5, OGS-12 (all ca.
+1.6–1.5 Ma), and BSN-17 (1.3 Ma), with ages estimated by stratigraphic position in the Gona sequence [@quade2008; @semaw2020].
+The Busidima Formation accumulated through fluvial deposition by the ancestral Awash River (Type I context) and its smaller tributary channels [Type II context; @quade2004; @quade2008].
 Oldowan sites at Gona all occur in Type I sediments, indicating channel bank/margin (OGS-7) or proximal floodplain (EG-10, EG-12) contexts close to the large, hetero-lithic clasts available from point bars in the axial river channel [@quade2004; @stout2005].
-Acheulean sites continue to occur in Type I contexts (BSN-17, DAN-5) but are also found in Type II sediments (OGS-5, OGS-12,) reflecting increased utilization of large perennial tributaries to the ancestral Awash River [@quade2008].
+Acheulean sites continue to occur in Type I contexts (BSN-17, DAN-5) but are also found in Type II sediments (OGS-5, OGS-12) reflecting increased utilization of large perennial tributaries to the ancestral Awash River [@quade2008].
 Clasts locally available in these tributaries were relatively small, implying that the large flakes and cobbles used to produce Acheulean artifacts were initially sourced from the axial river.
 A similar pattern of habitat diversification and increasing lithic transport distances has been described at other sites and may be typical of the early Acheulean [@rogers1994; @linaresmatás2021; @hay1976].
-As with other early (i.e.
-\>1.0 mya [@stoutStoneToolmakingEvolution2011; @presnyakova2018]) Acheulean assemblages, the Gona collections include numerous "crudely made" handaxes and picks on large flake blanks and cobbles, as well as large (\> 10cm) unmodified flakes, flaked pieces interpreted typo-technologically as Mode 1 cores (see Figure 1af), and smaller debitage [@semaw2018; @semaw2020].
+As with other early [i.e., \> 1.0 mya; @stoutStoneToolmakingEvolution2011; @presnyakova2018] Acheulean assemblages, the Gona collections include numerous “crudely made” handaxes and picks on large flake blanks and cobbles, as well as large (\> 10 cm) unmodified flakes, flaked pieces interpreted typo-technologically as Mode 1 cores, and smaller debitage [@semaw2018; @semaw2020].
 
-<!-- ```{r fig2, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Gona site map."} -->
+# Materials and methods
 
-<!-- knitr::include_graphics(here::here("figure/site map.png")) -->
+## Archaeological sample
 
-<!-- ``` -->
+Artifact collections analyzed here include in situ pieces excavated from intact stratigraphic contexts and surface pieces systematically collected from the sediments eroding from these layers.
+Surface pieces are included because the current technological analysis does not require more precise spatial association, stratigraphic, and chronological control.
+Our sample comprises the total collection of flaked pieces [@isaac1997] from each site, regardless of size, as well as any large (\> 10 cm) unmodified detached pieces (total $n = 226$).
+As summarized in Table 1, our analytic categories thus include Flaked Cobbles, Flaked Flakes, Flaked Pieces on Indeterminate bases, and Unmodified Large Flakes.
+Note that the Flaked Piece categories are not size dependent, and a number of small Flaked Flakes from Oldowan sites are thus included.
+Our intention is to avoid a priori assumptions about artifact typology, technology, or function.
+All artifact collections from Gona are housed at the National Museum of Ethiopia.
 
-# Materials and Methods
+## Artifact classification and comparison
 
-## Materials
+Artifacts were classified according to initial form (pebble/cobble, detached piece, or indeterminate), presence/absence of flaking, technological interpretation (“Mode 1” vs.
+“Mode 2” FP), and archaeological context (Oldowan vs.
+early Acheulean sites).
+Mode 1 vs.
+Mode 2 classification of flaked pieces was based on qualitative assessment of LCT criteria reviewed in the Introduction (i.e.
+elongation, flattening, the creation of a point and, in the case of flakes, size).
+LCTs were additionally classified as handaxes, knives, or picks drawing on definitions from Kleindienst [-@kleindienst1962] and Leakey [-@leakey1971] as we understand them to be broadly employed in recent early Achuelean literature [e.g., @lepre2011; @beyene2013; @delatorre2018].
+In this scheme, handaxes include bifacially or unifacially flaked tools that are relatively flat and have a rough “teardrop” shape.
+Picks are thicker tools, often with trihedral or quadrangular sections and a clear point.
+Knives are lightly worked large flakes characterized by flaking mostly along one edge, often opposite the platform.
+The utility of classification to Mode and type was assessed through discriminant function analyses (DFAs) using artifact shape and reduction intensity variables.
+The aim of DFA analysis was to see if qualitative artifact types corresponded to analytically relevant patterning in our quantitative shape and reduction measures.
+Evidence for such patterning was on the basis of DFA classification success rates, visual assessment of clustering in DF plots, and using Hartigans’ dip test for unimodality/multimodality of DF score distributions.
+Comparisons between artifact classes (initial form, presence of modification, Mode, typology) and contexts (Oldowan vs.
+Acheulean) with conducted using t-tests to assess significance and Cohen’s d as a standardized measure of effect size.
 
-### Archaeological Sample
+## Artifact measurement
 
-Artifact collections analyzed here include *in situ* pieces excavated from intact stratigraphic contexts and surface pieces systematically collected from the sediments eroding from these layers.
-Surface pieces are included because the current technological analysis does not require more precise spatial association, stratigraphic, and chronological control.
-Our sample comprises the total collection of flaked pieces [@isaac1997] and large (\>10 cm) detached pieces from each site, regardless of typo-technological interpretation.
+Conventional linear measures capture the direction (e.g., length \> width) but not the location of geometric relations (e.g., position of maximum breadth).
+We address this by collecting linear measures defined by homologous semi-landmarks.
+All artifacts were oriented along their maximum dimension, without respect to any other technological or morphological landmarks, which was measured and defined as “length” (L).
+The next largest dimension orthogonal to length was used to define the plane of “width,” (W) with the dimension orthogonal to this plane defined as “thickness” (T).
+Additional width (W~1~, W~2~, W~3~) and thickness (T~1~, T~2~, T~3~) measures were then collected at 25%, 50%, and 75% of Length, with the piece oriented so that W~1~ \> W~3~.
+To partition variation in shape from variation in size, we employed geometric mean size-adjustment [@lycett2006].
+For each artifact, we divided each linear measure (L, W~1~, W~2~, W~3~, T~1~, T~2~, T~3~) by the geometric mean for that artifact (the seventh root of the product of these seven measurements).
+Geometric mean (GM)-adjusted variables were then submitted to a principal components analysis (PCA; covariance matrix) to identify the main dimensions of shape variation.
 
-## Methods
+To estimate surface area, we combined our measure of length (L) with maximum width (W~m~) and thickness (T~m~) measures in a prism-based formula (2LW~m~ + 2LT~m~ + 2 W~m~T~m~).
+Surface area calculated in this way correlates with mass^2/3^ at $r= 0.966$ in our sample.
+Calculated surface area was then used to derive the Scar Density Index: SDI = number of flake scars \> 1 cm per unit surface area [@clarkson2013; @shiptonFlakeScarDensity2015].
+The length (maximum dimension) of each flake scar was also measured.
+The Flaked Area Index [FAI: flaked area/total surface area; @li2015] was estimated directly “by eye” as a percentage (nearest 5%) of the total artifact surface.
 
-### Artifact Classification
+To assess the adequacy of shape descriptions based on our linear measures, we directly compared these with shape as quantified by 3D methods on the 31 artifacts for which scans are available.
+GM-transformed linear measures from these 31 artifacts were submitted to a variance-covariance matrix PCA.
+Principal components (PCs) with an eigenvalue greater than the mean were retained for analysis and the results compared qualitatively (morphological interpretation of PCs) and quantitatively (correlation of artifact factor scores) to 3D results.
+Stepwise regression was used to assess the amount of variance accounted for in cases where multiple 3DGM-PCs were found to be correlated with a single linear PC.
+Accuracy of surface area and flaked area estimates was assessed by correlation with 3D results.
 
-Artifacts were classified according to initial form (pebble/cobble, detached piece, or indeterminate), presence/absence of retouch, technological interpretation ("Mode 1" core vs.
-"Mode 2" LCT), and archaeological context (Oldowan vs.
-Early Acheulean sites).
-LCTs were additionally classified as handaxes, knives, or picks following definitions from Kleindienst [-@kleindienst1962].
-The validity of technological interpretations and typological classifications was assessed through cluster analysis based on artifact shape and reduction intensity variables.
+## Three-dimensional methods
 
-### Artifact Measurement
-
-Conventional linear measures capture the direction (e.g., length \> breadth) but not the location of geometric relations (e.g., position of maximum breadth).
-We address this by collecting linear measures defined by homologous semi-landmarks.
-All artifacts were oriented along their maximum dimension, which was measured and defined as "length".
-The next largest dimension orthogonal to length was used to define the plane of "breadth", with the dimension orthogonal to this plane defined as "thickness" Width (W~1~, W~2~, W~3~) and thickness (T~1~, T~2~, T~3~) measures were then collected at 25%, 50%, and 75% of length, oriented so that 25% Width \> 75% Width.
-To partition variation in shape from variation in size, we divided all linear measures by the geometric mean [@lycett2006].
-GM-transformed variables were then submitted to a Principal Components Analysis (covariance matrix) to identify the main dimensions of shape variation.
-
-Our semi-landmark measurement system allowed us to improve on the prism-based surface area formula (2LW + 2LT + 2 WT) by using our 7 recorded dimensions to more tightly fit three prisms around the artifact: SA = W~1~T~1~ + 2(.33L \* W~1~) + 2(.33L \* T~1~) + 2(.33L \* W~2~) + 2(.33L \* T~2~) + 2(.33L \* W~3~) + 2(.33L \* T~3~) + W~3~T~3~.
-Surface area calculated in this way correlates with mass^2/3^ at r^2^ = 0.947 in our sample.
-Calculated surface area was then used to derive the Scar Density Index: SDI = number of flake scars \> 1cm per unit surface area [@clarkson2013; @shiptonFlakeScarDensity2015].
-The Flaked Area Index (FAI: flaked area divided by total surface area)[@li2015] was estimated directly "by eye" as a percentage of the total artifact surface.
-
-To assess the adequacy of shape descriptions based on our linear measures, we directly compared these with shape as quantified by 3D methods on the 33 artifacts for which scans are available.
-GM-transformed linear measures from these 33 artifacts were submitted to a variance-covariance matrix PCA.
-PCs with an eigenvalue greater than the mean were retained for analysis and the results compared qualitatively (morphological interpretation of PCs) and quantitatively (correlation of artifact factor scores) to 3D results.
-Accuracy of surface area and flaked area estimates was also assessed by correlation with 3D results.
-
-### 3D Methods
-
-3DGM analysis was conducted using the *AGMT3-D* program of Herzlinger and Grosman [-@herzlinger2018].
+3DGM analysis was conducted using the AGMT3-D [@herzlinger2018].
 Artifacts were automatically oriented according to the axis of least asymmetry, then manually oriented following the same orthogonal conventions described in section 2.3.
-Then, a grid of 200 homologous semi-landmarks were overlain on each artifact's surface.
-Generalized Procrustes and Principal Component analyses were then undertaken to explore the shape variability of the sample.
-The surface area of each artifact was calculated using the *Artifact3-D* program of Grosman et al. [-@grosman2022].
-*Artifact3-D* was also used to automatically identify the flake scar boundaries and compute each scar's surface area, using the scar analysis functions of Richardson et al. [-@richardson2014].
+Then, a grid of 200 homologous semi-landmarks were overlain on each artifact’s surface.
+Generalized Procrustes and PCA were then undertaken to explore the shape variability of the sample.
+The surface area of each artifact was calculated using the Artifact3-D program [@grosman2022].
+Artifact3-D was also used to automatically identify the flake scar boundaries and compute each scar’s surface area, using the scar analysis functions of Richardson et al.
+[-@richardson2014].
+While flaked and unflaked areas were mostly easily identifiable on the 3D scans, scar counts and photographs taken during the original analysis were used to resolve any ambiguities in the presence or location of cortex, ventral surface, and scars.
+In cases where this automatic scar segmentation did not match the real scar boundaries, the Merge function of the program was used to manually join or separate the erroneous regions.
+These boundaries were used to calculate the total 3D surface area of each artefact, as well as its flaked and unflaked area.
 
@@
 
-## Measurement Validation
+## Measurement validation
 
-A PCA on GM transformed linear measures of the 33 artifacts identified two PCs accounting for 75.6% of the variance.
-Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance.
-Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness).
-This A PCA on GM transformed linear measures the 33 artifacts identified two PCs accounting for 75.6% of the variance.
-Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance.
-Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness).
+A PCA on GM transformed linear measures of the 31 scanned artifacts identified two PCs accounting for 71.5% of the variance.
+Linear-PC1 (Eigenvalue = 0.161) explained 49.7% of the variance.
+Factor loadings (**Table** 2) for Linear-PC1 reflect artifact elongation (i.e., a negative correlation of length vs.
+distal width and thickness).
+
 This closely parallels the length vs.
-width and thickness tradeoff captured by 3DGM-PC1 (**Figure** \@ref(fig:fig3d)) and is reflected in a tight correlation of artifact scores produced by the two PCs (r = 0.903, p \< 0.001, **Figure** \@ref(fig:figcomparison)A).
-A second factor (Linear-PC2, eigenvalue = 0.07) explained an additional 20.4% of variance.
-This factor was less strongly correlated with its 3DGM counterpart (3DGM-PC2; r = 0.344, p = 0.050) probably because Linear-PC2 describes anticorrelated variation in width and thickness (i.e., broad and flat vs.
-thick and pointed; **Table** \@ref(tab:tab1)) whereas 3DGM-PC2 more purely isolates convergence (**Figure** \@ref(fig:fig3d)).
-The remainder of the shape variability explained by Linear-PC2 is captured by higher order 3DGM-PCs 3 through 5, which comprise the contribution of the left and right lateral margins to relative thickness.
+width and thickness tradeoff captured by 3DGM-PC1 (**Fig.** \@ref(fig:fig3)) and is reflected in a tight correlation of artifact scores produced by the two PCs ($r = 0.890$, $p  < 0.001$; **Fig.** \@ref(fig:fig4)a).
+A second factor (Linear-PC2, eigenvalue = 0.07) explained an additional 21.8% of variance.
+This factor was less strongly correlated with its 3DGM counterpart (3DGM-PC2; $r = -0.458$, $p = 0.010$) probably because the remainder of the shape variability explained by Linear-PC2 is captured distributed across the higher order 3DGM-PCs 3 through 5 (Supplemental Fig.
+1)..
+These PCs describe localized variation in relative thickness (especially of the base, 3DGM-PC 3 & 5) and breath (especially toward the middle/tip, 3DGM-PC 4).
+Although these PCs partition variation on the right margin (3DGM-PC3) from the left (3DGM-PC 4 & 5), none appears straightforwardly related to artifact symmetry.
+These 3DGM descriptions are of course much more detailed than is possible with our 7 linear measures, but we consider them to be broadly summarized by linear-PC2, which describes variation ranging from broad tips and thin butts on one extreme to pointed tips and thick butts on the other.
+Together, 3DGM-PC2, 3DGM-PC3 ($r = -0.359$, $p = 0.047$), 3DGM-PC4 ($r = 0.381$, $p = 0.035$), and 3DGM-PC5 ($r = -0.584$, $p < 0.001$) all correlate with Linear-PC2, cumulatively capturing whether the items are broad and flat or thick and pointed.
+A stepwise regression ($r^2 = 0.824$, F(4,26) = 30.408, $p < 0.001$, Probability-of-F-to-enter $\leq$ 0.050; Probability-of-F-to-remove $\geq$ 0.100) with Linear-PC2 as the dependent variable retained all four of these 3DGM-PCs as significant predictors.
 Use of high-resolution, coordinate-based scan data thus generates PCs that identify more specific shape attributes, but the underlying morphological variability captured by the linear and 3D analyses remains similar.
-Together, 3DGM-PC2 (r = 0.344, p = 0.050), 3DGM-PC3 (r = -0.416, p = 0.016), 3DGM-PC4 (r = 0.458, p = 0.007), and 3DGM-PC5 (r = -0.352, p = 0.044) correlate well with Linear-PC2, cumulatively capturing whether the items are broad and flat or thick and pointed.
-A stepwise regression ($r^2$=0.625, F(4,28)=11.697, p\<0.001, Probability-of-F-to-enter \<= 0.050; Probability-of-F-to-remove \>= 0.100) with Linear-PC2 as the dependent variable retained all four of these 3DGM-PCs as significant predictors.
+In fact, for questions like our about basic variation in overall artifact shape, simple linear descriptions may be easier to work with, summarize, and interpret.
 
-```{r tab1, echo=FALSE, message=FALSE}
-library(tidyverse)
-df1 <- data.frame("Linear metrics (GM transformed)" = c("Length", "W1", "W2", "W3", "T1", "T2", "T3"), 
-                 "Linear-PC1" = c("0.989", "0.303", "0.403", "-0.176", "-0.135", "-0.369", "-0.607"), 
-                 "Linear-PC2" = c("-0.107", "0.350", "0.767", "0.790", "-0.679", "-0.623", "-0.282") ) 
-kableExtra::kbl(df1, booktabs = T, caption = "Component loadings for linear metric PCs on scanned sample.") 
-```
-
-```{r fig3d, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="3D models displaying the first two PCs."}
-knitr::include_graphics(here::here("figure/Fig3a.png"))
-```
-
-```{r figcomparison, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Comparison between linear measurement and 3d measurement."}
-knitr::include_graphics(here::here("figure/Fig.3D comparison new1.png"))
+```{r fig3, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Principal Components 1 and 2 from the 3DGM analysis."}
+knitr::include_graphics(here::here("figure/Fig.3.png"))
 ```
@@
 We thus concluded that linear measures are adequate to capture relevant shape variation and proceeded with a PCA on our full sample.
-We identified two PCs accounting for 80.0% of the variance.
-PC1 (Eigenvalue = 0.216) explained 56.4% of the variance.
-Factor loadings (**Table** \@ref(tab:tab2)) for PC1 reflect artifact flatness (i.e., an anti-correlation of length and width vs. thickness) such that higher values indicate relatively thinner pieces.
-PC2 (Eigenvalue = 0.090) explained 23.6% of the variance.
-Factor loadings (**Table** \@ref(tab:tab2)) show that PC2 captures artifact convergence or "pointedness" (i.e., an anti-correlation of tip width with length and butt thickness) such that higher values indicate shorter, less pointed forms.
+This identified two PCs accounting for 79.7% of the variance.
+PC1 (Eigenvalue = 0.214) explained 55.9% of the variance.
+Factor loadings (**Table** 3) for PC1 reflect artifact flatness (i.e., an negative correlation of length and width vs.
+thickness), such that higher values indicate relatively thinner pieces.
+PC2 (Eigenvalue = 0.091) explained 23.8% of the variance.
+Factor loadings (**Table** 3) show that PC2 captures artifact pointedness (i.e., a negative correlation of tip width with length and butt thickness) such that higher values indicate shorter, less pointed forms.
 
-```{r tab2, echo=FALSE, message=FALSE}
-library(tidyverse)
-df1 <- data.frame("Linear metrics (GM transformed)" = c("Length", "W1", "W2", "W3", "T1", "T2", "T3"), 
-                 "PC1" = c("0.905", "0.635", "0.680", "0.486", "-0.510", "-0.684", "-0.719"), 
-                 "PC2" = c("-0.414", "0.299", "0.571", "0.761", "-0.526", "-0.525", "0.023") ) 
-kableExtra::kbl(df1, booktabs = T, caption = "Component loadings for linear metric PCs in the full sample.") 
+```{r fig4, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Comparison of 3D shape (A), scar density (B), and flaked area (C) measures with corresponding linear measurements and estimates.
+Abbreviations: PC = principal component; 3DGM = three-dimensional geometric morphometrics; SDI: corrected scar density index."}
+knitr::include_graphics(here::here("figure/Fig.4.png"))
 ```
@@
 We also tested the validity of our two reduction measures, SDI and FAI.
-In agreement with Clarkson (2013), we found that surface area estimated from caliper measures displayed a strong correlation with (r2=0.975, p \< 0.001) but linear over-estimation of ($\beta$ = 1.58) 3D surface area.
+In agreement with Clarkson [-@clarkson2013], we found that surface area estimated from caliper measures displayed a strong correlation with ($r^2 = 0.982$, $p < 0.001$) but linear over-estimation of ($\beta = 0.452$) 3DGM surface area.
 This results in a systematic underestimation of SDI that scales with core size.
-However, a simple correction of the caliper estimate (dividing by the slope, 1.58) eliminates surface area over-estimation (mean difference = $256mm^2$ [\<1.7% of mean], p=0.040) and produces SDI values that agree with 3D values (r2=0.975, p \< 0.001, $\beta$ = 0.98) (**Figure** \@ref(fig:figcomparison)B).
+However, a simple correction of the caliper estimate (multiplying by 0.5) eliminates surface area over-estimation and produces SDI values that agree with 3D values (\$r$2 = 0.988$, $p < 0.001$, $\beta = 0.910$; **Fig.** \@ref(fig:fig4)b).
 We thus proceeded to apply this correction to surface area estimates in the full sample.
-Insofar as these relationships are driven by basic geometry, we expect these methods (including correction) to be generalizable to other ESA assemblages.
+Insofar as these relationships are driven by basic geometry (roughly 2:1 ratio between the surface area of a rectangular prism vs.
+ellipsoid), we expect this correction to be generalizable to other Early Stone Age (ESA) assemblages.
 
-Visual estimation of flaked area approximated 3DGM measurement very closely (r2=0.932, p \< 0.001, $\beta$ = 1.051; **Figure** \@ref(fig:figcomparison)C) and without any systematic bias (paired t-test: mean difference = -0.015%, 95% CI = 2.02% to 1.99%, p = 0.987).
-Individual errors ranged between -10.46% and 14.24%.
+Visual estimation of flaked area approximated 3D measurement very closely ($r^2 = 0.929$; $p < 0.001$, $\beta = 0.868$; **Fig.** \@ref(fig:fig4)c) and without any systematic bias (paired t-test: mean difference = -0.24%, 95% CI = -2.35% to 1.86%, $p = 0.811$).
+Individual errors ranged between extremes of -10.46% and 14.24%.
 We thus considered visual estimation to be reliable in our sample.
@@
 
-We first conducted a stepwise DFA on all flaked pieces (n=192) with inferred technological Mode (one vs.
-two) as the grouping variable and PCs 1 and 2, corrected SDI (cSDI), and FAI as the independent variables.
-All variables were retained, yielding one canonical DF (eigenvalue=1.825, Wilks Lambda = 0.354, p\< 0.001) which correctly classified 93.8% of artifacts.
-We thus accepted the validity of classification by Mode in our sample and employed this distinction in subsequent analyses.
-There was no discernable difference in discriminant scores for Mode 1 cores from Oldowan (n=37) vs. Acheulean (n=39) contexts (, p = 0.746).
-Mode 1 cores from Oldowan contexts (n=37) do include 10 (27%) small, retouched flakes.
-Only one retouched flake from an Acheulean context was classified by the DFA as a Mode 1 core.
-This piece, typologically classified as a "knife", is the smallest (93mm) retouched flake in the Acheulean sample.
-When retouched flakes are excluded from the comparison, there are no significant differences in shape between Mode 1 cores from Oldowan and Acheulean contexts.
-Interestingly, however, Acheulean Mode 1 cobble-cores are much larger (mean weight 480g vs.
-186g, Cohen's d=1.137, p\<0.001) and less heavily reduced (mean cSDI 0.057 vs.
-0.103, Cohen's d = -0.884, p \< 0.001) despite having similar FAI (mean 0.52 vs.
-0.46, Cohen's d=0.271, p=0.266).
+We first conducted a stepwise DFA on all flaked pieces ($n = 191$) with inferred technological Mode (one vs.
+two) as the grouping variable and shape PCs 1 and 2, corrected SDI (cSDI), and FAI as the independent variables.
+All variables were retained, yielding one canonical discriminant function DF (eigenvalue = 2.152, Wilks Lambda = 0.317, $p < 0.001$) which correctly classified 95.3% of artifacts.
+We thus concluded that Mode captures relevant patterning in our quantitative data and employed this distinction in subsequent analyses.
+There was no discernable difference in discriminant scores for Mode 1 flaked pieces from Oldowan ($n = 37$) vs.
+Acheulean ($n = 36$) contexts ($p = 0.703$).
+Mode 1 FPs from Oldowan contexts include 7 (19%) small, flaked flakes, three of which were misclassified by the DFA as Mode 2.
+When flaked flakes are excluded from the comparison, there are no significant differences in shape or DFA scores between Mode 1 flaked pieces from Oldowan vs.
+Acheulean contexts.
+Interestingly, however, Acheulean Mode 1 flaked cobbles are much larger (mean weight 481 g vs.
+186 g, Cohen’s d = 1.137, $p < 0.001$) and less heavily reduced (mean cSDI 0.06 vs.
+0.10, Cohen’s d = -0.949, $p < 0.001$) despite having similar FAI (mean 0.52 vs.
+0.46, Cohen’s d = 0.271, $p = 0.266$).
 
-Next, we conducted a stepwise DFA on all flaked Mode 2 pieces (i.e.
-excluding unmodified large flakes, n = 115) with typology (handaxe, pick, knife) as the grouping variable and the same four independent variables.
-Both shape PCs and FAI were retained, while cSDI was not entered.
-This produced two DFs (DF1: Eigenvalue=1.536, 91.3% of variance; DF2: Eigenvalue = 0.146, 8.7% of variance; Wilks Lambda = 0.344; p\<0.001) which correctly classified 71.3% of artifacts.
-Inspection of DF coefficients shows that DF1 captures an inverse correlation between flaked area and pointedness (Linear-PC2) whereas DF2 captures a positive relationship between flaked area and elongation (Linear-PC2).
-A bivariate plot (Figure 5) illustrates the fact that DF1 captures a range of variation from pointed, heavily-worked picks, through handaxes, to knives, with substantial overlap between adjacent types (Table 2).
-As an intermediate type, "handaxes" were correctly classified only 43.9% of the time.
-In agreement with others [@duke2021; @kuhn2020; @presnyakova2018] we conclude that these typological labels artificially partition a continuum of variation and abandon them in subsequent analyses.
+Next, we conducted a stepwise DFA on all flaked Mode 2 pieces (i.e., excluding unmodified large flakes, $n = 118$) with typology (handaxe, pick, knife) as the grouping variable and the same four independent variables.
+All four variables were again retained, producing two DFs (DF1: Eigenvalue = 1.02, 86.3% of variance; DF2: Eigenvalue = 0.240, 13.8% of variance; Wilks Lambda = 0.322; $p < 0.001$) which correctly classified 73.7% of artifacts.
+A plot of artifact DF values (**Fig.** \@ref(fig:fig5)a) shows substantial overlap between types, with handaxes occupying an intermediate position between picks and knives.
+As such an intermediate type, handaxes were correctly classified by only 53.7% of the time (Table 4).
+Visually, the DF1 distribution appears weakly bimodal with the handaxe centroid falling between two peaks, although the distribution is not actually significantly non-unimodal (Hartigans’ dip test for unimodality/multimodality, D = 0.0433, $p = 0.2073$).
+Inspection of DF coefficients shows that DF1 captures a correlation between FAI and Linear-PC2 such that increasing FAI is associated with greater pointedness.
+The distribution of DF1 values thus reflects a distinction between pointed, heavily worked “picks” and less pointed, more lightly worked “knives.” This, in turn largely corresponds to the different morphology of flaked cobbles vs.
+flaked flakes (**Fig.** \@ref(fig:fig5)b).
+We thus find no evidence that traditional Achuelean LCT types are a useful way to organize our data, and we abandon them in subsequent analyses, focusing instead on reduction Mode and initial blank form.
+
+```{r fig5, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="DFA results."}
+knitr::include_graphics(here::here("figure/Fig.5.png"))
+```
 
@@
 
-To assess the influence of flake removals on core form, we examined the association between our reduction measures (cSDI and FAI) and core shape (PC1, PC2).
-In the complete sample of flaked pieces (n=192), we observed weak but significant effects of cSDI (r=-0.294, p \<0.001) and FAI (r=-0.294, p\<0.001) on PC1(flatness) and of FAI only on PC2(pointedness) (r=-0.436, p\<0.001).
-However, it is clear that these overall effects conflate different trends in Mode 1 vs. Mode 2 cores, as well as in cores executed on flake vs. pebble/cobble bases.
-Within categories, we observed no significant effects of cSDI whereas FAI had variable relationships with core form.
+To assess the influence of flake removals on Flaked Piece (FP) form, we examined the association between our reduction measures (cSDI and FAI) and FP shape (PC1, PC2).
+Summary statistics are presented in Table 5.
+In the complete sample of flaked pieces ($n = 191$), we observed weak but significant effects of cSDI ($r = -0.361$, $p < 0.001$) and FAI ($r = -0.294$, $p < 0.001$) on linear-PC1(flatness) and of FAI only on linear-PC2 (pointedness; $r = -0.443$, $p < 0.001$).
+However, these overall effects conflate different trends in Mode 1 vs.
+Mode 2 FPs, as well as in FPs executed on flake vs.
+pebble/cobble bases.
+Within categories, we observed no significant effects of cSDI whereas FAI had variable relationships with FP form.
 
-In Mode 1 cores, a weak negative effect of FAI on PC2(pointedness) (n = 76, $\beta$ =-0.006, r=0.240, p= 0.037) suggests that increasing extent of modification tends to slightly increase pointedness (Figure 6a).
-No effects of FAI on PC1 (flatness) or of SDIc on either PC approach significance.
-FAI and SDIc are moderately correlated ($\beta$=0.001, r=0.524, p\<0.001), an effect that reflects increases in the upper limit of scar density values as flaked area increases (i.e., flaked area constrains maximum cSDI: Figure 6b).
-
-In Mode 2 cores (n=115), reduction measures have different effects depending on initial blank form.
-In general, the contrast between flat, lightly-worked flake bases and rounder, more-heavily worked cobble bases tends to inflate relationships between FAI and core shape (**Figure** \@ref(fig:FAIBASE)).
-Mode 2 cobble cores (n=37) display a negative effect of FAI on PC1(Flatness) ($\beta$=-0.011, r=0.367, p = 0.026), no effect of FAI on pointedness, and no association between FAI and cSDI (r=0.141, p = 0.406).
-Flake cores (n=69) show negative effects of FAI on PC1(Flatness) ($\beta$=-0.007, r=0.284, p = 0.018) and, more strongly, PC2(pointedness) ($\beta$=-0.019, r=0.443, p \< 0.001).
-Comparison of Mode 2 flake cores with unmodified large flakes from Acheulean contexts (n=35) shows that ranges of shape variation substantially overlap but that flaking generally reduces both PC1 and PC2.
-In fact, regressions of PC1 and PC2 on FAI show y-intercepts that closely approximate unmodified flake mean values (0.852 vs.
-0.774 and 0.693 vs.
-0.594, respectively) and significantly negative slopes as FAI increases (**Figure** \@ref(fig:FAIFLAKED)).
-Flake cores show a small but significant positive effect of FAI on cSDI that, in contrast to Mode 1 cores, represents actual covariance rather than simply a relaxation of constraint (i.e.
-both upper and lower limits of cSDI increase with FAI).
-
-```{r FAIBASE, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Mode 2 trends across blank types."}
-knitr::include_graphics(here::here("figure/Fig.FAI by base new NEW.png"))
+```{r fig6, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Effect of FAI on Mode 2 flaked piece shape across cobble, flake, and indeterminate bases.
+Abbreviations: PC = principal component; FAI: Flaked area index."}
+knitr::include_graphics(here::here("figure/Fig.6.png"))
 ```
 
-```{r FAIFLAKED, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Effect of FAI compared to unmodified large flakes from Acheulean contexts."}
-knitr::include_graphics(here::here("figure/Fig.FAI by flaked new123.png"))
+In Mode 1 FPs, a weak negative effect of FAI on PC2 ($n = 73$, $\beta = -0.007$, $r = 0.271$, $p = 0.020$) suggests that increasing extent of modification tends to slightly increase pointedness.
+No effects of FAI on PC1 or of cSDI on either PC approach significance.
+FAI and cSDI are moderate-to-weakly correlated ($\beta = 0.001$, $r = 0.446$, $p < 0.001$).
+
+In Mode 2 FPs ($n = 118$), reduction measures have different effects depending on initial blank type.
+In the complete sample of Mode 2 FPs (**Fig.** \@ref(fig:fig6)), pieces range from broad and flat, lightly worked flake bases to rounder, more pointed, and more heavily worked cobble bases.
+However, effects of reduction within base categories are less consistent.
+
+Mode 2 flaked cobbles ($n = 37$) display a weak negative effect of FAI on PC1(flatness; $\beta = -0.011$, $r = 0.368$, $p = 0.025$), no effect of FAI on PC2(pointedness), and no association between FAI and cSDI ($r = 0.114$, $p = 0.502$), thus showing minimal evidence of shaping.
+Flaked flakes (n = 72) show negative effects of FAI on PC1 ($\beta = -0.008$, $r = 0.289$, $p = 0.014$) and, more strongly, PC2 ($\beta = -0.018$, $r = 0.425$, $p < 0.001$) as well as a moderate correlation between FAI and cSDI ($r = 0.570$, $p < 0.001$).
+Thus, increasing flaked area is associated with increasingly thicker and more pointed forms and there is a tendency for additional flake scars to increase flaked area.
+Comparison of Mode 2 flaked flakes with unmodified large flakes from Acheulean contexts ($n = 35$) shows that ranges of shape variation substantially overlap but that flaking generally reduces both PC1 and PC2.
+In fact, regressions of PC1 and PC2 on FAI show y-intercepts that closely approximate unmodified flake mean values (0.839 vs.
+0.780 and 0.679 vs.
+0.591, respectively) and significantly negative slopes as FAI increases (**Fig.** \@ref(fig:fig7)).
+
+```{r fig7, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Effect of FAI on large flake shape.
+Silhouettes depict the GM-transformed Thickness (A) and Breadth (B) values of the extreme cases labeled in the plots.
+Abbreviations: PC = principal component; FAI: Flaked area index."}
+knitr::include_graphics(here::here("figure/Fig.7.png"))
 ```
 
-A small number of Mode 2 cores (n=9) were so heavily modified (mean FAI=89%, range 65-100) that the blank form could not be determined.
-These heavily worked "indeterminate" cores tend to show similar PC1 (flatness) and PC2 (pointedness) values to cobble-base cores (i.e.
-thicker and more pointed than flake cores) and follow general trends of shape change with increasing FAI observed across base types (**Figure** \@ref(fig:FAIBASE)).
+A small number of Mode 2 FPs ($n = 9$) were so heavily modified (mean FAI = 89%, range 65-100) that the blank form could not be determined.
+These heavily worked “indeterminate” cores tend to show similar PC1 (flatness) and PC2 (pointedness) values to cobble-base cores (i.e., thicker and more pointed than flake cores) and to follow general trends of shape change with increasing FAI observed across base types (**Fig.** \@ref(fig:fig6)).
+
+## Cobble selection
+
+Although there is little evidence of intentional shaping of cobbles at Gona, it is also clear that Mode 2 flaked cobbles are more elongated than Mode 1 flaked cobbles and represent an extreme with the range of cobble shapes and sizes found in Type 1 channels in the Busidima Formation [@quade2004].
+This is not simply an analytical artifact of classifying elongated flaked cobbles as “Mode 2” while selectively omitting rounded forms: the complete sample of large ($\geq$ 750 g) cores in our sample ($n = 25$) is dominated by relatively elongated forms (median L/W = 1.66).
+The small number ($n = 5$) of large, rounded (L/W \< 1.33) cores in the sample are also anomalous in being very lightly worked compared to more elongated large FPs (mean scar count = 4.2 vs.
+10.45, $p < 0.001$; mean cSDI = 0.013 vs.
+0.029, $p < 0.001$; mean FAI = 36 vs.
+65, $p = 0.016$).
+Thus, there is strong evidence for the intentional selection of large, elongated cobbles at Gona early Acheulean sites.
 
@@
 
-Our analyses of flaked pieces from Oldowan and Early Acheulean contexts at Gona support the presence of two distinct reduction modes (1 and 2) that can be reliably discriminated using measures of artifact shape, flaking intensity, and extent of surface modification.
-However, they fail to support further sub-division between conventional Mode 2 tool types (handaxe, pick, knife).
-Following criteria proposed in the Introduction, we find strong evidence of imposed form in the retouched large flakes from Acheulean contexts at Gona, weak evidence of shaping in Mode 2 cobble cores from Acheulean contexts, and little to no evidence of shaping in Mode 1 cores from both Acheulean and Oldowan contexts.
+Our analyses of flaked pieces from Oldowan and early Acheulean contexts at Gona support the presence of two distinct reduction Modes (1 and 2) that can be reliably discriminated using measures of artifact shape, flaking intensity, and extent of surface modification.
+However, they fail to justify further sub-division between conventional Mode 2 tool types (handaxe, pick, knife).
+Rather shape and reduction variation in our sample seems more closely related to initial blank form (flake vs.
+cobble).
+These results are consistent with the view [@duke2021; @kuhn2020; @presnyakova2018] that typological labels arbitrarily partition a continuum of artifact variation, although it remains to be seen if other variables (e.g.
+location of flake removals) would reveal more discrete clustering.
+For current purposes, we focused on analysis by flaking Mode and initial form.
+
+Following criteria proposed in the Introduction, we find strong evidence of imposed form in the large flaked flakes from Acheulean contexts at Gona, weak evidence of shaping in Mode 2 flaked cobbles from Acheulean contexts, and little to no evidence of shaping in Mode 1 cores from both Acheulean and Oldowan contexts.
 Our framing of this research question and its proposed test criteria follows the preponderance of archaeological literature in conceptualizing artifact design as a binary presence/absence of intentionally imposed form [@dibbleMajorFallaciesSurrounding2017].
@@
 As we argued in the Introduction, the systematic production of large flakes indicates attention to artifact size as a desired design feature.
-Many of these flakes would have possessed substantial lengths of sharp edge and could have been immediately useful as cutting tools without further retouch.
-In fact, 33% of the large flakes in our total sample are unretouched.
-Notably, this is also true of the sub-sample from sites (OGS-5 and OGS-12) located in Type II, "tributary" sediments (29 of 82 unmodified, 35%).
+Many of these flakes would have possessed substantial lengths of sharp edge and could have been immediately useful as cutting tools without further modification.
+In fact, 33% of the large flakes in our total sample are unmodified.
+Notably, this is also true of the sub-sample from sites (OGS-5 and OGS-12) located in Type II, “tributary” sediments (29 of 82 unmodified, 35%).
 This likely reflects transport and discard of unmodified flakes from production sites closer to the axial river system sources of large clasts, consistent with the idea that such flakes were themselves treated as useful tools.
-We consider the alternative possibility, that unmodified flakes were unintended byproducts of local LCT production from transported boulder cores, to be less likely considering the energetic inefficiency of this strategy as well as the absence of large boulder cores in the assemblages (but see discussion of LCTs on cobble cores below).
+We consider the alternative possibility, that unmodified flakes were unintended byproducts of local LCT production from transported boulder cores, to be less likely considering the energetic inefficiency of this strategy, as well as the absence of large boulder cores in the assemblages (but see discussion of LCTs on cobbles below).
 
-Whether or not unmodified flakes were themselves desired "end products", a majority of large flakes in the sample were retouched to some extent.
+Whether or not unmodified flakes were themselves desired “end products”, a majority of large flakes in the sample were flaked to some extent.
 Our analyses indicate that this flaking was performed in a manner that produced systematic and directional shape changes.
-Specifically, increasing FAI is associated with progressive reduction in relative breadth (increasing relative thickness and elongation; PC1) and increasing convergence (PC2).
+Specifically, increasing FAI is associated with progressive reduction in relative breadth (i.e., increasing relative thickness and elongation; PC1) and increasing pointedness (PC2).
 In contrast, variation in cSDI is not associated with artifact shape.
@@
 This clearly differs from classic re-sharpening models, which emphasize shape change with decreasing artifact size as a byproduct of concentrated and repeated mass removal from particular areas of the artifact, such a working edges [@dibble1987; @mcpherron2003].
-It also clearly differs from the expected effects of un-patterned debitage on flake blanks, which does tend to increase relative thickness but also decreases elongation and is not associated with convergence [@moore2016].
+It also clearly differs from the expected effects of un-patterned debitage on flake blanks, which does tend to increase relative thickness but also decreases elongation and is not associated with pointedness [@moore2016].
 
-It is possible that the observed pattern reflects the intentional pursuit of an explicit morphological target or "mental template." This is generally seen as a strong claim implying rich mental representation and manipulation of spatial features [@wynn2002], cultural mechanisms for the intersubjective sharing of ideas of "appropriate" form [@wynn1995], and social reproduction of the skills required to reliably achieve these forms [@liuDifferentialEffectsKnapping2023].
-If there is such a target, however, it would seem to be a very loose one considering the observed range of shape variation in retouched flakes (Figure 5).
-A more cognitively and culturally "lean" interpretation is that the pattern reflects a procedural and/or functional bias toward establishing (or rejuvenating original) cutting edges at one end of the long axis.
-In the absence of effective thinning and volume management techniques, such a bias would naturally lead to narrowing/thickening of the piece overall and especially at the "tip," as is seen in our sample.
-This bias might itself be a socially learned behavioral convention or "habit" (sensu Isaac, 1986), but could plausibly emerge directly from individual responses to the morphology of large flake blanks.
-Such flakes are typically elongated and thus present: 1) two longer edges along the maximum dimension that are a natural focus for retouch, and 2) an ergonomic polarity defined by a thick (platform and percussion bulb) "butt" vs.
-a feather terminated "tip" [cf.
+It is possible that the observed pattern reflects the intentional pursuit of an explicit morphological target or “mental template.” This is generally seen as a strong claim implying rich mental representation and manipulation of spatial features [@wynn2002], cultural mechanisms for the intersubjective sharing of ideas of “appropriate” form [@wynn1995], and social reproduction of the skills required to reliably achieve these forms [@liuDifferentialEffectsKnapping2023].
+A more cognitively and culturally “lean” alternative is that the pattern reflects a procedural and/or functional preference to establish (or rejuvenate) cutting edges at one end of the long axis.
+In the absence of effective thinning and volume management techniques, such a preference would naturally lead to narrowing/thickening of the piece overall and especially at the “tip,” as is seen in our sample.
+This preference might itself be a socially learned convention or “habit” [sensu @isaac1986], but it could also emerge from individual assessments of the technological affordances of large flake blanks.
+Such flakes are typically elongated and thus present: 1) two longer edges along the maximum dimension that are a natural focus for edge imposition, and 2) an ergonomic polarity defined by a thick (platform and percussion bulb) “butt” vs.
+a feather terminated “tip” [cf.
 @key2016; @wynn2018].
-Even with this minimal interpretation, however, we would argue that LFB flaking strategies at Gona present evidence of design in the sense of being intended to produce and/or maintain desired functional features of the LCT.
+Even on this minimal interpretation, however, we would argue that LFB flaking strategies at Gona present evidence of design in the sense of being intended to produce and/or maintain desired functional features of the LCT.
 This contrasts with patterns observed in cores classified as Mode 1 in our analyses which, as expected, appear to reflect simple debitage without intended morphological targets.
@@
 
-A Discriminant Function Analysis strongly supported the presence of two distinct morphological types in our sample (based on overall shape and flake scar characteristics), which we had previously typo-technologically interpreted as reflecting Mode 1 (debitage) and 2 (façonnage) flaking strategies.
-Consistent with this interpretation, cores assigned to Mode 1 showed little or no evidence of core shape management.
-One potential exception is an unexpected tendency of increasing FAI to be associated with increasing convergence.
-However, this effect is an order of magnitude weaker than that seen in LFB Acheulean cores and may be an artifact of our orienting protocol, which aligns cores such that width at 25% length is always less than width at 75% length.
+A Discriminant Function Analysis strongly supported the presence of two distinct morphological types in our sample (based on overall shape and flake scar characteristics), which we had previously typo-technologically interpreted as reflecting Mode 1 (débitage) and 2 (façonnage) flaking strategies.
+Consistent with this interpretation, flaked pieces assigned to Mode 1 showed little or no evidence of shaping .
+One potential exception is an unexpected tendency of increasing FAI to be associated with increasing pointedness.
+However, this effect is an order of magnitude weaker than that seen in large Acheulean flaked flakes and may be an artifact of our orienting protocol, which aligns cores such that width at 25% length is always less than width at 75% length.
 If reduction tends to increase asymmetry from initially rounded cobble forms, this could account for the weak convergence effect we observe.
-In support of this interpretation, Mode 1 cores display a relatively strong correlation between FAI and cSDI driven by an increase in the upper bound of cSDI values as FAI increases.
-In other words, we find that greater flaked area on Mode 1 cores can, but does not necessarily, accommodate a greater number of flake scars.
-This is what would be expected from knapping unconstrained by any systematic core-shaping strategy.
 
-The presence (n=14) of relatively large (\> 100mm) Mode 1 cores at Gona Acheulean sites raises the further possibility that at least some of these might be depleted cores from LFB production.
-However, none of these display scars \> 90mm (mean = 51.8 mm), most are very lightly reduced (mean = 5.7 scars), and none meet the 150mm length cut-off used by Sharon (2009) to identify "large" cores suitable for LFB production.
-The largest Mode I core in our sample is 137 mm long, has five scars ranging from 25- 85 mm long, and an FAI of 25%.
-We thus consider it unlikely that large Mode I cores from Acheulean contexts were used for LFB production.
-The behavioral significance of relatively large, lightly-worked Mode 1 cores in Acheulean contexts is a question for future research.
+The presence ($n = 14$) of relatively large (\> 100 mm) Mode 1 flaked cobbles at Gona Acheulean sites raises the further possibility that at least some of these might be depleted cores from LFB production.
+However, none display scars \> 90 mm (mean = 51.8 mm), most are very lightly reduced (mean = 5.7 scars), and none meet the 150 mm length cut-off used by Sharon [-@sharon2009] to identify “large” cores suitable for LFB production.
+The largest Mode I flaked cobble in our sample is 137mm long, has five scars ranging from 25–85 mm long, and an FAI of 25%.
+We thus consider it unlikely that large Mode I flaked cobbles from Acheulean contexts were used for LFB production.
+The behavioral significance of relatively large, lightly-worked Mode 1 flaked pieces in Acheulean contexts is a question for future research.
 
@@
 
-Our most complicated results come from Mode 2 cobble cores.
+Our most complicated results come from Mode 2 flaked cobbles.
 Evidence of intentional shaping is limited: FAI has a moderate effect on PC1 such that increasing flaked area produces relatively shorter, thicker, and narrower pieces.
-This violates expectations that intentional LCT shaping should increase elongation and convergence.
-However, it also fails to align with a reported tendency for un-patterned debitage to decrease the thickness and increase the elongation of cobble cores (Moore & Perston, 2016).
-This may reflect the fact that Mode 2 cobble cores at Gona begin and remain relatively elongated (Table 3: mean L/W = 1.68) throughout reduction compared to starting (n=29, mean L/W = 1.29) and ending (mean L/W = 1.34) cobble cores in the experiment of Moore and Perston [-@moore2016].
-The moderate trend toward thickening and shortening of cores with increasing FAI might plausibly be an effect of particularly elongated initial core form.
-In the absence of intentional thinning techniques [@stout2014] flaking around the perimeter of an oblong core would be expected to preferentially reduce length and especially breath (due to the long axis providing more potential platforms, especially if knappers seek to maximize flake size) while preserving thickness.
-Along these lines, Toth [-@toth1985] reported that "roller" shaped cobbles tended to produce bifacial choppers in his unstructured debitage experiments.
+This violates expectations that intentional LCT shaping should increase elongation and pointedness.
+However, it also fails to align with a reported tendency for un-patterned debitage to decrease the thickness and increase the elongation of cobble cores [@moore2016].
+This may reflect the fact that Mode 2 flaked cobbles at Gona begin and remain relatively elongated (Table 4: mean L/W = 1.68) throughout reduction compared to starting ($n = 29$, mean L/W = 1.29) and ending (mean L/W = 1.34) cobble cores in the experiment of Moore and Perston [-@moore2016].
+The moderate trend toward thickening and shortening of flaked cobbles with increasing FAI might plausibly be an effect of particularly elongated initial cobble form.
+In the absence of intentional thinning techniques [@stout2014], flaking around the perimeter of an oblong cobble would be expected to preferentially reduce length and especially breath (due to the long axis providing more potential platforms, especially if knappers seek to maximize flake size) while preserving thickness.
+Along these lines, Toth [-@toth1985] reported that “roller” shaped cobbles tended to produce bifacial choppers in his unstructured debitage experiments.
 
-Whether or not shaping occurred, it is clear that the elongated Mode 2 cobble cores at Gona do provide evidence of initial size and shape selection within the range of rounded cobbles available from Type 1 channels in the Busidima Formation [@quade2004].
-This pattern is not simply an analytical artifact of classifying elongated cores as "Mode 2" while selectively omitting rounded forms: the complete sample of large (\>=750g) cores in our sample (n=25) is dominated by relatively elongated forms (median L/W = 1.66).
-The small number (n=5) of large, rounded (L/W \< 1.33) cores in the sample are also anomalous in being very lightly worked compared to more elongated large cores (mean scar count = 4.2 vs.
-10.45, p \< 0.001; mean cSDI = 0.14 vs.
-0.31, p= 0.002; mean FAI = 36 vs.
-65, p = 0.016).
-
-This could reflect the intentional selection of large, elongated blanks for LCT production, as has been proposed at other early Acheulean sites [@duke2021; @harmand2009; @texier2018].
+Whether or not shaping occurred, it is clear that the elongated Mode 2 flaked cobbles at Gona do provide evidence of the selection of relatively large and elongated cobbles from within the range of rounded cobbles available from Type 1 channels in the Busidima Formation [@quade2004].
+This could reflect the intentional selection of large, elongated blanks suitable for LCT production, as has been proposed at other early Acheulean sites [@harmand2009; @texier2018; @duke2021].
 We would characterize such selection as a design choice even in the absence of subsequent shaping.
-However, large-cobble shape preferences might equally reflect biases related to LFB production.
+However, large-cobble shape preferences might equally reflect preferences related to efficient LFB production.
 Geometrically, elongated cobbles afford greater potential maximum dimensions for detached flakes than do rounder cobbles of a similar mass.
@@
 @wynn2018] cutting edges than do round flakes.
-Flatter and more elongated cobbles are also generally easier to open and exploit [@texier2018; @toth1982; @whittaker1994].
+Flatter and more elongated cobbles are also generally easier to open and exploit [@toth1982; @whittaker1994; @texier2018].
 This is an increasingly important consideration as size increases due to the increased percussive force needed to detach larger flakes and the challenges of supporting and positioning larger cores [@semaw2009], particularly with cobbles that are large but not sufficiently massive to act as stationary targets for thrown hammerstones [@toth2019].
-We thus find it plausible that at least some typological picks, handaxes, and knives made on cobble cores at Gona could have been by-products of large flake production rather than intended forms.
-Toth and Schick [-@toth2019: 741] similarly note that "heavily reduced boulder cores can assume the form of smallish discoids and polyhedrons [and] may not be identified as sources of large flake blanks by many archaeologists".
+We thus find it plausible that at least some typological picks, handaxes, and knives made on cobbles at Gona could have been by-products of large flake production rather than intended forms.
+Toth and Schick [@toth2019: 741] similarly note that “heavily reduced boulder cores can assume the form of smallish discoids and polyhedrons [and] may not be identified as sources of large flake blanks by many archaeologists”.
 If this is true, such cores are distinguished from classic Mode 1 cores by allometric effects on raw material choice and flake placement rather than a shift to deliberate shaping.
 
-The possibility that some Mode 2 cobble cores at Gona might be the remains of cores used for large flake production was initially suggested by informal, "experiential" knapping with Gona cobbles, during which attempts at LFB production produced pick-like forms as a byproduct.
-Consistent with this hypothesis, the mean (152.9mm) and median (156mm) length of Mode 2 cobble cores are both above the 150mm cut-off proposed by Sharon [-@sharon2009] and the maximum scar length on each Mode 2 cobble core substantially overlaps with the size range of unmodified large flakes (Table 3).
-Importantly, this is also true of the largest retouched flakes at Gona, nine of which exhibit flake scars \>100mm and thus could have been used as LFB cores.
+The possibility that some Mode 2 flaked cobbles at Gona might be the remains of cores used for large flake production was initially suggested by informal, “experiential” knapping with Gona cobbles, during which attempts at LFB production produced pick-like forms as a byproduct.
+Consistent with this hypothesis, the mean (152.9mm) and median (156mm) length of Mode 2 cobble cores are both above the 150mm cut-off proposed by Sharon [-@sharon2009] and the maximum scar length on each Mode 2 flaked cobble substantially overlaps with the size range of unmodified large flakes (Table 3).
+Importantly, this is also true of the largest flaked flakes at Gona, nine of which exhibit flake scars \>100mm and thus could have been used as LFB cores.
 This possibility has also been suggested at Koobi Fora [@presnyakova2018] and might be relevant to understanding very large LCTs on flakes at other sites [e.g., @beyene2013].
 
-Nevertheless, the fact remains that the preponderance of scars on large cobble cores are from the removal of smaller flakes.
+Nevertheless, the fact remains that the preponderance of scars on large cobbles are from the removal of smaller flakes.
 Although these smaller scars could record preparatory flaking for LFB predetermination [@delatorre2018] and/or a subsequent stage of small flake debitage [@shea2010], it is difficult to rule out a role in shaping.
-Indeed, some pieces (e.g., Figure 1 ac, ad, ae) exhibit delicate points and sharp edges that are strongly suggestive of intentional shaping.
+Indeed, some pieces (e.g., Fig.
+2 aa, ab, ac) exhibit delicate points and sharp edges that are strongly suggestive of intentional shaping.
 If these pieces did in fact begin as cores for LFB production, it is possible that they were subsequently shaped into cutting tools in their own right.
-Such lithic "upcycling" of depleted LFB cores would help to explain the transport of these pieces away from the axial river system.
-However, we should also be cautious not to overinterpret a small number of suggestive pieces pulled from a wider range of variation, as they may simply represent low frequency "spandrels" of un-structured debitage [@moore2016; @moore2020].
-These various possibilities remain to be tested, but we note that there is no a priori reason to think that Acheulean tool makers would have neglected potentially useful cores because they "belonged" to a particular reduction sequence or technological type.
+Such lithic “upcycling” of depleted LFB cores would help to explain the transport of these pieces away from the axial river system.
+However, we should also be cautious not to overinterpret a small number of suggestive pieces pulled from a wider range of variation, as they may simply represent low frequency “spandrels” of un-structured debitage [@moore2016; @moore2020].
+These various possibilities remain to be tested, but we note that there is no a priori reason to think that Acheulean tool makers would have neglected potentially useful cores because they “belonged” to a particular reduction sequence or technological type.
 
-## Implications
+## Implications for understanding the early Acheulean
 
@@
 However, discussion of the technological, cognitive, and cultural implications of this transition has often focused on the emergence (or not) of intentional shape imposition [e.g., @duke2021; @holloway1969; @gowlett1986; @isaac1976; @roche2005; @wynn1995; @wynn2002].
-Our results suggest that this emphasis may be misplaced, and that distinctive Early Acheulean artifact forms might represent secondary accommodations to the primary goal of increasing tool size.
-From this perspective, Early Acheulean shape imposition is best understood as reflecting an interaction between functional design goals and technological constraints [@kuhn2020; @wynn2018].
+Our results suggest that this emphasis may be misplaced, and that distinctive early Acheulean artifact forms might represent secondary accommodations to the primary goal of increasing tool size.
+From this perspective, early Acheulean shape imposition is best understood as reflecting an interaction between functional design goals and technological constraints [@kuhn2020; @wynn2018].
+Note that this would still imply intentional design choices.
+What is at issue is the nature of the goals being pursued and the strategies used to achieve them.
 
@@
 Somewhat more controversially, Oldowan flaking may have included preferred debitage patterns [@stout2019] and/or intentional core maintenance and rejuvenation strategies [@delagnes2005].
-Our failure to find systematic effects of reduction on Oldowan core shape at Gona is consistent with this broad characterization.
+Our failure to find systematic effects of reduction on Oldowan flaked piece shape at Gona is consistent with this broad characterization.
 Starting from this Oldowan baseline, the production of larger cutting tools could in principle be achieved by increasing the size of cores and detached flakes and/or by attempting to produce and maintain cutting edges on cores themselves rather than on the smaller pieces detached from them.
-Both strategies are evident in the Early Acheulean at Gona.
+Both strategies are evident in the early Acheulean at Gona.
 
 At Gona, large flakes appear to have been transported across the landscape and discarded either in their original form or after relatively light modification to impose desired cutting edges.
-This behavior combines the two size-maximizing strategies identified above by increasing flake size and then using these large detached pieces as supports for retouched edges.
-On the other hand, large cobble cores at Gona display shape variation more consistent with debitage than shaping.
-This leads us to suggest that these pieces might be depleted cores from large flake production (size-maximizing strategy 1) that have been "upcycled" as large core tools (size-maximizing strategy 2).
-In this way, the full range of artifact types and patterns of shape variation in the Gona Early Acheulean can be parsimoniously accounted for as the expression of two strategies for increasing cutting tool size.
+This behavior combines the two size-maximizing strategies identified above by increasing flake size and then using these large, detached pieces as supports for flaked edges.
+On the other hand, the sample of large flaked cobbles at Gona displays shape variation more consistent with debitage than shaping.
+This leads us to suggest that some of these pieces might be depleted cores from large flake production (size-maximizing strategy 1) that have been “upcycled” as large core tools (size-maximizing strategy 2).
+In this way, the full range of artifact types and patterns of shape variation in the Gona early Acheulean can be parsimoniously accounted for as the expression of two strategies for increasing cutting tool size.
 
@@
 First, it supports the view that the Acheulean initially emerged as a set of technological strategies for increasing cutting tool size, especially by striking large flakes [@isaac1969].
-Hypothetically, this size increase might have been favored by novel functional priorities, such as extended use-life, enhanced transportability, and utility for heavy-duty cutting [@shea2010; @toth2019].
-It may also have been motivated and/or enabled by increases in hominid body size.
+Hypothetically, this size increase might have been favored by novel functional priorities of hominins for their tools, such as extended use-life, enhanced transportability, and utility for heavy-duty cutting [@shea2010; @toth2019].
+It may also have been motivated and/or enabled by increases in hominin body size.
 These overlapping possibilities appear likely to be mutually reinforcing rather than mutually exclusive.
-In fact, the appearance of the Acheulean starting 1.95 mya [@mussi2023] is roughly contemporaneous with an increase in the number, size, and ecogeographic range of stone tool sites in general [@plummer2004], as well as evidence of habitat diversification and greater lithic transport distances for Acheulean sites specifically [@hay1976; @quade2008; @rogers1994; @delatorre2008], increasing emphasis on large animal butchery [@linaresmatás2021], and increases in hominid brain and body size [@antón2014] including the first appearance of *Homo erectus* [@mussi2023].
-Increased cutting tool size is a plausible response and/or enabling factor to many of these shifts and is in that sense appears "overdetermined" by available evidence.
-Importantly, the current study does not resolve the extent to which the production and transport of Early Acheulean tools around the landscape [@presnyakova2018] occurred through planned logistical [@binford1980] activities of individuals and/or collaborating groups vs.
+In fact, the appearance of the Acheulean between 1.95 Ma [@mussi2023, but see @gossa2024] and 1.7 Ma [@lepre2011; @beyene2013; @duke2021] is roughly contemporaneous with an increase in the number, size, and ecogeographic range of stone tool sites in general [@plummer2004], as well as evidence of habitat diversification and greater lithic transport distances for Acheulean sites specifically [@hay1976; @rogers1994; @quade2008; @delatorre2008], increasing emphasis on large animal butchery [@linaresmatás2021], and increases in hominin brain and body size [@antón2014] including the first appearance of *Homo erectus* between 2.04–1.95 Ma [@herries2020].
+Increased cutting tool size is a plausible response and/or enabling factor to many of these shifts and is in that sense appears “overdetermined” by available evidence.
+Importantly, the current study does not resolve the extent to which the production and transport of early Acheulean tools around the landscape [@presnyakova2018] occurred through planned logistical [@binford1980] activities of individuals and/or collaborating groups vs.
 the cumulative effects of independent transport and recycling events over time [@reeves2023].
@@
 Second, it reinforces a growing consensus [e.g., @duke2021; @kuhn2020; @presnyakova2018] that early Acheulean LCT types artificially partition a continuum of variation rather than representing distinct target forms.
-More specifically, we propose that these forms emerge as the expression of a generic goal of cutting tool size maximization implemented across varying material constraints and opportunities.
-This is particularly relevant to the interpretation of "picks," which have long been viewed as a morphologically distinct but functionally mysterious tool type.
+More specifically, we propose that these forms emerge as the expression of a generic goal of cutting tool size maximization implemented across varying material constraints and
+opportunities.
+This is particularly relevant to the interpretation of “picks,” which have long been viewed as a morphologically distinct but functionally mysterious tool type.
 Toth and Schick [-@toth2019] note that picks do not appear designed to provide good cutting edges but also do not show use-wear consistent with digging.
 This leads them to speculate that picks may have been specialized weapons used to dispatch large, wounded animals with a blow to the head.
-However, picks are a common, persistent, and morphologically variable artifact type in the Early Acheulean, which argues against such a narrow function.
-We suggest that typological picks may lump together a variable mix of depleted LFB cores and large cutting tools made on relatively thick (cobble or flake) blanks, possibly including upcycled LFB cores.
-Within this interpretation, picks would be part of a continuum of morphological variation produced by different raw material forms and complex reduction histories, rather than a distinct tool type designed for novel function.
+However, picks are a common, persistent, and morphologically variable artifact type in the early Acheulean, which argues against such a narrow function.
+We suggest that typological picks may lump together a variable mix of depleted LFB
+cores and large cutting tools made on relatively thick (cobble or flake) blanks, possibly including upcycled LFB cores.
+On this interpretation, picks would be part of a continuum of morphological variation produced by different raw material forms and complex reduction histories, rather than a distinct tool type designed for novel function.
 Due to such constraints, picks as a class may appear less well designed for cutting than LCTs produced on thinner blanks.
 However, the relationship between tool morphology and cutting efficiency is complex and multivariate [@keyIntegratingMechanicalErgonomic2016], and the relative utility of short and/or higher-angle edges on a more massive tool is not well understood across diverse cutting tasks.
+
 Picks as an artifact class have not received the same attention as handaxes and cleavers and we are unaware of any broad comparative synthesis.
 Beyene et al.
-[-@beyene2013] do report that picks at Konso are relatively conservative in shape and scar counts over a period from 1.75-1.2 Ma during which associated handaxes show clear increases in refinement.
-They thus suggest that pick function may already have been effectively optimized by 1.75-1.6 Ma whereas the cutting function of handaxes continued to be enhanced.
-This is broadly consistent with the current suggestion that typological picks identify a morphological extreme produced by raw material features (esp.
-thickness) that constrain or discourage the refinement that might be developed on other blanks.
-From 1.75-1.2 Ma, these ad hoc and/or less refined cutting tools appear to have remained a stable element in the tool kit.
-However, they are nearly absent at 0.85 Ma, coincident with the appearance of "considerably refined" handaxes.
+[-@beyene2013] report that picks at Konso are relatively conservative in shape and scar counts over a period from 1.75–1.2 Ma during which associated handaxes show clear increases in refinement.
+They thus suggest that pick function may already have been effectively optimized by 1.75 to 1.6 Ma whereas the cutting function of handaxes continued to be enhanced.
+This is broadly consistent with the current suggestion that typological picks identify a morphological extreme produced by raw material features (especially thickness) that constrain or discourage the refinement that might be developed on other blanks.
+From 1.75–1.2 Ma, these ad hoc and/or less refined cutting tools appear to have remained a stable element in the Konso tool kit.
+However, they are nearly absent at 0.85 Ma, coincident with the appearance of “considerably refined” handaxes.
 Indeed, picks are not commonly reported to co-occur with refined handaxes in later Acheulean contexts anywhere.
-On a traditional functional interpretation, this would imply that their originally designed function also became less important in these contexts.
+On a traditional functional interpretation, this would imply that whatever it is that picks were originally designed to do also became less important in these contexts.
 The alternative possibility suggested here is that increased investment in handaxe refinement decreased the perceived value of producing and using ad hoc core tools for similar functions.
 
-Lastly, our interpretation suggests that LCT variation at Gona probably does not reflect intentional imposition of morphological norms [cf. @holloway1969] or detailed mental templates [@deetz1967].
+Lastly, our interpretation suggests that LCT variation at Gona probably does not reflect intentional imposition of morphological norms [cf.
+@holloway1969] or detailed mental templates [@deetz1967].
 Instead, we would focus attention on the cognitive and cultural implications of size maximization strategies.
-As reviewed by Stout [-@stoutStoneToolmakingEvolution2011], these stem from 1) the physical and strategic challenges of quarrying large flakes and 2) the increased complexity of knapping action hierarchies resulting from the addition of novel sub-goals (e.g.
-create a cutting edge) involved in shaping large flake and cobble cores into useful tools.
+As reviewed by Stout [-@stoutStoneToolmakingEvolution2011], these stem from 1) the physical and strategic challenges of quarrying large flakes and 2) the increased complexity of knapping action hierarchies resulting from the addition of novel sub-goals (e.g., create a cutting edge) involved in shaping large flake and cobble cores into useful tools.
+
 Systematic large flake production strategies and techniques are qualitatively different from small-flake debitage [@isaac1969; @toth2019] and thus add to the volume of technical knowledge and know-how that must be acquired by individuals.
@@
 Stout [-@stout2002] provides a modern example of large flake production that illustrates the potential scope of such demands.
-Interestingly, much earlier large flake production has been reported at the 3.3 Ma site of Lomekwi 3 but is argued to be poorly controlled with minimal core reduction, numerous steps and hinges, and extensive platform battering.
-Although concerns have been raised regarding the dating and context of this site [@archer2020; @dominguez-rodrigo2019], it may provide an interesting technological comparison to Early Acheulean large flake production.
+Interestingly, much earlier large flake production has been reported at the 3.3 Ma site of Lomekwi 3 [@harmand2015] but is argued to be poorly controlled with minimal core reduction, numerous steps and hinges, and extensive platform battering.
+Although concerns have been raised regarding the dating and context of this site [@archer2020; @dominguez-rodrigo2019], it may provide an interesting technological comparison to early Acheulean large flake production.
 
@@
 Such complexity has implications for assessing cognitive demands including relational integration, temporal abstraction, and goal abstraction [@stoutStoneToolmakingEvolution2011].
-For this reason, our study was designed to address concerns that intentional shaping might not actually be characteristic of Early Acheulean technology [e.g., @moore2016].
-We did find evidence for (at least) the imposition of cutting edges on LFBs, and offer this a foundation for "minimum necessary competence" [sensu @killin2021] cognitive interpretations.
-We remain agnostic as to whether similar knapping complexity was demonstrated in the preceding Oldowan, but note that the rare and un-standardized [@gallotti2018] occurrence of debitage on flakes prior to 2.0 mya does not provide evidence of intentional shaping.
-Stout [-@stoutStoneToolmakingEvolution2011] argued that elaborated small flake debitage methods contemporary with the Early Acheulean, such as single platform Karari scrapers [@isaac1997] and hierarchical centripetal cores [@delatorre2003], document complex goal structures including the intentional modification of the core morphology to enable subsequent flake detachments.
+For this reason, our study was designed to address concerns that intentional shaping might not actually be characteristic of early Acheulean technology [e.g., @moore2016].
+We did find evidence for (at least) the imposition of cutting edges on LFBs and offer this as a foundation for “minimum necessary competence” [sensu @killin2021] cognitive interpretations.
+We remain agnostic as to whether similar knapping complexity was demonstrated in the preceding Oldowan but note that the rare and un-standardized [@gallotti2018] occurrence of debitage on flakes prior to 2.0 Ma does not provide evidence of intentional shaping.
+Stout [-@stoutStoneToolmakingEvolution2011] argued that elaborated small flake debitage methods contemporary with the early Acheulean, such as single platform Karari scrapers [@isaac1997] and hierarchical centripetal cores [@delatorre2003], document complex goal structures including the intentional modification of the core morphology to enable subsequent flake detachments.
 Others have seen evidence of such intentions in earlier bifacial [@duke2021] and unifacial [@delagnes2005] debitage strategies.
-However, all of these interpretations are based on qualitative assessments of flake scar pattering and/or refitting and remain open to critique from quantitative and experimentally-based approaches contending that similar knapping patterns can emerge without "top down" intentions [@moore2016; @moore2020; @toth1985].
+However, all of these interpretations are based on qualitative assessments of flake scar pattering and/or refitting and remain open to critique from quantitative and experimentally-based approaches contending that similar knapping patterns can emerge without “top down” intentions [@moore2016; @moore2020; @toth1985].
 
@@
 Logically, the minimum capacities required to support a novel behavior must be present before the behavioral innovation can occur [@stout2023] and may predate it substantially.
-We would thus argue that technological innovation in the Early Acheulean was more likely stimulated by changing behavioral and ecological strategies that placed a premium on cutting tool size, rather than by the sudden emergence of new cognitive capacities.
-In this context, it is important to note that the preponderance of cognitive and especially neuroarchaeological investigations of handaxe manufacture [@hecht2023; @stout2015; @stoutStoneToolmakingEvolution2011; @stout2008] have studied refined, later Acheulean forms and should not be generalized to discussions regarding the emergence of the Early Acheulean [cf.
+We would thus argue that technological innovation in the early Acheulean was more likely stimulated by changing behavioral and ecological strategies that placed a premium on cutting tool size, rather than by the sudden emergence of new cognitive capacities.
+In this context, it is important to note that the preponderance of cognitive and especially neuroarchaeological investigations of handaxe manufacture [@hecht2023; @stout2015; @stoutStoneToolmakingEvolution2011; @stout2008] have studied refined, later Acheulean forms and should not be generalized to discussions regarding the emergence of the early Acheulean [cf.
 @duke2021].
 These neuroarchaeological studies were made possible by a robust understanding of the specific knapping behaviors being modeled [@stout2023].
-Similar studies modeling Early Acheulean technology are clearly needed and we hope that investigations like the current one can help to provide the necessary behavioral foundation.
+Similar studies modeling early Acheulean technology are clearly needed and we hope that investigations like the current one can help to provide the necessary behavioral foundation.
 
-## Generalizability
+## Generalizability to other early Acheulean sites
 
-It remains to be seen to what extent current results can be generalized to other Early Acheulean sites.
-It is clear that LFB production as seen at Gona is typical [@isaac1969] of the African Acheulean from its first known appearance [@mussi2023].
+It remains to be seen to what extent current results can be generalized to other early Acheulean sites.
+It is clear that LFB production as seen at Gona is typical [@isaac1969] of the African Acheulean from its first known appearances [@mussi2023; @lepre2011; @beyene2013].
 Similarly, the transport and discard of unmodified large flakes has also been reported at Peninj [@delatorre2008] and Koobi Fora [@presnyakova2018] and may be more widespread than has been specifically noted in publication.
-The general pattern of light, non-invasive retouch we observe also appears typical of Early Acheulean sites [reviewed by @presnyakova2018].
+The general pattern of light, non-invasive flaking we observe also appears typical of early Acheulean sites [reviewed by @presnyakova2018].
 However, the effects of reduction on LCT shape have not been systematically tested across sites in ways comparable to the current analyses.
-This makes it difficult to determine if rare examples of "well shaped" Early Acheulean LCTs [e.g., @diez-martín2019] represent distinct technological behaviors or extreme points along a continuum of variation.
-Presnyakova et al. [-@presnyakova2018] argue for the latter, supported by shape effects of reduction intensity identified using artifact size and edge angle as proxies.
-However, these authors do not consider SDI or FAI and their analyses excluded any flaked pieces without "clearly defined tips and bases," thus limiting comparability with current results.
+This makes it difficult to determine if rare examples of “well shaped” early Acheulean LCTs [e.g., @diez-martín2019] represent distinct technological behaviors or extreme points along a continuum of variation.
+Presnyakova et al.
+[-@presnyakova2018] argue for the latter, supported by shape effects of reduction intensity identified using artifact size and edge angle as proxies.
+However, these authors do not consider SDI or FAI and their analyses excluded any flaked pieces without “clearly defined tips and bases,” thus limiting comparability with current results.
 
-The occurrence of numerous large cobble cores (typological LCTs on cobble bases) at Gona is atypical for East Africa and may reflect local raw materials availability.
-Kuman [-@kuman2007] reports that river cobbles from the nearby Blaaubank gravels are the dominant blank form in Gauteng "Cradle of Humankind" Early Acheulean collections, possibly due to the rarity of larger boulder-size clasts.
+The occurrence of numerous large flaked cobbles (typological LCTs on cobble bases) at Gona is atypical for East Africa and may reflect local raw materials availability.
+Kuman [-@kuman2007] reports that river cobbles from the nearby Blaaubank gravels are the dominant blank form in Gauteng “Cradle of Humankind” early Acheulean collections, possibly due to the rarity of larger boulder-size clasts.
 Relatively high frequencies of LCTs on cobbles have also been reported from Gadeb in Ethiopia [@delatorreEarlyStoneAge2011].
@@
 [-@presnyakova2018: Fig.
-8] illustrate a boulder core with refitting large flake from FxJj65 that appears to be comparable in size (\~150mm long), shape (L/W $\approx$ 1.5), and reduction intensity to some of the more lightly worked cobble cores at Gona, providing at least anecdotal support for our interpretations.
-In sum, however, systematic comparison across published Early Acheulean sites is not currently possible due to differences in research questions, methods, and data reporting practices.
-Clearly much work is needed to enable such comparison, but we do find some encouragement in the current finding that conventional linear measures, simple scar counts, and visual flaked area estimations are sufficient to describe relevant variation in shape and reduction
+8] illustrate a boulder core with refitting large flake from FxJj65 that appears to be comparable in size (\~150 mm long), shape (L/W $\approx$ 1.5), and reduction intensity to some of the more lightly worked flaked cobbles at Gona, providing at least anecdotal support for our interpretations.
+In sum, however, systematic comparison across published early Acheulean sites is not currently possible due to differences in research questions, methods, and data reporting practices.
+Clearly much work is needed to enable such comparison, but we do find some encouragement in the current finding that conventional linear measures, simple scar counts, and visual flaked area estimations are sufficient to describe relevant variation in shape and reduction intensity/extent.
 
@@
 
-Characteristic Early Acheulean artifact forms including large retouched flakes and core tools began to appear ca.
-2.0 Ma [@mussi2023] as part of a pervasive ecological and behavioral shift encompassing changing diet, ranging patterns, and habitat usage; increasing site frequency, density, and ecogeographic distribution; and the first appearance of *Homo erectus*.
+Characteristic early Acheulean artifact forms including large flaked flakes and core tools began to appear ca.
+2.0-1.7 Ma as part of a pervasive ecological and behavioral shift encompassing changing diet, ranging patterns, and habitat usage; increasing site frequency, density, and ecogeographic distribution; and the first appearance of *H.
+erectus*.
 Although the cognitive and cultural significance of this new lithic industry is often framed in terms of the emergence of discrete tool types and/or intentionally imposed morphological norms, we find little evidence at Gona to support this.
 Instead, we observe systematic patterns of raw material selection and core surface modification organized around the production and maintenance of useful cutting edges on large(r) supports.
-Such larger cutting tools may have been prioritized in newly emerging hominid behavioral adaptations due to their extended use-life, enhanced transportability, and/or suitability for heavy-duty cutting.
+Such larger cutting tools may have been prioritized in newly emerging hominin life ways due to their extended use-life, enhanced transportability, and/or suitability for heavy-duty cutting.
 
-This interpretation is consistent with a characterization of Early Acheulean tool forms as emerging from a set of "design imperatives" for large hand-held cutting tools [@gowlett2006; @wynn2018], specifically including elongation and ergonomic positioning of cutting edges [[\@key2016; \@keyIntegratingMechanicalErgonomic2016](Key%20et%20al.,%202016;%20Key,%202016)].
-Our results thus provide support for the presence of imposed form in the broad theoretical sense developed by Kuhn [-@kuhn2020]: not as the realization of a "holographic model" in the mind of the maker, but rather as the outcome of technological choices made in view of material constraints and affordances [@moore2020].
-To the extent that some, but not all [@muller2022], later Acheulean assemblages provide stronger evidence of normative form imposition [@garcía-medrano2019; @liuDifferentialEffectsKnapping2023; @shipton2020] this underlines the heterogeneity of lithic technologies subsumed within the "Acheulean Industry" and argues for more context-specific approaches to evaluating behavioral, cognitive, and cultural implications.
+This interpretation is consistent with a characterization of early Acheulean tool forms as emerging from a set of “design imperatives” for large hand-held cutting tools [@gowlett2006; @wynn2018] , specifically including elongation and ergonomic positioning of cutting edges [@key2016; @keyIntegratingMechanicalErgonomic2016].
+Our results thus provide support for the presence of imposed form in the broad theoretical sense developed by Kuhn [-@kuhn2020]: not as the realization of a “holographic model” in the mind of the maker, but rather as the outcome of intentional technological choices made in view of material constraints and affordances [@moore2020].
+To the extent that some, but not all [@muller2022], later Acheulean assemblages provide stronger evidence of normative form imposition [@garcía-medrano2019; @shipton2020; @liuDifferentialEffectsKnapping2023] this underlines the heterogeneity of lithic technologies subsumed within the “Acheulean Industry” and argues for more context-specific approaches to evaluating behavioral, cognitive, and cultural implications.
 

### 2024-08-19 manuscript/manuscript_word.Rmd
finished another round of proofreading.

--- 2024-08-19 manuscript/manuscript.Rmd
@@
 - Sileshi Semaw^[Centro Nacional de Investigación sobre la Evolución Humana (CENIEH), Burgos, Spain; sileshi.semaw@cenieh.es]
-abstract:
-|
-  ```{=latex}
-  The appearance of ‘Large Cutting Tools’ in the early Acheulean is widely regarded as the first evidence for the imposition of intended forms on artifacts, with major implications for hominin cognitive and cultural capacities.
+abstract: The appearance of ‘Large Cutting Tools’ in the early Acheulean is widely regarded as the first evidence for the imposition of intended forms on artifacts, with major implications for hominin cognitive and cultural capacities.
 However, the nature and extent of explicit design documented by these forms remains open to debate.
@@
 Our results thus provide support for the presence of Acheulean design, not as the concrete realization of a ‘holographic model’ in the mind of the maker, but rather in the broader sense of deliberate technological choices made in view of behavioral goals and material constraints.
-
   \textbf{Keywords:} Lithic technology; Early Stone Age; Handaxe; Large Cutting Tool; Pick; Shaping
-  ```
 date: "`r Sys.Date()`"
 fontsize: 11pt
-output: bookdown::pdf_document2
+output: word_document
 header-includes:

### 2024-08-19 code/Visualization.R

@@ -13,7 +13,9 @@ library(diptest)
 gona_linear<-readr::read_csv("data/Gona_EA_for_Cheng_020824.csv")
 gona_3d<-read.csv("data/3D-cleaned-for-Cheng-080224.csv")
 
-## Figure for 3d comparison
+# Visualization
+
+## Figure 4 (3d and caliper measurement comparison)
 
 fig4a <- ggplot(gona_3d, aes(X.3DGMPC110x10, FAC1_2)) + 
   geom_point() + 
@@ -33,11 +35,92 @@ fig4c <- ggplot(gona_3d, aes(Flaked, FlakedVisual)) +
   labs(x ="Flaked percentage (3D)", y = "Flaked percentage (visual)")
 
 patchwork <- (fig4a + fig4b + fig4c)
-patchwork + plot_annotation(tag_levels = 'A')
-ggplot2::ggsave("Fig.3D comparison new1.png", path="figure.", width = 9, height = 3, dpi = 300)
+patchwork + plot_annotation(tag_levels = 'a')
+ggplot2::ggsave("Fig.4.png", path="figure.", width = 9, height = 3, dpi = 300)
 
 
-## Figure for FAI by base
+## Figure 5 (DFA results)
+### LDFA (typology based)
+
+theme_set(theme_classic()) 
+gona_linearfil <- gona_linear %>% filter(Typology == "Pick" | Typology =="Handaxe" | Typology == "Knife")
+gona_lineardfa <- dplyr::select(gona_linearfil, FAI, FAC1_1, cSDI, FAC2_1, Typology)
+
+# Split the data into training (80%) and test set (20%) 
+set.seed(123) 
+training.individuals <- gona_lineardfa$Typology %>%  
+  createDataPartition(p = 0.8, list = FALSE) 
+train.data <- gona_lineardfa[training.individuals, ] 
+test.data <- gona_lineardfa[-training.individuals, ] 
+
+# Estimate preprocessing parameters 
+preproc.parameter <- train.data %>%  
+  preProcess(method = c("center", "scale")) 
+
+# Transform the data using the estimated parameters 
+train.transform <- preproc.parameter %>% predict(train.data) 
+test.transform <- preproc.parameter %>% predict(test.data) 
+
+# Fit the model 
+model <- lda(Typology~., data = train.transform) 
+
[…]

### 2024-08-20 code/Visualization.R
figure update.

@@ -7,6 +7,7 @@ library(MASS)
 library(caret) 
 library(lsr)
 library(diptest)
+library(DFA.CANCOR)
 
 # Data preparation
 ## loading data
@@ -42,82 +43,64 @@ ggplot2::ggsave("Fig.4.png", path="figure.", width = 9, height = 3, dpi = 300)
 ## Figure 5 (DFA results)
 ### LDFA (typology based)
 
+
 theme_set(theme_classic()) 
 gona_linearfil <- gona_linear %>% filter(Typology == "Pick" | Typology =="Handaxe" | Typology == "Knife")
 gona_lineardfa <- dplyr::select(gona_linearfil, FAI, FAC1_1, cSDI, FAC2_1, Typology)
 
-# Split the data into training (80%) and test set (20%) 
-set.seed(123) 
-training.individuals <- gona_lineardfa$Typology %>%  
-  createDataPartition(p = 0.8, list = FALSE) 
-train.data <- gona_lineardfa[training.individuals, ] 
-test.data <- gona_lineardfa[-training.individuals, ] 
 
-# Estimate preprocessing parameters 
-preproc.parameter <- train.data %>%  
-  preProcess(method = c("center", "scale")) 
+DFA_Field=DFA.CANCOR::DFA(data = gona_lineardfa, 
+              groups = 'Typology', 
+              variables = c('FAI', 'FAC1_1', 'cSDI', 'FAC2_1'),
+              predictive = TRUE, 
+              priorprob = 'EQUAL',   
+              covmat_type='within', # altho better to use 'separate' for these data
+              verbose = TRUE)
+DFA_Field$dfa_scores$Function.2<- (-1)*(DFA_Field$dfa_scores$Function.2)
 
-# Transform the data using the estimated parameters 
-train.transform <- preproc.parameter %>% predict(train.data) 
-test.transform <- preproc.parameter %>% predict(test.data) 
+P0<-ggplot(DFA_Field$dfa_scores, aes(Function.1, Function.2)) + geom_point(aes(color = group))+
+  labs(x ="DF1", y = "DF2")
+P1<-ggExtra::ggMarginal(P0+ theme(legend.position = "left"),type="histogram")
 
-# Fit the model 
-model <- lda(Typology~., data = train.transform) 
[…]

### 2024-08-22 code/Analysis.R
readme update and folder cleaning.

@@ -1,260 +1,0 @@
-# Add packages
-library(tidyverse)
-library(factoextra)
-library(patchwork)
-
-# Data preparation
-## loading data
-gona_linear<-readr::read_csv("data/Gona_ESA_Linear.csv")
-gona_3d<-readr::read_csv("data/Gona_ESA_3D.csv")
-## creating a new column of unique identifier for merging
-gona_linear$Name <- paste(gona_linear$Site, gona_linear$Catalog, sep= "-")
-## merging two datasets based on Name
-joined_gona <-dplyr::left_join(gona_linear, gona_3d, by = c("Name" = "Name"))
-## Reordering column order for principle component analysis
-joined_gona <- joined_gona %>% dplyr::relocate(GM_L,GM_Br1,GM_Br2,GM_Br3,GM_T1,GM_T2,GM_T3)
-## Subsetting a dataframe that contains both linear and 3d data for measuremnt validation
-overlap_gona <- joined_gona %>% tidyr::drop_na(Flaked.y)
-
-
-# Measurement Validation
-## Principle component analysis
-gona_pca_overlap <-prcomp(overlap_gona[1:7], scale = FALSE)
-overlap.res.var <- get_pca_var(gona_pca_overlap)
-overlap_variable_loadings<-data.frame(overlap.res.var$coord) 
-## Screeplot showing percentage of explained variances
-fviz_eig(gona_pca_overlap, addlabels=TRUE, hjust = -0.1)
-## Results for individuals
-overlap.res.ind <- get_pca_ind(gona_pca_overlap)
-Indiv_scores_overlap<-data.frame(overlap.res.ind$coord)
-## add data back to dataset
-overlap_gona$overlapPC1<-Indiv_scores_overlap$Dim.1
-overlap_gona$overlapPC2<-Indiv_scores_overlap$Dim.2
-## Comparison of linear PC1 and 3D PCA
-results_expr <- statsExpressions::corr_test(overlap_gona, "overlapPC1", "@3DGMPC110x10")$expression[[1]]
-fig4a<-ggplot(overlap_gona, aes(x="overlapPC1", y="@3DGMPC110x10")) + 
-  geom_point() + 
-  geom_smooth(method=lm) +
-  labs(subtitle = results_expr)
-
[…]

### 2024-08-22 code/Visualization.R

@@ -215,412 +215,6 @@ leftfig<-joined_gona %>%
 
 
 
-
-
-
-
-## figures for demonstrating the conceptual model as requested by R3
-# gona_linear<- gona_linear %>% unite("FULLNAME", Locality:Catalog, remove = FALSE)
-### mapping SDI and FAI
-# fig0a <- ggplot(gona_linear, aes(cSDI, FAI, color = Typology)) +
-#   geom_point() + 
-#   # geom_smooth(method=lm)+
-#   ggrepel::geom_text_repel(aes(label=FULLNAME))+
-#   labs(x ="cSDI", y = "FAI")
-
-# 
-# 
-# ### correlation between SDI/PC1 correlation and FAI/PC1
-# correlation1<- gona_linear %>% group_by(Locality) %>%  summarise(SDIPC1R = abs(cor(cSDI, FAC1_1)))
-# correlation2<- gona_linear %>% group_by(Locality) %>%  summarise(FAIPC1R = abs(cor(FAI, FAC1_1)))
-# correlation12<- merge(correlation1,correlation2,by="Locality")
-# 
-# ### correlation between SDI/PC2 correlation and FAI/PC2
-# 
-# correlation3<- gona_linear %>% group_by(Locality) %>%  summarise(SDIPC2R = abs(cor(cSDI, FAC2_1)))
-# correlation4<- gona_linear %>% group_by(Locality) %>%  summarise(FAIPC2R = abs(cor(FAI, FAC2_1)))
-# correlation34<- merge(correlation3,correlation4,by="Locality")
-
-
-# ### correlation between SDI/PC1 correlation and FAI/PC1
-# correlation1<- gona_linear %>% group_by(Typology) %>%  summarise(SDIPC1R = abs(cor(cSDI, FAC1_1)))
-# correlation2<- gona_linear %>% group_by(Typology) %>%  summarise(FAIPC1R = abs(cor(FAI, FAC1_1)))
-# correlation12<- merge(correlation1,correlation2,by="Typology")
-
-### correlation between SDI/PC2 correlation and FAI/PC2
-# 
-# correlation3<- gona_linear %>% group_by(Typology) %>%  summarise(SDIPC2R = abs(cor(cSDI, FAC2_1)))
-# correlation4<- gona_linear %>% group_by(Typology) %>%  summarise(FAIPC2R = abs(cor(FAI, FAC2_1)))
[…]

### 2024-08-22 code/old/txt_reading_in_mass.R

@@ -1,236 +1,0 @@
-library(tidyverse)
-
-list_of_files <- list.files(path = "/Users/raylc/Downloads/Boxgrove Taylor work/Handaxes/boxgrove_measurements/plan_2", recursive = TRUE, full.names = TRUE)
-
-# Read all the files and create a FileName column to store filenames
-df <- list_of_files %>%
-  set_names(.) %>%
-  map_df(read_table2, .id = "FileName")
-Handaxes/boxgrove_measurements
-
-
-
-
-
-
-
-
-
-
-
-
-
-plan_filenames_160<-list.files("Handaxes/boxgrove_measurements/plan_822_160", pattern="*.txt")
-profile_filenames_160<-list.files("Handaxes/boxgrove_measurements/profile_822_160", pattern="*.txt")
-plan_filenames_162<-list.files("Handaxes/boxgrove_measurements/plan_2", pattern="*.txt")
-profile_filenames_162<-list.files("Handaxes/boxgrove_measurements/profile_2", pattern="*.txt")
-
-
-
-merge_shape_data_function_160<-function(plan_names,profile_names,plan_pathway,profile_pathway) {
-  
-  # combine txt files for each recording set
-  handaxe_plan_measurements<-do.call(rbind,lapply(plan_names,open_read,plan_pathway))
-  handaxe_profile_measurements<-do.call(rbind,lapply(profile_names,open_read,profile_pathway))
-  
-  # clean and reshape plan data 
-  handaxe_plan_measurements_reshape<-handaxe_plan_measurements %>%
-    mutate(variable=recode(handaxe_plan_measurements$variable, Width = "width"),
-           measurement_point = paste(variable, measurement_point, sep="_"),
-           shape_width_mm=measurement*25.4) %>%
-    select(-c(variable,measurement)) %>%
-    dcast(individual ~ measurement_point,value.var="shape_width_mm")
-  
-  # clean and reshape profile data 
-  handaxe_profile_measurements_reshape<-handaxe_profile_measurements %>%
-    mutate(variable=dplyr::recode(variable, Width = "thickness"),
-           measurement_point = paste(variable, measurement_point, sep="_"),
-           shape_thickness_mm=measurement*25.4) %>%
-    select(-c(variable,measurement)) %>%
[…]
